### 2019-11-09 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
reprocessing script for feature counts

---
title: "MSSM-Penn-Pitt - ACC and DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
author: "Kelsey Montgomery"
date: "11/8/2019"
output: html_document
---

```{r knit2synapse, eval=FALSE}
library(synapser)
library(knit2synapse)

synLogin()

knit2synapse::createAndKnitToFolderEntity(file = "./MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
                                 parentId ="syn9872607",
                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model")
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)
library(R.utils)
library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(lme4)
library(biomaRt)

library(synapser)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

synLogin()

library(doParallel)
library(foreach)

cl = makeCluster(detectCores()-2)

registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn9872607';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of GRCh38 feature counts with CQN normalisation (ACC and DLPFC)';

thisFileName <- 'MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd'

# Github link
thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='master')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
downloadFile <- function(id){
  fread(synGet(id)$path, data.table = F)
}
downloadFile_version <- function(id , version){
  fread(synGet(id, version = version)$path, data.table = F)
}
# Download reprocessed counts
counts = 'syn21071622'
ALL_USED_IDs = counts
counts = downloadFile(counts) %>% data.frame()
counts$Chr <- NULL
counts$Start <- NULL
counts$End <- NULL
counts$Strand <- NULL

# Parse gene lengths
gene_lengths = counts[,c("Geneid", "Length")]
counts$Length <- NULL
gene_lengths <- rename(gene_lengths, gene_id = Geneid)

# Get ancestry vector calculated using gemtools
ANCESTRY_ID = 'syn2511399'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
ANCESTRY = downloadFile(ANCESTRY_ID) %>% 
  plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'SNP_report:Genotyping_Sample_ID'))

# Get genotype ids from synapse
GENOTYPE_ID = 'syn16816490'
ALL_USED_IDs = c(ALL_USED_IDs, GENOTYPE_ID)
GENOTYPE = downloadFile(GENOTYPE_ID) %>% 
  dplyr::select(Individual_ID, `SNP_report:Genotyping_Sample_ID`) %>%
  dplyr::inner_join(ANCESTRY)

#Get Metadata
metadata = 'syn16816488' 
ALL_USED_IDs[length(ALL_USED_IDs)+1] = metadata
metadata = downloadFile_version(metadata, version = 13)  %>% 
  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
                Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch','rnaSeq_dissection:Brain_Region',
                                      'rnaSeq_report:Exclude?',
                                      'rnaSeq_report:Mapped_Reads',
                                      'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
                                      'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
                                      'rnaSeq_report:Expression_Profiling_Efficiency',
                                      'rnaSeq_report:rRNA_Rate',
                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Percent_Aligned'))

METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>%
  dplyr::rename(PMI = `PMI_(in_hours)`,
                ReportExclude = `rnaSeq_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                Tissue = `rnaSeq_dissection:Brain_Region`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, 
                IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, 
                GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, 
                rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, 
                AlignmentRate = `rnaSeq_report:Percent_Aligned`,
                RIN = `rnaSeq_isolation:RIN`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, Age_of_Death, PMI, Dx, 
                RIN, ReportExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, RibozeroBatch, Tissue, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate) %>% 
  dplyr::filter(Institution %in% c("MSSM", "Penn", "Pitt")) %>% 
  dplyr::filter(SampleID %in% colnames(counts), !is.na(SampleID)) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx,  Other = "AFF", Other = "BP", Control = "Control", SCZ = "SCZ"))
```

### Data preprocessing
```{r preprocess.data}
ind = METADATA$SampleID[which(METADATA$ReportExclude == 1)]
writeLines(paste('Following',length(ind),'samples are marked exclude'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Ethnicity) | is.na(METADATA$Institution) | is.na(METADATA$Dx)]
writeLines(paste('Following', length(ind), 'counts are missing clinical metadata'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$PMI)]
writeLines(paste('Following', length(ind), 'counts are missing PMI'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Reported_Gender)]
writeLines(paste('Following', length(ind), 'counts are missing gender'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Age_of_Death)]
writeLines(paste('Following', length(ind), 'counts are missing age of death'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$EV.1)]
writeLines(paste('Following', length(ind), 'counts are missing ancestry information'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))
```

```{r preprocess.data1, results='asis'}
# Match covariates to expression data
indToRetain = intersect(METADATA$SampleID, colnames(counts))

rownames(counts) <- counts$Geneid
counts = counts[,indToRetain]

rownames(METADATA) = METADATA$SampleID
METADATA = METADATA[indToRetain,]

METADATA %>% 
  group_by(Dx, Tissue) %>% 
  summarise(count = n()) %>% 
  spread(Tissue, count) %>%
  kable()

METADATA = METADATA %>%
  dplyr::mutate(Dx.Tissue = paste(Dx, Tissue, sep = '.'))
```
Following sample are removed `r paste(removedIDs, collapse = ',')`

```{r gene.param}
## Get GC content from biomart
backgroundGenes = data.frame(gene_id = rownames(counts)) %>%
  dplyr::mutate(id = gene_id) %>%
  tidyr::separate(id, c('ensembl_gene_id','position'), sep = '\\.')

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "sep2019.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>% 
  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
```

### Sex Check
```{r Reported_Gender_plots}
counts$gene_id = rownames(counts) 
REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
  left_join(counts) %>%
  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
  mutate(value = log(value)) %>%
  rename(`counts(log)`= value) %>% 
  rename(SampleID = item) %>%
  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
  rename(`Reported Gender` = Reported_Gender)

##XIST and UTY expression 
#ENSG00000229807.10 and ENSG00000183878.15 
FILT <- REPORTED.GENDER.COUNTS %>% 
  filter(gene_id == "ENSG00000229807.10" | gene_id == "ENSG00000183878.15") %>% 
  dplyr::select(-one_of("chromosome_name")) %>% 
  tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
  mutate(XIST = as.numeric(`ENSG00000229807.10`)) %>% 
  mutate(UTY = as.numeric(`ENSG00000183878.15`))

p = ggplot(FILT, aes (x= XIST, y = UTY)) 
p = p + geom_point(aes(color=`Reported Gender`)) 
p
```
### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID', "Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads","AlignmentRate", 
                     "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){
  x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- METADATA$SampleID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], function(x){
  x = as.numeric(as.character(gsub('[\\,\\%]','',x)))
})

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
tmp = COVARIATES.CORRELATION$ESTIMATE
tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
draw(h, heatmap_legend_side = 'left')
```

## Explore metatdata
```{r data.explore, fig.width = 12, fig.height = 12}
my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Age_of_Death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme

# Transcripts Detected
p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntragenicRate)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Intragenic Rate') + my.theme

# Mapped Reads
p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme

# PercentAligned
p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TotalReads)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('Total Reads') + my.theme

# rRNARate
p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot()
p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme

multiplot(plotlist = p, cols = 3)

# Institution
# vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES)

# Gender
# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
```{r cpmnormalisation}
counts$gene_id = NULL

genesToAnalyze = dlply(METADATA, .(Dx.Tissue, Reported_Gender), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, counts)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC.CONT$gene_id[!is.na(GENE.GC.CONT$percentage_gene_gc_content)]) %>%
  intersect(gene_lengths$gene_id[!is.na(gene_lengths$Length)])

#Subset union of all genes and convert subset to logcpm values 
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(counts[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
  dplyr::rename(gene_id = genes) %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  group_by(gene_biotype) %>%
  summarise(fraction  = n()) %>%
  filter(fraction > 100) %>%
  dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
```
`r dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]` genes are used for the analysis.

Following is the distribution of biotypes
`r kable(pct.pc)`

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")

indToRemove = c(indToRemove, c("MSSM_RNA_ACC_224", "MSSM_RNA_ACC_240", "MSSM_RNA_ACC_BP_11", "MSSM_RNA_ACC_42", "MSSM_RNA_ACC_BP_15", "PENN_RNA_ACC_003", "MSSM_RNA_ACC_374"))

indToRemove = c(indToRemove, "PITT_RNA_BP_PFC_1770")

# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
  dplyr::mutate(label = SampleID)
plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

# Plot abberent distribution of logcpm counts
tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% indToRemove),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
p

indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
COVARIATES = COVARIATES[indToRetain,]

tmp = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue'), sep = '_') %>%
  group_by(Dx, Tissue) %>%
  summarise(count = n()) %>%
  spread(Tissue, count)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals

Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`

Distribution of samples are:
`r kable(tmp)`

### Winsorise counts (Gene outliers in samples)
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(LOG.CPM, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = NA
  x[x > (mn+3*std.dev)] = NA
  return(x)
}) %>% t

# Back calculate counts for further analysis
NEW.COUNTS = PROCESSED_COUNTS$filteredExprMatrix$counts
NEW.COUNTS[is.na(LOG.CPM)] = 0
```

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
gene_lengths = tibble::column_to_rownames(gene_lengths, var = 'gene_id') %>% as.data.frame()
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gene_gc_content'],
                          lengths = gene_lengths[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
NEW.COUNTS[log.mat] = NA
CQN.GENE_EXPRESSION$E[log.mat] = NA
```

### Co-expression distribution
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
cr = cor(t(tmp))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
PC <- prcomp(tmp, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
tree = hclust(as.dist(t(tmp)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r temp, include = F}
dev.off()
gc()
```

### Distribution of samples (log cpm)
```{r lcpm.dist, cache=FALSE}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$E) %>%
  rownameToFirstColumn('gene_id') %>%
  tidyr::gather(SampleID, logCPM, -gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID'))

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx.Tissue, scale = 'free')
p
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
preAdjustedSigCovars = runPCAandPlotCorrelations(tmp, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Model Identification
```{r fxn}
# Function to fit a mixed effect model and return BIC
fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){

  # Get design matrix
  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]

  # Estimate voom weights
  cnt[is.na(cnt)] = 0
  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights

  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
                              .combine = rbind,
                              .packages = c('lme4', 'dplyr'),
                              .export = c('voomWeights')) %dopar% {

                                # Perform weighted mixed linear models for every gene
                                expr = data.frame(Gx = lcpm[i,])
                                data = cbind(expr,
                                             designMat[rownames(expr),],
                                             Individual_ID = groupingVar)
                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')

                                weights = voomWeights[i,]
                                BIC = NA
                                tryCatch({
                                  mdl = lmer(formula = formula, 
                                             data = data, 
                                             weights = weights)
                                  BIC = BIC(mdl)
                                }, error = function(e){
                                  BIC = NA
                                })
                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
                                                BIC = BIC))
                            }
  return(modelBIC)
  }

# Function to select covariates in a step-wise regression fashion
selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){

  varsSelected = c()
  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
  while(length(covarsToSelect) != 0){
    allModelBIC = foreach(i = 1:length(covarsToSelect),
                          .combine = c,
                          .packages = pkgs,
                          .export = c('fitMixedEffectModel')) %dopar% {

                            # Fit a mixed effect linear model and get all BIC
                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                           covars[colnames(cnt),'Individual_ID'], 
                                                           lcpm, 
                                                           cnt, 
                                                           covars) %>%
                            plyr::rename(c('BIC' = covarsToSelect[i]))

                            return(list(modelBIC))
                          }
    allModelBIC = plyr::join_all(allModelBIC)

    # Find number of genes that are changed based on BIC
    nChange = allModelBIC %>%
      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
      dplyr::left_join(baseModelBIC) %>%
      dplyr::mutate(diff = baseModel - Value) %>%
      dplyr::group_by(ModelName) %>%
      dplyr::summarise(better = sum(diff >= 5), 
                       worse = sum(diff <= -5)) %>%
      dplyr::mutate(nchange = better - worse)

    print(nChange)

    # Select variable
    tmp = nChange %>%
      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
      dplyr::top_n(1, nchange)

    if (dim(tmp)[1] == 0){
      covarsToSelect = c()
    } else {
      baseModelBIC = allModelBIC %>%
        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
      colnames(baseModelBIC)[2] = 'baseModel'
      varsSelected = c(varsSelected, tmp$ModelName)
      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
      print(paste('Variable chosen to be added:',tmp$ModelName))
    }
  }

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('Dx.Tissue')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase2.covars = c('LibraryBatch', 'FlowcellBatch')
vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase3.covars = c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected", "rRNARate", "TotalReads")
vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`

### Normalisation
1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect
```{r iterative.normalisation, results='asis'}
writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))

# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate voom weights
tmp = NEW.COUNTS
tmp[is.na(tmp)] = 0
VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)

# Estimate consensus correlation coefficient
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION = voom(tmp, 
                            design=DM1$design, plot=F,
                            block = COVARIATES$Individual_ID, 
                            correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
tmp = RESIDUAL.GENE_EXPRESSION
tmp[is.na(tmp)] = 0
residualSigCovars = runPCAandPlotCorrelations(tmp, COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Residual calculation
Calculate weighted residuals and add back "Dx.Tissue" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

### Coexpression of residual expression 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
tmp = RESIDUAL.GENE_EXPRESSION
tmp[is.na(tmp)] = 0
cr = cor(t(tmp))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(tmp, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
p
```

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Institution')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(tmp)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Institution')]))
```

```{r temp1, include=F}
dev.off()
gc()
```

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:6, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('gene_id') %>% 
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(gene_lengths %>% rownameToFirstColumn('gene_id'))
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
  group_by(Comparison, Tissue) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
  group_by(Comparison, Tissue) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))

kable(full_join(tmp,tmp1))

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE %>% 
                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE %>% 
                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = percentage_gene_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gene_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE %>% 
                    arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

multiplot(plotlist = pl, cols = 3)
```

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
diffexp.genes = DE %>%
  dlply(.(Comparison, Tissue), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

# Download all related genesets from synapse
geneset.files = synGetChildren("syn3240583", includeTypes=list("file"))$asList()
genesets.to.test = purrr::map(geneset.files, .f = function(fileList){
  tmp = downloadFile(fileList$id) %>%
    dlply(.(Name), .fun = function(x){
      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
    })
  }) %>%
  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
  do.call(c,.)

# Perform enrichment analysis
enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
    }, x, backgroundGenes, .parallel = TRUE) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, genesets.to.test, backgroundGenes) %>%
  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
  dplyr::rename(Comparison = Name, Category = .id)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Set annotations
all.annotations = list(
  dataType = 'geneExpression',
  dataSubtype = 'processed',
  assay	 = 'rnaSeq',

  tissue	= 'dorsolateral prefrontal cortex, anterior cingulate cortex', 
  study = 'CMC',

  species = 'Human',
  consortium	= 'CMC',

  analysisType	= "data normalization",

  transcriptNormalizationMethod	= 'CQN',
  transcriptQuantificationMethod = 'featureCounts',
  referenceSet = 'GRCh38'
)

# Code
CODE <- Folder(name = "MSSM-Penn-Pitt - ACC and DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = synapser::File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(COV_OBJ, annotations = all.annotations)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
synSetAnnotations(COUNT_OBJ, annotations = all.annotations)

# Store logCPM
CQN.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(LCOUNT_OBJ, annotations = all.annotations)

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(OFFSET_OBJ, annotations = all.annotations)

# Store design matrix
DM1$design %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'CMC_ACC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('CMC_ACC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(DM_OBJ, annotations = all.annotations)

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(nEXP_OBJ, annotations = all.annotations)

# Store differential expression results
write.table(DE, file = 'CMC_ACC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_ACC_DLPFC_DiffExpression.tsv', 
                name = 'Differential Expression Results (Dx.Tissue)', 
                parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(DEXP_OBJ, annotations = all.annotations)

# Store enrichment results
write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                  parentId = CODE$properties$id)
ENRICH_OBJ = synStore(ENRICH_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
```
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

### Source Rmd
[Source markdown](`r thisFile`)

### 2019-12-17 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
run dlpfc samples for hbcc

---
title: "HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
author: "Kelsey Montgomery and Thanneer Perumal"
date: "`r date()`"
output: html_document
---

```{r knit2synapse, eval=FALSE}
library(synapser)
library(knit2synapse)

synLogin()

knit2synapse::createAndKnitToFolderEntity(file = "./HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
                                          parentId ="syn21435176",
                                          folderName = 'HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file

## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(lme4)
library(biomaRt)
library(ComplexHeatmap)
library(statmod)

library(synapser)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

synLogin()

library(doParallel)
library(foreach)

cl = makeCluster(detectCores()-2)

registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn21435176';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of GRCh38 feature counts with CQN normalisation (DLPFC)';

thisFileName <- 'HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd'

# Github link
thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='master')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))
```
### Data download
#### Obtain effective count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
downloadFile <- function(id){
  fread(synGet(id)$path, data.table = F)
}
downloadFile_version <- function(id , version){
  fread(synGet(id, version = version)$path, data.table = F)
}

# Get RNASeq QCmetadata
METADATA_QC_DLPFC_ID = 'syn16816488' 
ALL_USED_IDs = METADATA_QC_DLPFC_ID
metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 13) %>%
  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
                Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?',
                                      'rnaSeq_report:Mapped_Reads',
                                      'rnaSeq_report:Intragenic_Rate','rnaSeq_report:Intronic_Rate',
                                      'rnaSeq_report:Intergenic_Rate','rnaSeq_report:Genes_Detected',
                                      'rnaSeq_report:Expression_Profiling_Efficiency',
                                      'rnaSeq_report:rRNA_Rate',
                                      'rnaSeq_report:Total_Reads', 'rnaSeq_report:Percent_Aligned',
                                      'rnaSeq_report:Transcripts_Detected')) %>% 
  dplyr::filter(Institution %in% c("NIMH-HBCC"))

## Mapping file for misnamed HBCC samples 
maps <- downloadFile("syn17115622") %>% 
  dplyr::filter()

## Download reprocessed counts (DLPFC)
counts = 'syn21382250'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = counts
COUNT = downloadFile(counts) %>% data.frame()
COUNT$Chr <- NULL
COUNT$Start <- NULL
COUNT$End <- NULL
COUNT$Strand <- NULL

# Parse gene lengths
gene_lengths = COUNT[,c("Geneid", "Length")]
COUNT$Length <- NULL
gene_lengths <- rename(gene_lengths, gene_id = Geneid)

# Get ancestry vector calculated using gemtools
ANCESTRY_ID = 'syn9922992'
ALL_USED_IDs = c(ALL_USED_IDs, ANCESTRY_ID)
ANCESTRY = downloadFile(ANCESTRY_ID) %>% 
  plyr::rename(c(ID = 'SNP_report:Genotyping_Sample_ID'))

# Get genotype ids from synapse
GENOTYPE_ID = 'syn16816490'
ALL_USED_IDs = c(ALL_USED_IDs, GENOTYPE_ID)
GENOTYPE = downloadFile(GENOTYPE_ID) %>% 
  dplyr::select(Individual_ID, `SNP_report:Genotyping_Sample_ID`, `SNP_report:Exclude?`) %>% 
  dplyr::inner_join(ANCESTRY)
```

### Data preprocessing
```{r preprocess.data}
# Merge metadata
METADATA = metadata %>%
  dplyr::left_join(GENOTYPE) %>% 
  dplyr::rename(PMI = `PMI_(in_hours)`,
                RIN = `rnaSeq_isolation:RIN`, 
                ReportExclude = `rnaSeq_report:Exclude?`,
                GenotypeExclude = `SNP_report:Exclude?`,
                SampleID = Sample_RNA_ID,
                LibraryBatch = `rnaSeq_report:Library_Batch`,
                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
                MappedReads = `rnaSeq_report:Mapped_Reads`,
                IntragenicRate = `rnaSeq_report:Intragenic_Rate`, IntronicRate = `rnaSeq_report:Intronic_Rate`, 
                IntergenicRate = `rnaSeq_report:Intergenic_Rate`, GenesDetected = `rnaSeq_report:Genes_Detected`,
                ExpProfEfficiency = `rnaSeq_report:Expression_Profiling_Efficiency`, rRNARate = `rnaSeq_report:rRNA_Rate`,
                TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, Age_of_Death, PMI, Dx, 
                RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
                LibraryBatch, FlowcellBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
  dplyr::mutate(Dx = forcats::fct_recode(Dx, Other = "BP", Other = "undetermined", Control = "Control", SCZ = "SCZ"))

# Map to HBCC Sample Ids
METADATA <- dplyr::select(maps, ID, RNA_ID) %>% 
  dplyr::right_join(METADATA, by = c("RNA_ID" = "SampleID")) %>% 
  dplyr::mutate(SampleID = dplyr::coalesce(ID, RNA_ID))

# Limit HBCC DLPFC Counts
COUNT <- COUNT[,colnames(COUNT) %in% METADATA$SampleID]

ind = METADATA$SampleID[which(METADATA$ReportExclude == 1 | METADATA$GenotypeExclude)]
writeLines(paste('Following',length(ind),'samples are marked exclude'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Ethnicity) | is.na(METADATA$Institution) | is.na(METADATA$Dx)]
writeLines(paste('Following', length(ind), 'counts are missing any metadata'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$PMI)]
writeLines(paste('Following', length(ind), 'counts are missing PMI'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Reported_Gender)]
writeLines(paste('Following', length(ind), 'counts are missing gender'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$Age_of_Death)]
writeLines(paste('Following', length(ind), 'counts are missing age of death'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))

ind = METADATA$SampleID [is.na(METADATA$EV.1)]
writeLines(paste('Following', length(ind), 'counts are missing ancestry information'))
writeLines(paste(ind, collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))
```

```{r preprocess.data1, results='asis'}
# Match covariates to expression data
indToRetain = intersect(METADATA$SampleID, colnames(COUNT))
removedIDs = setdiff(colnames(COUNT), METADATA$SampleID)

COUNT = COUNT[,indToRetain]

rownames(METADATA) = METADATA$SampleID
METADATA = METADATA[indToRetain,]

METADATA %>% 
  group_by(Dx) %>% 
  summarise(count = n()) %>% 
  kable()
```
Following sample are removed `r paste(removedIDs, collapse = ',')`

```{r gene.param, include = F}
## Get GC content from biomart
backgroundGenes = data.frame(gene_id = rownames(COUNT)) %>% 
  dplyr::mutate(id = gene_id) %>% 
  tidyr::separate(id, c('ensembl_gene_id','position'), sep = '\\.')

# Define biomart object
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "ensembl.org", dataset = "hsapiens_gene_ensembl")

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>% 
  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
```
### Sex Chromosome-specific Gene Expression Patterns

```{r Reported_Gender_plots}
COUNT$gene_id = rownames(COUNT) 
REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
  left_join(COUNT) %>%
  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
  mutate(value = log(value)) %>%
  rename(`counts(log)`= value) %>% 
  rename(SampleID = item) %>%
  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
  rename(`Reported Gender` = Reported_Gender)

my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))

p = list()
p[[1]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "X"), aes(x = `Reported Gender`, y = `counts(log)`)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('X') + my.theme
p[[2]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y"), aes(x = `Reported Gender`, y = `counts(log)`)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('Y') + my.theme
multiplot(plotlist = p, cols = 2)

##XIST and UTY expression 
#ENSG00000229807.10 and ENSG00000183878.15 
FILT <- REPORTED.GENDER.COUNTS %>% 
  filter(gene_id == "ENSG00000229807.10" | gene_id == "ENSG00000183878.15") %>% 
  dplyr::select(-one_of("chromosome_name")) %>% 
  tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
  mutate(XIST = as.numeric(`ENSG00000229807.10`)) %>% 
  mutate(UTY = as.numeric(`ENSG00000183878.15`)) %>% 
  mutate(UTY = ifelse(UTY == -Inf, 0, UTY)) %>% 
  mutate(XIST = ifelse(XIST == -Inf, 0, XIST))

p = ggplot(FILT, aes (x= XIST, y = UTY)) 
p = p + geom_point(aes(color=`Reported Gender`)) + 
  ggtitle("Sex Check HBCC Cohort") + 
  theme(plot.title = element_text(hjust = 0.5, size = 15)) +
  labs(colour = "Reported Gender")
p
```

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c('Individual_ID',"Institution", "Reported_Gender", "LibraryBatch", "Dx", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
                    "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

# Find inter relation between factor covariates
COVARIATES = METADATA[,c(FactorCovariates,ContCovariates),drop=F]
COVARIATES[,FactorCovariates] <- data.frame(lapply(COVARIATES[,FactorCovariates],function(x){
  x <- sapply(x,function(y){str_replace_all(as.character(y),'[^[:alnum:]]','_')})}))
rownames(COVARIATES) <- METADATA$SampleID

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], function(x){
  x = as.numeric(as.character(gsub('[\\,\\%]','',x)))
})

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
tmp = COVARIATES.CORRELATION$ESTIMATE
tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
draw(h, heatmap_legend_side = 'left')
```

## Explore metatdata
```{r data.explore, fig.width = 12, fig.height = 12}
my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))

# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = Dx, y = RIN)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + my.theme

# Age of Death
p[[2]] = ggplot(COVARIATES, aes(x = Dx, y = Age_of_Death)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme

# PMI
p[[3]] = ggplot(COVARIATES, aes(x = Dx, y = PMI)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme

# Intronic Rate
p[[4]] = ggplot(COVARIATES, aes(x = Dx, y = IntronicRate)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme

# IntergenicRate
p[[5]] = ggplot(COVARIATES, aes(x = Dx, y = IntergenicRate)) + geom_boxplot()
p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme

# Transcripts Detected
p[[6]] = ggplot(COVARIATES, aes(x = Dx, y = IntragenicRate)) + geom_boxplot()
p[[6]] = p[[6]] + ggtitle('Intragenic Rate') + my.theme

# Mapped Reads
p[[7]] = ggplot(COVARIATES, aes(x = Dx, y = MappedReads)) + geom_boxplot()
p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme

# PercentAligned
p[[8]] = ggplot(COVARIATES, aes(x = Dx, y = TotalReads)) + geom_boxplot()
p[[8]] = p[[8]] + ggtitle('Total Reads') + my.theme

# rRNARate
p[[9]] = ggplot(COVARIATES, aes(x = Dx, y = rRNARate)) + geom_boxplot()
p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme

multiplot(plotlist = p, cols = 3)

# Institution
# vcd::mosaic(~ Institution + Dx, data = COVARIATES)

# Gender
# vcd::mosaic(~ Reported_Gender + Dx, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx and per Dx.sex. Also remove genes with missing gene length and percentage GC content

```{r cpmnormalisation}
rownames(COUNT) = COUNT$gene_id
COUNT$gene_id = NULL

genesToAnalyze = dlply(METADATA, .(Dx, Reported_Gender), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>% 
  intersect(GENE.GC.CONT$gene_id[!is.na(GENE.GC.CONT$percentage_gene_gc_content)]) %>% 
  intersect(GENE.LEN$gene_id[!is.na(GENE.LEN$Length)])

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
  dplyr::rename(gene_id = genes) %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  group_by(gene_biotype) %>%
  summarise(fraction  = n()) %>%
  filter(fraction > 100) %>%
  dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
```
`r dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]` genes are used for the analysis.

Following is the distribution of biotypes
`r kable(pct.pc)`

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")

# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  dplyr::mutate(label = SampleID)
plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
p <- p + theme_bw() %+replace% theme(legend.position="right")
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

# Plot abberent distribution of logcpm counts
tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID'))

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% indToRemove),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme_bw() %+replace% theme(legend.position = 'none') + facet_grid(Dx~., scale = 'free')
p

indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
COVARIATES = COVARIATES[indToRetain,]

tmp = COVARIATES %>%
  group_by(Dx) %>%
  summarise(count = n())
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals

Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`

Distribution of samples are:
`r kable(tmp)`

### Winsorise counts (Gene outliers in samples)
```{r winsorise.data}
LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
LOG.CPM = apply(LOG.CPM, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))

  x[x < (mn-3*std.dev)] = NA
  x[x > (mn+3*std.dev)] = NA
  return(x)
}) %>% t

# Back calculate counts for further analysis
NEW.COUNTS = PROCESSED_COUNTS$filteredExprMatrix$counts
NEW.COUNTS[is.na(LOG.CPM)] = 0
```

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
GENE.LEN = tibble::column_to_rownames(GENE.LEN, var = 'gene_id') %>% as.data.frame()
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gene_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
NEW.COUNTS[log.mat] = NA
CQN.GENE_EXPRESSION$E[log.mat] = NA
```

### Co-expression distribution
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
cr = cor(t(tmp))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
PC <- prcomp(tmp, scale.=T, center = T)

# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA MSSM/Penn/Pitt Cohort") + 
  guides(colour = guide_legend(nrow = 2),
         shape = guide_legend(nrow = 2),
         size = guide_legend(nrow = 2))
p
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)

#MDS
tmp.mds <- dist(t(tmp))
mds <- cmdscale(tmp.mds)
#Plot MDS
plotdata <- data.frame(SampleID = rownames(mds),
                       MDS1= mds[,1],
                       MDS2= mds[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(MDS1,MDS2))
p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("MDS MSSM/Penn/Pitt Cohort") + 
  guides(colour = guide_legend(nrow = 2),
         shape = guide_legend(nrow = 2),
         size = guide_legend(nrow = 2))
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
tree = hclust(as.dist(t(tmp)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp))
```
```{r temp, include = F}
dev.off()
gc()
```

### Distribution of samples (log cpm)
```{r lcpm.dist, cache=FALSE}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$E) %>%
  rownameToFirstColumn('gene_id') %>%
  tidyr::gather(SampleID, logCPM, -gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID'))

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx, scale = 'free')
p
```

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
preAdjustedSigCovars = runPCAandPlotCorrelations(tmp, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Model Identification
```{r fxn}
# Function to fit a mixed effect model and return BIC
fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){

  # Get design matrix
  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]

  # Estimate voom weights
  cnt[is.na(cnt)] = 0
  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights

  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
                              .combine = rbind,
                              .packages = c('lme4', 'dplyr'),
                              .export = c('voomWeights')) %dopar% {

                                # Perform weighted mixed linear models for every gene
                                expr = data.frame(Gx = lcpm[i,])
                                data = cbind(expr,
                                             designMat[rownames(expr),],
                                             Individual_ID = groupingVar)
                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')

                                weights = voomWeights[i,]
                                BIC = NA
                                tryCatch({
                                  mdl = lmer(formula = formula, 
                                             data = data, 
                                             weights = weights)
                                  BIC = BIC(mdl)
                                }, error = function(e){
                                  BIC = NA
                                })
                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
                                                BIC = BIC))
                            }
  return(modelBIC)
  }

# Function to select covariates in a step-wise regression fashion
selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){

  varsSelected = c()
  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
  while(length(covarsToSelect) != 0){
    allModelBIC = foreach(i = 1:length(covarsToSelect),
                          .combine = c,
                          .packages = pkgs,
                          .export = c('fitMixedEffectModel')) %dopar% {

                            # Fit a mixed effect linear model and get all BIC
                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                           covars[colnames(cnt),'Individual_ID'], 
                                                           lcpm, 
                                                           cnt, 
                                                           covars) %>%
                            plyr::rename(c('BIC' = covarsToSelect[i]))

                            return(list(modelBIC))
                          }
    allModelBIC = plyr::join_all(allModelBIC)

    # Find number of genes that are changed based on BIC
    nChange = allModelBIC %>%
      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
      dplyr::left_join(baseModelBIC) %>%
      dplyr::mutate(diff = baseModel - Value) %>%
      dplyr::group_by(ModelName) %>%
      dplyr::summarise(better = sum(diff >= 5), 
                       worse = sum(diff <= -5)) %>%
      dplyr::mutate(nchange = better - worse)

    print(nChange)

    # Select variable
    tmp = nChange %>%
      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
      dplyr::top_n(1, nchange)

    if (dim(tmp)[1] == 0){
      covarsToSelect = c()
    } else {
      baseModelBIC = allModelBIC %>%
        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
      colnames(baseModelBIC)[2] = 'baseModel'
      varsSelected = c(varsSelected, tmp$ModelName)
      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
      print(paste('Variable chosen to be added:',tmp$ModelName))
    }
  }

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('Dx')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)
# Iteratively identify covariates in phase I
phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)
# Iteratively identify covariates in phase I
phase2.covars = c('LibraryBatch', 'FlowcellBatch')
vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)
# Iteratively identify covariates in phase I
phase3.covars = c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected",
                  "ExpProfEfficiency", "rRNARate", "TotalReads")
vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`

### Normalisation
1. Dx is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect
```{r iterative.normalisation, results='asis'}
writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))

# Post adjusted design matrix
DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

# Estimate voom weights
tmp = NEW.COUNTS
tmp[is.na(tmp)] = 0
VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)

# Estimate consensus correlation coefficient
correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)

# Re-calculate voom weights with correlation of random effects
VOOM.GENE_EXPRESSION = voom(tmp, 
                            design=DM1$design, plot=F,
                            block = COVARIATES$Individual_ID, 
                            correlation = correlation$cor)

# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
tmp = RESIDUAL.GENE_EXPRESSION
tmp[is.na(tmp)] = 0
residualSigCovars = runPCAandPlotCorrelations(tmp, COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

### Coexpression of residual expression 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
tmp = RESIDUAL.GENE_EXPRESSION
tmp[is.na(tmp)] = 0
cr = cor(t(tmp))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(tmp, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx, scales = 'free_y')
p
```

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx','Institution')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(tmp)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx','Institution')]))
```

```{r temp1, include=F}
dev.off()
gc()
```

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
# Fit contrast
contrast = makeContrasts(contrasts=c("DxOther-DxControl",
                                     "DxSCZ-DxControl",
                                     "DxSCZ-DxOther"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:3, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('gene_id') %>%
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(Ensemble2HGNC) %>%
  left_join(GENE.LEN %>%
              rownameToFirstColumn('gene_id'))
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))

kable(full_join(tmp,tmp1))

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE %>% 
                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('red','grey','green'))
pl[[1]] = pl[[1]] + theme_bw() %+replace% theme(legend.position = 'top')

pl[[2]] = ggplot(DE %>% 
                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = percentage_gene_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gene_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('red','grey','green'))
pl[[2]] = pl[[2]] + theme_bw() %+replace% theme(legend.position = 'top')

pl[[3]] = ggplot(DE %>% 
                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('red','grey','green'))
pl[[3]] = pl[[3]] + theme_bw() %+replace% theme(legend.position = 'top')

multiplot(plotlist = pl, cols = 3)
```

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
diffexp.genes = DE %>%
  dlply(.(Comparison), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

# Download all related genesets from synapse
geneset.files = synGetChildren("syn3240583", includeTypes=list("file"))$asList()
genesets.to.test = purrr::map(geneset.files, .f = function(fileList){
  tmp = downloadFile(fileList$id) %>%
    dlply(.(Name), .fun = function(x){
      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
    })
  }) %>%
  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
  do.call(c,.)

# Perform enrichment analysis
enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
    }, x, backgroundGenes, .parallel = TRUE) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, genesets.to.test, backgroundGenes) %>%
  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
  dplyr::rename(Comparison = Name, Category = .id)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Set annotations
all.annotations = list(
  dataType = 'geneExpression',
  dataSubtype = 'processed',
  assay	 = 'rnaSeq',

  tissue	= 'dorsolateral prefrontal cortex, anterior cingulate cortex', 
  study = 'CMC',

  species = 'Human',
  consortium	= 'CMC',

  analysisType	= "data normalization",

  transcriptNormalizationMethod	= 'CQN',
  transcriptQuantificationMethod = 'featureCounts',
  referenceSet = 'GRCh38'
)

# Code
CODE <- Folder(name = "HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_HBCC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = synapser::File('CMC_HBCC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'covariates'
synSetAnnotations(COV_OBJ, annotations = all.annotations)

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_HBCC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_HBCC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'filteredCounts'
synSetAnnotations(COUNT_OBJ, annotations = all.annotations)

# Store logCPM
CQN.GENE_EXPRESSION$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_HBCC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('CMC_HBCC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'filteredLCPM'
synSetAnnotations(LCOUNT_OBJ, annotations = all.annotations)

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_HBCC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('CMC_HBCC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'offset'
synSetAnnotations(OFFSET_OBJ, annotations = all.annotations)

# Store design matrix
DM1$design %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'CMC_HBCC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('CMC_HBCC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'designMatrix'
synSetAnnotations(DM_OBJ, annotations = all.annotations)

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_HBCC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_HBCC_DLPFC_netResidualExpression.tsv', 
                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'residualGeneExpForNetAnlz'
synSetAnnotations(nEXP_OBJ, annotations = all.annotations)

# Store differential expression results
write.table(DE, file = 'CMC_HBCC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_HBCC_DLPFC_DiffExpression.tsv', 
                name = 'Differential Expression Results (Dx.Tissue)', 
                parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'diffExp'
synSetAnnotations(DEXP_OBJ, annotations = all.annotations)

# Store enrichment results
write.table(enrichment.results, file = 'CMC_HBCC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('CMC_HBCC_DLPFC_EnrichResults.tsv', 
                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                  parentId = CODE$properties$id)
ENRICH_OBJ = synStore(ENRICH_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
all.annotations$dataSubType = 'enrichResults'
synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

### Source Rmd
[Source markdown](`r thisFile`)

### 2019-12-17 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix file relative path

@@
 
-knit2synapse::createAndKnitToFolderEntity(file = "./HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
+knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
                                           parentId ="syn21435176",

### 2019-12-17 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix dropped gene ids in rownames
explicit specification of mart host

@@
 # Limit HBCC DLPFC Counts
+rownames(COUNT) = COUNT$Geneid
 COUNT <- COUNT[,colnames(COUNT) %in% METADATA$SampleID]
@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "ensembl.org", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "oct2018.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
 

### 2019-12-17 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
XIST ver change in new ensembl build

@@
 ##XIST and UTY expression 
-#ENSG00000229807.10 and ENSG00000183878.15 
+#ENSG00000229807.11 and ENSG00000183878.15 
 FILT <- REPORTED.GENDER.COUNTS %>% 
-  filter(gene_id == "ENSG00000229807.10" | gene_id == "ENSG00000183878.15") %>% 
+  filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>% 
   dplyr::select(-one_of("chromosome_name")) %>% 

### 2019-12-17 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
XIST ver change in new ensembl build

@@
   tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
-  mutate(XIST = as.numeric(`ENSG00000229807.10`)) %>% 
+  mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>% 
   mutate(UTY = as.numeric(`ENSG00000183878.15`)) %>% 

### 2019-12-19 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
point to updated metadata file

@@
 ALL_USED_IDs = METADATA_QC_DLPFC_ID
-metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 13) %>%
-  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
+metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 15) %>%
+  dplyr::select(`Individual_ID`, Institution, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx,
                 Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
+                                      'rnaSeq_dissection:Brain_Region',
                                       'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
@@
                                       'rnaSeq_report:Transcripts_Detected')) %>% 
-  dplyr::filter(Institution %in% c("NIMH-HBCC"))
-
-## Mapping file for misnamed HBCC samples 
-maps <- downloadFile("syn17115622") %>% 
-  dplyr::filter()
+  dplyr::filter(Institution %in% c("NIMH-HBCC")) %>% 
+  dplyr::filter(`rnaSeq_dissection:Brain_Region` %in% c("DLPFC"))
 
@@
                 TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
-  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, Age_of_Death, PMI, Dx, 
+  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
                 RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
@@
 
-# Map to HBCC Sample Ids
-METADATA <- dplyr::select(maps, ID, RNA_ID) %>% 
-  dplyr::right_join(METADATA, by = c("RNA_ID" = "SampleID")) %>% 
-  dplyr::mutate(SampleID = dplyr::coalesce(ID, RNA_ID))
-
 # Limit HBCC DLPFC Counts
@@
 METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))
+
+qc <- readRDS(synGet("syn21420622")$path)
+
 ```

### 2019-12-27 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove institution covar

--- 2019-12-17 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
 ALL_USED_IDs = METADATA_QC_DLPFC_ID
-metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 13) %>%
-  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
+metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 14) %>%
+  dplyr::select(`Individual_ID`, Institution, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
                 Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
+                                      'rnaSeq_dissection:Brain_Region',
                                       'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
@@
   dplyr::left_join(GENOTYPE) %>% 
-  dplyr::rename(PMI = `PMI_(in_hours)`,
+  dplyr::rename(Region = `rnaSeq_dissection:Brain_Region`,
+                PMI = `PMI_(in_hours)`,
                 RIN = `rnaSeq_isolation:RIN`, 
@@
 
-ind = METADATA$SampleID [is.na(METADATA$Age_of_Death)]
+ind = METADATA$SampleID [is.na(METADATA$ageOfDeath)]
 writeLines(paste('Following', length(ind), 'counts are missing age of death'))
@@
 ```{r covariates.clustering}
-FactorCovariates <- c('Individual_ID',"Institution", "Reported_Gender", "LibraryBatch", "Dx", "FlowcellBatch")
-ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
+FactorCovariates <- c('Individual_ID', "Reported_Gender", "LibraryBatch", "Dx", "FlowcellBatch")
+ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                     "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
@@
 # Age of Death
-p[[2]] = ggplot(COVARIATES, aes(x = Dx, y = Age_of_Death)) + geom_boxplot()
+p[[2]] = ggplot(COVARIATES, aes(x = Dx, y = ageOfDeath)) + geom_boxplot()
 p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme
@@
   intersect(GENE.GC.CONT$gene_id[!is.na(GENE.GC.CONT$percentage_gene_gc_content)]) %>% 
-  intersect(GENE.LEN$gene_id[!is.na(GENE.LEN$Length)])
+  intersect(gene_lengths$gene_id[!is.na(gene_lengths$Length)])
 
@@
 ```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
-indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
-                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
-                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
-                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
-                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
-                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
-                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
-                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
-                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
-                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
-                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")
+# indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
+#                 'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
+#                 'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
+#                 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
+#                 'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
+#                 "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
+#                 "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
+#                 "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
+#                 "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
+#                 "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
+#                 "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")
 
@@
                        PC2=PC$rotation[,2])
-
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
@@
 plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''
-
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
+p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
 p <- p + theme_bw() %+replace% theme(legend.position="right")
@@
 p
-
 # Plot abberent distribution of logcpm counts
@@
               rownameToFirstColumn('SampleID'))
-
-p = ggplot(tmp1 %>%
-             dplyr::filter(SampleID %in% indToRemove),
-           aes(x = logCPM, color = SampleID)) + geom_density() 
-p = p + theme_bw() %+replace% theme(legend.position = 'none') + facet_grid(Dx~., scale = 'free')
-p
+# p = ggplot(tmp1 %>%
+#              dplyr::filter(SampleID %in% indToRemove),
+#            aes(x = logCPM, color = SampleID)) + geom_density() 
+# p = p + theme_bw() %+replace% theme(legend.position = 'none') + facet_grid(Dx~., scale = 'free')
+# p
 
@@
 # Compute offset for gene length and gc content
-GENE.LEN = tibble::column_to_rownames(GENE.LEN, var = 'gene_id') %>% as.data.frame()
+gene_lengths = tibble::column_to_rownames(gene_lengths, var = 'gene_id') %>% as.data.frame()
 CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                           x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gene_gc_content'],
-                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
+                          lengths = gene_lengths[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                           lengthMethod = "smooth", 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
+p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
 p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA MSSM/Penn/Pitt Cohort") + 
@@
 p <- ggplot(plotdata, aes(MDS1,MDS2))
-p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
+p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
 p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("MDS MSSM/Penn/Pitt Cohort") + 
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx")])
+COVARIATES.tmp = data.matrix(COVARIATES[,c("Reported_Gender", "LibraryBatch", "Dx")])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
 # Iteratively identify covariates in phase I
-phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
+phase1.covars = c('Reported_Gender', 'PMI', 'ageOfDeath', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
 vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
+p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
 p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx, scales = 'free_y')
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx','Institution')])
+COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx')])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
                            main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp[,c('Dx','Institution')]))
+                           groupLabels = colnames(COVARIATES.tmp[,c('Dx')]))
 ```
@@
   left_join(Ensemble2HGNC) %>%
-  left_join(GENE.LEN %>%
+  left_join(gene_lengths %>%
               rownameToFirstColumn('gene_id'))

### 2019-12-27 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
automate outlier id - 4SDs from mean

@@
 ```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
-# indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
-#                 'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
-#                 'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
-#                 'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
-#                 'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
-#                 "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
-#                 "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
-#                 "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
-#                 "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
-#                 "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
-#                 "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")
-
 # Find principal components of expression to plot
@@
                        PC2=PC$rotation[,2])
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  dplyr::mutate(label = SampleID)
-plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''
+
+# Percentage from each PC
+eigen <- PC$sdev^2
+pc1 <- eigen[1]/sum(eigen)
+pc2 <- eigen[2]/sum(eigen)
+
+# Identify outliers - samples 4SDs from the mean
+outliers <- as.character(plotdata$SampleID[c(which(plotdata$PC1 < mean(plotdata$PC1) - 4*sd(plotdata$PC1)),
+                              which(plotdata$PC1 > mean(plotdata$PC1) + 4*sd(plotdata$PC1))), drop = T])
+
+outliers <- c(outliers, as.character(plotdata$SampleID[c(which(plotdata$PC2 < mean(plotdata$PC2) - 4*sd(plotdata$PC2)),
+                                           which(plotdata$PC2 > mean(plotdata$PC2) + 4*sd(plotdata$PC2))), drop = T] ))
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, "SampleID")) %>%
+  dplyr::mutate(label = SampleID) %>% 
+  dplyr::mutate(label = ifelse((label %in% outliers), label, NA))
+
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
@@
 
-indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
+indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), outliers)
 PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]

### 2019-12-28 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
automate outlier id - 4SDs from mean
update to sep2019 ENSEMBL host

@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "oct2018.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "sep2019.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
 
@@
 
-Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`
+Based on the expression pattern following samples were tagged as outliers: `r paste(outliers, collapse = ', ')`
 
@@
                            main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp[,c('Dx')]))
+                           groupLabels = "Dx")
 ```

### 2020-01-02 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
update metadata schema, automate outlier computation

@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = metadata
-metadata = downloadFile_version(metadata, version = 13)  %>% 
-  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, Age_of_Death, `PMI_(in_hours)`, Dx,
+metadata = downloadFile_version(metadata, version = 14)  %>% 
+  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx,
                 Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
@@
                 RIN = `rnaSeq_isolation:RIN`) %>%
-  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, Age_of_Death, PMI, Dx, 
+  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
                 RIN, ReportExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
@@
 ##XIST and UTY expression 
-#ENSG00000229807.10 and ENSG00000183878.15 
+#ENSG00000229807.11 and ENSG00000183878.15 
 FILT <- REPORTED.GENDER.COUNTS %>% 
-  filter(gene_id == "ENSG00000229807.10" | gene_id == "ENSG00000183878.15") %>% 
+  filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>% 
   dplyr::select(-one_of("chromosome_name")) %>% 
   tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
-  mutate(XIST = as.numeric(`ENSG00000229807.10`)) %>% 
+  mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>% 
   mutate(UTY = as.numeric(`ENSG00000183878.15`))
@@
 FactorCovariates <- c('Individual_ID', "Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
-ContCovariates <- c("Age_of_Death", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
+ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
                     "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads","AlignmentRate", 
@@
 # Age of Death
-p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = Age_of_Death)) + geom_boxplot()
+p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = ageOfDeath)) + geom_boxplot()
 p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme
@@
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+# Percentage from each PC
+eigen <- PC$sdev^2
+pc1 <- eigen[1]/sum(eigen)
+pc2 <- eigen[2]/sum(eigen)
+
+# Identify outliers - samples 4SDs from the mean
+outliers <- as.character(plotdata$SampleID[c(which(plotdata$PC1 < mean(plotdata$PC1) - 4*sd(plotdata$PC1)),
+                              which(plotdata$PC1 > mean(plotdata$PC1) + 4*sd(plotdata$PC1))), drop = T])
+
+outliers <- c(outliers, as.character(plotdata$SampleID[c(which(plotdata$PC2 < mean(plotdata$PC2) - 4*sd(plotdata$PC2)),
+                                           which(plotdata$PC2 > mean(plotdata$PC2) + 4*sd(plotdata$PC2))), drop = T] ))
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, "SampleID")) %>%
     tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
-  dplyr::mutate(label = SampleID)
-plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''
+  dplyr::mutate(label = SampleID) %>% 
+  dplyr::mutate(label = ifelse((label %in% outliers), label, NA))
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
@@
 p = ggplot(tmp1 %>%
-             dplyr::filter(SampleID %in% indToRemove),
+             dplyr::filter(SampleID %in% outliers),
            aes(x = logCPM, color = SampleID)) + geom_density() 
@@
 
-Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`
+Based on the expression pattern following samples were tagged as outliers: `r paste(outliers, collapse = ', ')`
+
+Based on prior evidence, investigators elected to remove: `r setdiff(indToRemove, outliers)[setdiff(indToRemove,outliers) %in% colnames(counts)]`
 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
@@
 # Iteratively identify covariates in phase I
-phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'Age_of_Death', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
+phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'ageOfDeath', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
 vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
 p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')

### 2020-01-02 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix full filepath

@@
 
-knit2synapse::createAndKnitToFolderEntity(file = "./MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
+knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
                                  parentId ="syn9872607",

### 2020-01-02 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix full filepath

@@
 thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='master')
-thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
+thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))
 ```

### 2020-01-02 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
specify ComplexHeatmap package

@@
 h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
-draw(h, heatmap_legend_side = 'left')
+ComplexHeatmap::draw(h, heatmap_legend_side = 'left')
 ```

### 2020-01-02 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
rm deprecated synapser func

@@
 COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
-COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
+COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                      executed = thisFile, activityDescription = activityDescription)
 synSetAnnotations(COUNT_OBJ, annotations = all.annotations)

### 2020-01-02 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
mod inputs for HBCC ACC and STG cohort

--- 2019-12-28 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
 ---
-title: "HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
+title: "HBCC - ACC - STG - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
 author: "Kelsey Montgomery and Thanneer Perumal"
@@
 
-knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
+knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
                                           parentId ="syn21435176",
-                                          folderName = 'HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
+                                          folderName = 'HBCC - ACC - STG - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
 ```
@@
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of GRCh38 feature counts with CQN normalisation (DLPFC)';
-
-thisFileName <- 'HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd'
+activityDescription = 'Covariate analysis of GRCh38 feature counts with CQN normalisation (ACC and STG)';
+
+thisFileName <- 'HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd'
 
@@
 # Get RNASeq QCmetadata
-METADATA_QC_DLPFC_ID = 'syn16816488' 
-ALL_USED_IDs = METADATA_QC_DLPFC_ID
-metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 14) %>%
+METADATA_QC_ID = 'syn16816488' 
+ALL_USED_IDs = METADATA_QC_ID
+metadata = downloadFile_version(METADATA_QC_ID, version = 14) %>%
   dplyr::select(`Individual_ID`, Institution, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
@@
   dplyr::filter(Institution %in% c("NIMH-HBCC")) %>% 
-  dplyr::filter(`rnaSeq_dissection:Brain_Region` %in% c("DLPFC"))
-
-## Download reprocessed counts (DLPFC)
-counts = 'syn21382250'
+  dplyr::filter(`rnaSeq_dissection:Brain_Region` %in% c("ACC","STG"))
+
+## Download reprocessed counts (ACC and STG)
+counts = 'syn21420322'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = counts
@@
 
-# Limit HBCC DLPFC Counts
+# Limit HBCC ACC and STG Counts
 rownames(COUNT) = COUNT$Geneid
@@
 
-ind = METADATA$SampleID [is.na(METADATA$Ethnicity) | is.na(METADATA$Institution) | is.na(METADATA$Dx)]
-writeLines(paste('Following', length(ind), 'counts are missing any metadata'))
+ind = METADATA$SampleID [is.na(METADATA$Ethnicity)]
+writeLines(paste('Following', length(ind), 'counts are missing ethnicity information'))
 writeLines(paste(ind, collapse = ', '))
@@
 # Code
-CODE <- Folder(name = "HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "HBCC - ACC - STG - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
 CODE <- synStore(CODE)
@@
 COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
-write.table(COVARIATES, file = 'CMC_HBCC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = synapser::File('CMC_HBCC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+write.table(COVARIATES, file = 'CMC_HBCC_ACC-STG_Covariates.tsv', sep = '\t', row.names=F, quote=F)
+COV_OBJ = synapser::File('CMC_HBCC_ACC-STG_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
 COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_HBCC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('CMC_HBCC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
+  write.table(file = 'CMC_HBCC_ACC-STG_Counts.tsv', sep = '\t', row.names=F, quote=F)
+COUNT_OBJ = File('CMC_HBCC_ACC-STG_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
 COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_HBCC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('CMC_HBCC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
+  write.table(file = 'CMC_HBCC_ACC-STG_logCPM.tsv', sep = '\t', row.names=F, quote=F)
+LCOUNT_OBJ = File('CMC_HBCC_ACC-STG_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
 LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_HBCC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('CMC_HBCC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
+  write.table(file = 'CMC_HBCC_ACC-STG_offset.tsv', sep = '\t', row.names=F, quote=F)
+OFFSET_OBJ = File('CMC_HBCC_ACC-STG_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
 OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
   rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'CMC_HBCC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('CMC_HBCC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+  write.table(file = 'CMC_HBCC_ACC-STG_Design.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('CMC_HBCC_ACC-STG_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
 DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_HBCC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_HBCC_DLPFC_netResidualExpression.tsv', 
+  write.table(file = 'CMC_HBCC_ACC-STG_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+nEXP_OBJ = File('CMC_HBCC_ACC-STG_netResidualExpression.tsv', 
                 name = 'Normalised, covariates removed residual expression (for network analysis)', 
@@
 # Store differential expression results
-write.table(DE, file = 'CMC_HBCC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_HBCC_DLPFC_DiffExpression.tsv', 
+write.table(DE, file = 'CMC_HBCC_ACC-STG_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('CMC_HBCC_ACC-STG_DiffExpression.tsv', 
                 name = 'Differential Expression Results (Dx.Tissue)', 
@@
 # Store enrichment results
-write.table(enrichment.results, file = 'CMC_HBCC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
-ENRICH_OBJ = File('CMC_HBCC_DLPFC_EnrichResults.tsv', 
+write.table(enrichment.results, file = 'CMC_HBCC_ACC-STG_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
+ENRICH_OBJ = File('CMC_HBCC_ACC-STG_EnrichResults.tsv', 
                   name = 'Enrichment Analysis Results (Dx.Tissue)', 

### 2020-01-02 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix fct_recode

@@
                 IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
-  dplyr::mutate(Dx = forcats::fct_recode(Dx, Other = "BP", Other = "undetermined", Control = "Control", SCZ = "SCZ"))
+  dplyr::mutate(Dx = forcats::fct_recode(Dx, Other = "MDD", Other = "BP", Control = "Control", SCZ = "SCZ"))
 

### 2020-01-02 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix column header

@@
 COUNT$Strand <- NULL
+# Fix counts column header - remove "X" in sample name
+new_names <- gsub("X6", "6", colnames(COUNT))
+colnames(COUNT) <- new_names
 

### 2020-01-02 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
Remove missing covariates

@@
                                       'rnaSeq_dissection:Brain_Region',
-                                      'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
+                                      'rnaSeq_isolation:RNA_Isolation_Batch',
+                                      'rnaSeq_report:Library_Batch', 
                                       'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?',
@@
                 FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
-                RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
+                IsolationBatch = `rnaSeq_isolation:RNA_Isolation_Batch`, #isolation batch to replace ribozero batch
                 MappedReads = `rnaSeq_report:Mapped_Reads`,
@@
                 RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
-                LibraryBatch, FlowcellBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
+                LibraryBatch, FlowcellBatch, IsolationBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                 IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
@@
 ```{r covariates.clustering}
-FactorCovariates <- c('Individual_ID', "Reported_Gender", "LibraryBatch", "Dx", "FlowcellBatch")
+FactorCovariates <- c('Individual_ID', "Reported_Gender", "IsolationBatch", "Dx")
 ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c("Reported_Gender", "LibraryBatch", "Dx")])
+COVARIATES.tmp = data.matrix(COVARIATES[,c("Reported_Gender", "Dx")])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
 # Iteratively identify covariates in phase I
-phase2.covars = c('LibraryBatch', 'FlowcellBatch')
+phase2.covars = c('IsolationBatch')
 vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 

### 2020-01-03 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
patch script with new md ver

@@
 ALL_USED_IDs = METADATA_QC_ID
-metadata = downloadFile_version(METADATA_QC_ID, version = 14) %>%
+metadata = downloadFile_version(METADATA_QC_ID, version = 15) %>%
   dplyr::select(`Individual_ID`, Institution, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 

### 2020-01-04 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix Dx.Tissue covar

@@
   kable()
+
+METADATA = METADATA %>%
+  dplyr::mutate(Dx.Tissue = paste(Dx, Tissue, sep = '.'))
 ```
@@
 ```{r covariates.clustering}
-FactorCovariates <- c('Individual_ID', "Reported_Gender", "IsolationBatch", "Dx")
+FactorCovariates <- c('Individual_ID', "Reported_Gender", "IsolationBatch", "Dx.Tissue")
 ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
@@
 p = list()
-p[[1]] = ggplot(COVARIATES, aes(x = Dx, y = RIN)) + geom_boxplot()
+p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot()
 p[[1]] = p[[1]] + ggtitle('RIN') + my.theme
@@
 # Age of Death
-p[[2]] = ggplot(COVARIATES, aes(x = Dx, y = ageOfDeath)) + geom_boxplot()
+p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = ageOfDeath)) + geom_boxplot()
 p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme
@@
 # PMI
-p[[3]] = ggplot(COVARIATES, aes(x = Dx, y = PMI)) + geom_boxplot()
+p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot()
 p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme
@@
 # Intronic Rate
-p[[4]] = ggplot(COVARIATES, aes(x = Dx, y = IntronicRate)) + geom_boxplot()
+p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot()
 p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme
@@
 # IntergenicRate
-p[[5]] = ggplot(COVARIATES, aes(x = Dx, y = IntergenicRate)) + geom_boxplot()
+p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot()
 p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme
@@
 # Transcripts Detected
-p[[6]] = ggplot(COVARIATES, aes(x = Dx, y = IntragenicRate)) + geom_boxplot()
+p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntragenicRate)) + geom_boxplot()
 p[[6]] = p[[6]] + ggtitle('Intragenic Rate') + my.theme
@@
 # Mapped Reads
-p[[7]] = ggplot(COVARIATES, aes(x = Dx, y = MappedReads)) + geom_boxplot()
+p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot()
 p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme
@@
 # PercentAligned
-p[[8]] = ggplot(COVARIATES, aes(x = Dx, y = TotalReads)) + geom_boxplot()
+p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TotalReads)) + geom_boxplot()
 p[[8]] = p[[8]] + ggtitle('Total Reads') + my.theme
@@
 # rRNARate
-p[[9]] = ggplot(COVARIATES, aes(x = Dx, y = rRNARate)) + geom_boxplot()
+p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot()
 p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme
@@
 ### Filter genes
-Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx and per Dx.sex. Also remove genes with missing gene length and percentage GC content
+Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue and Reported Gender. Also remove genes with missing gene length and percentage GC content
 
@@
 
-genesToAnalyze = dlply(METADATA, .(Dx, Reported_Gender), .fun = function(mtd, count){
+genesToAnalyze = dlply(METADATA, .(Dx.Tissue, Reported_Gender), .fun = function(mtd, count){
   processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
+p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
 p <- p + theme_bw() %+replace% theme(legend.position="right")
@@
 
-tmp = COVARIATES %>%
-  group_by(Dx) %>%
-  summarise(count = n())
+tmp = tidyr::separate(COVARIATES, Dx.Tissue, c('Dx', 'Tissue'), sep = '_') %>%
+  group_by(Dx, Tissue) %>%
+  summarise(count = n()) %>%
+  spread(Tissue, count)
 ```
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
+p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
 p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA MSSM/Penn/Pitt Cohort") + 
@@
 p <- ggplot(plotdata, aes(MDS1,MDS2))
-p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
+p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
 p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("MDS MSSM/Penn/Pitt Cohort") + 
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c("Reported_Gender", "Dx")])
+COVARIATES.tmp = data.matrix(COVARIATES[,c("Reported_Gender", "Dx.Tissue")])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
 p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
-p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx, scale = 'free')
+p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx.Tissue, scale = 'free')
 p
@@
 # Fit a base model with the following covariates
-adjust.covars = c('Dx')
+adjust.covars = c('Dx.Tissue')
 baseModelBIC = fitMixedEffectModel(adjust.covars, 
@@
 ### Normalisation
-1. Dx is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
+1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
 2. Individual_ID is chosen as random effect
@@
 ### Residual calculation
-Calculate weighted residuals and add back "Dx" to the residuals
+Calculate weighted residuals and add back "Dx.Tissue" to the residuals
 ```{r varsToAddBack}
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx, scales = 'free_y')
+p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
+p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx.Tissue, scales = 'free_y')
 p
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx')])
+COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue')])
 COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
@@
                            main = "Sample dendrogram",
-                           groupLabels = "Dx")
+                           groupLabels = "Dx.Tissue")
 ```
@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("DxOther-DxControl",
-                                     "DxSCZ-DxControl",
-                                     "DxSCZ-DxOther"),
+contrast = makeContrasts(contrasts=c("Dx.TissueOther-Dx.TissueControl",
+                                     "Dx.TissueSCZ-Dx.TissueControl",
+                                     "Dx.TissueSCZ-Dx.TissueOther"),
                          levels = colnames(ADJUSTED.FIT$coefficients))
@@
   rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('Dx','',Comparison),
+  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
                 Direction = logFC/abs(logFC),
@@
 
-  tissue	= 'dorsolateral prefrontal cortex, anterior cingulate cortex', 
+  tissue	= 'superior temporal gyrus, anterior cingulate cortex', 
   study = 'CMC',

### 2020-01-04 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix Dx.Tissue covar

@@
   dplyr::left_join(GENOTYPE) %>% 
-  dplyr::rename(Region = `rnaSeq_dissection:Brain_Region`,
+  dplyr::rename(Tissue = `rnaSeq_dissection:Brain_Region`,
                 PMI = `PMI_(in_hours)`,

### 2020-01-04 ampad-rnaseq-de-clean/rosmap_cohort.R
specify dplyr::

@@ -3,12 +3,12 @@
 ############################
 # Get and Join/Bind Inputs #
 ############################
-
+Sys.setenv(R_CONFIG_ACTIVE = "default")
 # Join synIDs and select sample identifiers to map sequencing metadata
 sequencing_metadata <- get_data(config::get("sequencing metadata")$synID,
                                 config::get("sequencing metadata")$version) %>%
   full_join(get_data(config::get("synID mapping")$synID), by = c("sample" = "id")) %>%
-  select(specimenID, everything(), -sample, -versionNumber) %>%
+  dplyr::select(specimenID, everything(), -sample, -versionNumber) %>%
   distinct()
 
 # Remove one version of duplicate counts "syn4212581", "syn4212582"
@@ -19,7 +19,7 @@ counts <- get_data(config::get("counts")$synID)[-c(1:4),] %>%
   as.data.frame() %>%
   tibble::rownames_to_column(var = "sample") %>%
   left_join(get_data(config::get("synID mapping")$synID), by = c("sample" = "id")) %>%
-  select(-sample, -versionNumber) %>%
+  dplyr::select(-sample, -versionNumber) %>%
   tibble::column_to_rownames(var = "specimenID") %>%
   t()
 
@@ -49,12 +49,12 @@ metadata <- get_data(config::get("assay metadata")$synID,
 #######################
 
 # Identify variables with missing data
-removed_samples <- metadata$Sampleid[is.na(metadata$cogdx)|
-                                 is.na(metadata$braaksc)|
-                                 is.na(metadata$ceradsc)|
-                                 is.na(metadata$RINcontinuous)|
-                                 is.na(metadata$pmi)|
-                                 is.na(metadata$RnaSeqMetrics__INTRONIC_BASES)|
+removed_samples <- metadata$Sampleid[is.na(metadata$cogdx) |
+                                 is.na(metadata$braaksc) |
+                                 is.na(metadata$ceradsc) |
+                                 is.na(metadata$RINcontinuous) |
+                                 is.na(metadata$pmi) |
+                                 is.na(metadata$RnaSeqMetrics__INTRONIC_BASES) |
                                  is.na(metadata$age_death)]
 
 # Pick higher quality RIN batch

### 2020-01-04 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix Dx.Tissue covar

@@
                 TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
-  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
+  dplyr::select(SampleID, Individual_ID, Institution, Tissue, Reported_Gender, Sex, Ethnicity, ageOfDeath, PMI, Dx, 
                 RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,

### 2020-01-04 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix date

@@
 author: "Kelsey Montgomery"
-date: "11/8/2019"
+date: "`r date()`"
 output: html_document

### 2020-01-05 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix contrasts

@@
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueOther-Dx.TissueControl",
-                                     "Dx.TissueSCZ-Dx.TissueControl",
-                                     "Dx.TissueSCZ-Dx.TissueOther"),
+contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
+                                     "Dx.TissueOther_STG-Dx.TissueControl_STG",
+                                     "Dx.TissueSCZ_STG-Dx.TissueControl_STG",
+                                     "Dx.TissueSCZ_STG-Dx.TissueOther_STG"),
                          levels = colnames(ADJUSTED.FIT$coefficients))

### 2020-01-05 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix contrasts

@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, "SampleID")) %>%
+    tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
   dplyr::mutate(label = SampleID) %>% 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="right")
+p <- p + geom_point(aes(shape=Tissue, size=ageOfDeath))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
 p <- p + geom_text(aes(label= label), size=4, hjust=0)
 p
+
 # Plot abberent distribution of logcpm counts
@@
   left_join(COVARIATES %>%
-              rownameToFirstColumn('SampleID'))
-# p = ggplot(tmp1 %>%
-#              dplyr::filter(SampleID %in% indToRemove),
-#            aes(x = logCPM, color = SampleID)) + geom_density() 
-# p = p + theme_bw() %+replace% theme(legend.position = 'none') + facet_grid(Dx~., scale = 'free')
-# p
+              rownameToFirstColumn('SampleID') %>%
+              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))
+
+p = ggplot(tmp1 %>%
+             dplyr::filter(SampleID %in% outliers),
+           aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
+p
 
@@
 p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA MSSM/Penn/Pitt Cohort") + 
+p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA HBCC ACC-STG Cohort") + 
   guides(colour = guide_legend(nrow = 2),
@@
 p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("MDS MSSM/Penn/Pitt Cohort") + 
+p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("MDS HBCC ACC-STG Cohort") + 
   guides(colour = guide_legend(nrow = 2),
@@
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
 

### 2020-01-06 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix contrasts

@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx.Tissue, scales = 'free_y')
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
 p

### 2020-01-06 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix contrasts

--- 2020-01-05 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx.Tissue, scales = 'free_y')
+p <- p + geom_point(aes(shape=Tissue, size=ageOfDeath))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
 p

### 2020-01-06 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
note DREAM

--- 2020-01-02 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
 ---
-title: "MSSM-Penn-Pitt - ACC and DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
+title: "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
 author: "Kelsey Montgomery"
-date: "11/8/2019"
+date: "`r date()`"
 output: html_document
@@
                                  parentId ="syn9872607",
-                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model")
+                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model")
 ```
@@
 # Code
-CODE <- Folder(name = "MSSM-Penn-Pitt - ACC and DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
 CODE <- synStore(CODE)

### 2020-01-07 cmc-rnaseq-de/reproduce_dream.R
repro dream

@@ -1,0 +1,20 @@
+library("tibble")
+synLogin()
+param = BiocParallel::SnowParam(4, "SOCK", progressbar=TRUE)
+BiocParallel::register(param)
+
+metadata <- readr::read_csv(synGet("syn21488871")$path) %>% 
+  tibble::column_to_rownames(var = "SampleID")
+
+counts <- readr::read_csv(synGet("syn21488882")$path) %>% 
+  tibble::column_to_rownames(var = "geneID") %>% 
+  as.matrix()
+
+tmp = counts
+tmp[is.na(tmp)] = 0
+
+dream_formula <- ~ Dx.Tissue + (1|Individual_ID)
+
+vobjDream <- variancePartition::voomWithDreamWeights(counts = tmp, formula = dream_formula, data = metadata)
+
+

### 2020-01-16 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix date
add DREAM config to DE analysis

@@
                                   paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')
-
+                              writeLines(paste("This is the formula:",formula))
                                 weights = voomWeights[i,]
@@
 ```{r phase1}
-# Fit a base model with the following covariates
+# Fit a base model with the following covariates varsToFit, groupingVar, lcpm, cnt, covars
 adjust.covars = c('Dx.Tissue')
@@
 
-# Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
-DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
-# Estimate voom weights
+# Post adjusted formula
+random_effect <- "Individual_ID"
+formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+
+# Estimate voom weights with DREAM
 tmp = NEW.COUNTS
 tmp[is.na(tmp)] = 0
-VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)
-
-# Estimate consensus correlation coefficient
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION = voom(tmp, 
-                            design=DM1$design, plot=F,
-                            block = COVARIATES$Individual_ID, 
-                            correlation = correlation$cor)
+geneExpr = DGEList(tmp)
+geneExpr = edgeR::calcNormFactors(geneExpr)
+
+VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = tmp, 
+                                                               formula = formula,
+                                                               data = COVARIATES )
 
 # Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
-
-# Residuals after normalisation
+# CQN NA counts to 0 counts for input to dream()
+tmp = CQN.GENE_EXPRESSION$E
+tmp[is.na(tmp)] = 0
+VOOM.GENE_EXPRESSION$E = tmp
+ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
+                                        formula = formula,
+                                        data = COVARIATES)
+
+# Residuals after normalisation - residuals.MArrayLM() does not take MArrayLM2 objects
 RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
@@
 ```{r diffExp, fig.height=10, fig.width=15}
+
+formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+
+L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueOther_ACC", "Dx.TissueControl_ACC"))
+L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueControl_ACC"))
+L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueOther_ACC"))
+L4 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueOther_DLPFC", "Dx.TissueControl_DLPFC"))
+L5 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueControl_DLPFC"))
+
+L6 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueOther_DLPFC"))
+L = cbind(L1, L2, L3, L4, L5, L6)
+
+# Visualize contrast matrix
+plotContrasts(L)
+
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
-                                     "Dx.TissueOther_DLPFC-Dx.TissueControl_DLPFC",
-                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
-                                     "Dx.TissueSCZ_DLPFC-Dx.TissueOther_DLPFC"),
-                         levels = colnames(ADJUSTED.FIT$coefficients))
-FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
-FIT.CONTR = eBayes(FIT.CONTR)
-
-# Get differnetial expression
-DE = lapply(1:6, function(i, FIT){
-  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
+                                     formula = formula,
+                                     data = COVARIATES,
+                                     L = L)
+
+# Get differential expression
+DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
+  topTable(FIT, coef=i, number = 50000) %>%
     rownameToFirstColumn('gene_id') %>% 

### 2020-01-17 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove arbitrary cutoff number

@@
 DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
-  topTable(FIT, coef=i, number = 50000) %>%
+  topTable(FIT, coef=i, number = Inf) %>%
     rownameToFirstColumn('gene_id') %>% 

### 2020-01-19 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
increment to dream 1.17.3

@@
 library(biomaRt)
-
+devtools::install_github("GabrielHoffman/variancePartition@3287d7b")
+library(variancePartition)
 library(synapser)
@@
 
-VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = tmp, 
+VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
                                                                formula = formula,
-                                                               data = COVARIATES )
+                                                               data = COVARIATES)
 
@@
                                         formula = formula,
-                                        data = COVARIATES)
+                                        data = COVARIATES,
+                                        computeResiduals = TRUE)
 

### 2020-01-21 cmc-rnaseq-de/reproduce_dream.R
code to repo dream 1.17.3

@@ -1,20 +1,44 @@
-library("tibble")
+devtools::install_github("GabrielHoffman/variancePartition@3287d7b")
+library(variancePartition)
+library(tibble)
+library(synapser)
 synLogin()
 param = BiocParallel::SnowParam(4, "SOCK", progressbar=TRUE)
 BiocParallel::register(param)
 
-metadata <- readr::read_csv(synGet("syn21488871")$path) %>% 
+COVARIATES <- readr::read_csv(synGet("syn21488871")$path) %>% 
   tibble::column_to_rownames(var = "SampleID")
 
-counts <- readr::read_csv(synGet("syn21488882")$path) %>% 
-  tibble::column_to_rownames(var = "geneID") %>% 
-  as.matrix()
+NEW.COUNTS <- readRDS(synGet("syn21535143")$path)
 
-tmp = counts
+CQN.GENE_EXPRESSION <- readRDS(synGet("syn21535146")$path)
+
+# Post adjusted formula
+random_effect <- "Individual_ID"
+adjust.covars <- c("Dx.Tissue", "RIN")
+formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+
+tmp = NEW.COUNTS
 tmp[is.na(tmp)] = 0
+geneExpr = DGEList(tmp)
+geneExpr = edgeR::calcNormFactors(geneExpr)
 
-dream_formula <- ~ Dx.Tissue + (1|Individual_ID)
+VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
+                                                               formula = formula,
+                                                               data = COVARIATES)
 
-vobjDream <- variancePartition::voomWithDreamWeights(counts = tmp, formula = dream_formula, data = metadata)
+# Fit linear model using new weights and new design
+# CQN NA counts to 0 counts for input to dream()
+tmp = CQN.GENE_EXPRESSION$E
+tmp[is.na(tmp)] = 0
+VOOM.GENE_EXPRESSION$E = tmp
+ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
+                                        formula = formula,
+                                        data = COVARIATES,
+                                        computeResiduals = TRUE)
+
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+
 
 

### 2020-01-24 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
dream 1.17.5 to compute residuals

@@
 library(biomaRt)
-devtools::install_github("GabrielHoffman/variancePartition@3287d7b")
+#devtools::install_github("GabrielHoffman/variancePartition@613dfe9")
 library(variancePartition)
@@
 # Residuals after normalisation - residuals.MArrayLM() does not take MArrayLM2 objects
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
 
@@
 # Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
+varsToAddIn = grep("Dx.Tissue", colnames(ADJUSTED.FIT$design), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
 ```
@@
 
+contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
+              glue::glue_collapse(names(L2[L2 != 0]), "-"),
+              glue::glue_collapse(names(L3[L3 != 0]), "-"),
+              glue::glue_collapse(names(L4[L4 != 0]), "-"),
+              glue::glue_collapse(names(L5[L5 != 0]), "-"),
+              glue::glue_collapse(names(L6[L6 != 0]), "-"))
+
 # Visualize contrast matrix
@@
 }, FIT.CONTR) 
-names(DE) = colnames(contrast)
+names(DE) <- contrast
 

### 2020-01-24 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
add code chunk to compute cor between tissues

@@
 
+``` {r cor.repeat.measures}
+repeated <- COVARIATES %>%
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>% 
+  dplyr::group_by(Individual_ID, Tissue) %>% 
+  dplyr::summarize(Samples = n()) %>% 
+  dplyr::filter(Samples == 1) %>% 
+  dplyr::group_by(Individual_ID) %>% 
+  dplyr::summarize(multiSample_perTissue = n()) %>% 
+  dplyr::filter(multiSample_perTissue == 2)
+
+glue::glue("{length(unique(repeated$Individual_ID))} of {length(unique(COVARIATES$Individual_ID))} have repeat measurements")
+
+quant_cor <- function(count_matrix, metadata, repeated_vec){
+  correlations <- tribble(~ iid, ~ cor)
+  for (i in repeated_vec){
+     samps <- metadata$SampleID[metadata$Individual_ID %in% i]
+     cors <- cor(count_matrix[,samps[1]], count_matrix[,samps[2]])
+     entry <- tribble(~ iid, ~ cor,
+                      as.character(i), cors)
+     correlations <- bind_rows(correlations,entry)
+  }
+  correlations
+}
+
+tmp = NEW.COUNTS
+tmp[is.na(tmp)] = 0
+
+correlation_tissue <- quant_cor(tmp, METADATA, repeated$Individual_ID)
+```
+
 ### Sanity check
@@
 synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
+
+# Store correlation between tissues
+write.table(correlation_tissue, file = 'CMC_ACC_DLPFC_correlationTissue.tsv', sep = '\t', row.names = FALSE, quote = FALSE)
+COR_OBJ = File('CMC_ACC_DLPC_correlationTissue.tsv',
+               name = 'Correlation between Tissues, Same Individual',
+               parentId = CODE$properties$id)
+COR_OBJ = synStore(COR_OBJ, used = ALL_USED_IDs, activityName = activityName,
+                   executed = thisFile, activityDescription = acactivityDescription)
+synSetAnnotations(COR_OBJ, annotations = all.annotations)
 ```

### 2020-01-24 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
note branch

@@
 # Github link
-thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='master')
+thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='compute-tissue-cor')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))

### 2020-01-24 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
removed samples noted per clinical variable

@@
 counts$Length <- NULL
-gene_lengths <- rename(gene_lengths, gene_id = Geneid)
+gene_lengths <- dplyr::rename(gene_lengths, gene_id = Geneid)
 
@@
   group_by(Dx, Tissue) %>% 
-  summarise(count = n()) %>% 
+  dplyr::summarise(count = n()) %>% 
   spread(Tissue, count) %>%
@@
 ```
-Following sample are removed `r paste(removedIDs, collapse = ',')`
 

### 2020-01-24 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
note dplyr

@@
   mutate(value = log(value)) %>%
-  rename(`counts(log)`= value) %>% 
-  rename(SampleID = item) %>%
+  dplyr::rename(`counts(log)`= value) %>% 
+  dplyr::rename(SampleID = item) %>%
   left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
-  rename(`Reported Gender` = Reported_Gender)
+  dplyr::rename(`Reported Gender` = Reported_Gender)
 

### 2020-01-27 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
specify packages and design matrix output

@@
   group_by(gene_biotype) %>%
-  summarise(fraction  = n()) %>%
+  dplyr::summarise(fraction  = n()) %>%
   filter(fraction > 100) %>%
@@
   group_by(Dx, Tissue) %>%
-  summarise(count = n()) %>%
+  dplyr::summarise(count = n()) %>%
   spread(Tissue, count)
@@
 ```{r diffExp, fig.height=10, fig.width=15}
-
 formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
@@
 # Visualize contrast matrix
-plotContrasts(L)
+variancePartition::plotContrasts(L)
 
@@
 }, FIT.CONTR)
+```
+
+```{r testchunk}
 names(DE) <- contrast
@@
 # Store design matrix
-DM1$design %>%
+FIT.CONTR$design %>%
   rownameToFirstColumn('SampleID') %>%

### 2020-01-27 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix typo

@@
 COR_OBJ = synStore(COR_OBJ, used = ALL_USED_IDs, activityName = activityName,
-                   executed = thisFile, activityDescription = acactivityDescription)
+                   executed = thisFile, activityDescription = activityDescription)
 synSetAnnotations(COR_OBJ, annotations = all.annotations)

### 2020-01-27 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix parentId

@@
 knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
-                                 parentId ="syn9872607",
+                                 parentId ="syn21435176",
                                  folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model")

### 2020-01-27 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix typo

@@
 write.table(correlation_tissue, file = 'CMC_ACC_DLPFC_correlationTissue.tsv', sep = '\t', row.names = FALSE, quote = FALSE)
-COR_OBJ = File('CMC_ACC_DLPC_correlationTissue.tsv',
+COR_OBJ = File('CMC_ACC_DLPFC_correlationTissue.tsv',
                name = 'Correlation between Tissues, Same Individual',

### 2020-01-27 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix typo in markdown format

--- 2020-01-27 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
 synSetAnnotations(COR_OBJ, annotations = all.annotations)
-```
-```
+
+```
+```
+
 |  *Results*                                  |  *SynapseID*                     |
@@
 |  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |
+|  Correlation between Tissues                |  `r COR_OBJ$properties$id`       |
 

### 2020-01-27 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
new cohort

--- 2019-12-28 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
 ---
-title: "HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
+title: "HBCC - ACC-Nirmala - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
 author: "Kelsey Montgomery and Thanneer Perumal"
@@
 
-knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
+knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
                                           parentId ="syn21435176",
-                                          folderName = 'HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
+                                          folderName = 'HBCC - ACC-Nirmala - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
 ```
@@
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of GRCh38 feature counts with CQN normalisation (DLPFC)';
-
-thisFileName <- 'HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd'
+activityDescription = 'Covariate analysis of GRCh38 feature counts with CQN normalisation (ACC-Nirmala)';
+
+thisFileName <- 'HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd'
 
@@
 # Get RNASeq QCmetadata
-METADATA_QC_DLPFC_ID = 'syn16816488' 
-ALL_USED_IDs = METADATA_QC_DLPFC_ID
-metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 14) %>%
+METADATA_QC_ACC_ID = 'syn16816488' 
+ALL_USED_IDs = METADATA_QC_ACC_ID
+metadata = downloadFile_version(METADATA_QC_ACC_ID, version = 16) %>%
   dplyr::select(`Individual_ID`, Institution, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
-                Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
+                Sample_RNA_ID, one_of('Cohort',
+                                      'rnaSeq_isolation:RIN',
                                       'rnaSeq_dissection:Brain_Region',
@@
   dplyr::filter(Institution %in% c("NIMH-HBCC")) %>% 
-  dplyr::filter(`rnaSeq_dissection:Brain_Region` %in% c("DLPFC"))
-
-## Download reprocessed counts (DLPFC)
-counts = 'syn21382250'
+  dplyr::filter(`rnaSeq_dissection:Brain_Region` %in% c("ACC")) %>% 
+  dplyr::filter(Cohort %in% c("NIMH-HBCC-5"))
+
+## Download reprocessed counts (ACC)
+counts = 'syn21444441'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = counts
@@
 COUNT$Length <- NULL
-gene_lengths <- rename(gene_lengths, gene_id = Geneid)
+gene_lengths <- dplyr::rename(gene_lengths, gene_id = Geneid)
 
@@
                 IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
-  dplyr::mutate(Dx = forcats::fct_recode(Dx, Other = "BP", Other = "undetermined", Control = "Control", SCZ = "SCZ"))
-
-# Limit HBCC DLPFC Counts
+  dplyr::mutate(Dx = forcats::fct_recode(Dx, Other = "BP", Control = "Control", SCZ = "SCZ"))
+
+# Limit HBCC ACC Counts
 rownames(COUNT) = COUNT$Geneid
@@
   group_by(Dx) %>% 
-  summarise(count = n()) %>% 
+  dplyr::summarise(count = n()) %>% 
   kable()
@@
   group_by(Dx) %>%
-  summarise(count = n())
+  dplyr::summarise(count = n())
 ```
@@
 # Code
-CODE <- Folder(name = "HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "HBCC - ACC-Nirmala - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
 CODE <- synStore(CODE)
@@
 COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
-write.table(COVARIATES, file = 'CMC_HBCC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = synapser::File('CMC_HBCC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+write.table(COVARIATES, file = 'CMC_HBCC_ACC-Nirmala_Covariates.tsv', sep = '\t', row.names=F, quote=F)
+COV_OBJ = synapser::File('CMC_HBCC_ACC-Nirmala_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
 COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_HBCC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('CMC_HBCC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
+  write.table(file = 'CMC_HBCC_ACC-Nirmala_Counts.tsv', sep = '\t', row.names=F, quote=F)
+COUNT_OBJ = File('CMC_HBCC_ACC-Nirmala_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
 COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_HBCC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('CMC_HBCC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
+  write.table(file = 'CMC_HBCC_ACC-Nirmala_logCPM.tsv', sep = '\t', row.names=F, quote=F)
+LCOUNT_OBJ = File('CMC_HBCC_ACC-Nirmala_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
 LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_HBCC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('CMC_HBCC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
+  write.table(file = 'CMC_HBCC_ACC-Nirmala_offset.tsv', sep = '\t', row.names=F, quote=F)
+OFFSET_OBJ = File('CMC_HBCC_ACC-Nirmala_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
 OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
   rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'CMC_HBCC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('CMC_HBCC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
+  write.table(file = 'CMC_HBCC_ACC-Nirmala_Design.tsv', sep = '\t', row.names=F, quote=F)
+DM_OBJ = File('CMC_HBCC_ACC-Nirmala_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
 DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_HBCC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_HBCC_DLPFC_netResidualExpression.tsv', 
+  write.table(file = 'CMC_HBCC_ACC-Nirmala_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+nEXP_OBJ = File('CMC_HBCC_ACC-Nirmala_netResidualExpression.tsv', 
                 name = 'Normalised, covariates removed residual expression (for network analysis)', 
@@
 # Store differential expression results
-write.table(DE, file = 'CMC_HBCC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_HBCC_DLPFC_DiffExpression.tsv', 
+write.table(DE, file = 'CMC_HBCC_ACC-Nirmala_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('CMC_HBCC_ACC-Nirmala_DiffExpression.tsv', 
                 name = 'Differential Expression Results (Dx.Tissue)', 
@@
 # Store enrichment results
-write.table(enrichment.results, file = 'CMC_HBCC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
-ENRICH_OBJ = File('CMC_HBCC_DLPFC_EnrichResults.tsv', 
+write.table(enrichment.results, file = 'CMC_HBCC_ACC-Nirmala_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
+ENRICH_OBJ = File('CMC_HBCC_ACC-Nirmala_EnrichResults.tsv', 
                   name = 'Enrichment Analysis Results (Dx.Tissue)', 

### 2020-01-27 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix covars

@@
                                       'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
-                                      'rnaSeq_report:Flowcell_Batch', 'rnaSeq_report:Exclude?',
+                                      'rnaSeq_report:RNA_Isolation_Batch', 'rnaSeq_report:Exclude?',
                                       'rnaSeq_report:Mapped_Reads',
@@
                 LibraryBatch = `rnaSeq_report:Library_Batch`,
-                FlowcellBatch = `rnaSeq_report:Flowcell_Batch`,
+                IsolationBatch = `rnaSeq_report:RNA_Isolation_Batch`,
                 RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
@@
 ```{r covariates.clustering}
-FactorCovariates <- c('Individual_ID', "Reported_Gender", "LibraryBatch", "Dx", "FlowcellBatch")
+FactorCovariates <- c('Individual_ID', "Reported_Gender", "LibraryBatch", "Dx", "IsolationBatch")
 ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
@@
 ```{r covariates.correlation, fig.width=10, fig.height=8}
-COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
+COVARIATES.CORRELATION = CovariateAnalysis::getAssociationStatistics(COVARIATES, PVAL = 0.05)
 tmp = COVARIATES.CORRELATION$ESTIMATE
@@
 # Iteratively identify covariates in phase I
-phase2.covars = c('LibraryBatch', 'FlowcellBatch')
+phase2.covars = c('LibraryBatch', 'Isolation Batch')
 vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 

### 2020-01-27 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix covars

@@
                                       'rnaSeq_report:Ribozero_Batch', 'rnaSeq_report:Library_Batch', 
-                                      'rnaSeq_report:RNA_Isolation_Batch', 'rnaSeq_report:Exclude?',
+                                      'rnaSeq_isolation:RNA_Isolation_Batch', 'rnaSeq_report:Exclude?',
                                       'rnaSeq_report:Mapped_Reads',
@@
                 LibraryBatch = `rnaSeq_report:Library_Batch`,
-                IsolationBatch = `rnaSeq_report:RNA_Isolation_Batch`,
+                IsolationBatch = `rnaSeq_isolation:RNA_Isolation_Batch`,
                 RibozeroBatch = `rnaSeq_report:Ribozero_Batch`,
@@
                 RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
-                LibraryBatch, FlowcellBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
+                LibraryBatch, IsolationBatch, RibozeroBatch, MappedReads, TotalReads, GenesDetected, AlignmentRate,
                 IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%

### 2020-01-30 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
exclude missing RIN
typo

@@
 
+ind = METADATA$SampleID [is.na(METADATA$RIN)]
+writeLines(paste('Following', length(ind), 'counts are missing RIN'))
+writeLines(paste(ind, collapse = ', '))
+METADATA <- METADATA  %>% dplyr::filter(!(SampleID %in% ind))
+
 ind = METADATA$SampleID [is.na(METADATA$EV.1)]
@@
 # Iteratively identify covariates in phase I
-phase2.covars = c('LibraryBatch', 'Isolation Batch')
+phase2.covars = c('LibraryBatch', 'IsolationBatch')
 vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
@@
 # Estimate consensus correlation coefficient
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
+correlation <- CovariateAnalysis::parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
 

### 2020-01-30 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix fixed effect model

@@
 p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA MSSM/Penn/Pitt Cohort") + 
+p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA HBCC Cohort") + 
   guides(colour = guide_legend(nrow = 2),
@@
 p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("MDS MSSM/Penn/Pitt Cohort") + 
+p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("HBCC Cohort") + 
   guides(colour = guide_legend(nrow = 2),
@@
 ```{r fxn}
-# Function to fit a mixed effect model and return BIC
-fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){
+fitFixedEffectModel <- function(varsToFit, lcpm, cnt, covars){
 
@@
 
-  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
-                              .combine = rbind,
-                              .packages = c('lme4', 'dplyr'),
-                              .export = c('voomWeights')) %dopar% {
-
-                                # Perform weighted mixed linear models for every gene
-                                expr = data.frame(Gx = lcpm[i,])
-                                data = cbind(expr,
-                                             designMat[rownames(expr),],
-                                             Individual_ID = groupingVar)
-                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
-                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')
-
-                                weights = voomWeights[i,]
-                                BIC = NA
-                                tryCatch({
-                                  mdl = lmer(formula = formula, 
+  modelBIC = plyr::ldply(1:dim(lcpm)[1],
+                         .fun = function(i, lcpm, designMat, voomWeights){
+                           # Fit a linear models for every gene
+                           data = cbind(data.frame(Gx = lcpm[i,]),
+                                        designMat, 
+                                        weight = voomWeights[i,])
+
+                           formula = paste(setdiff(colnames(data), c('Gx', 'weight')), collapse = '+') %>%
+                             paste('Gx ~',., collapse = ' ')
+
+                           mdl = stats::lm(formula = formula, 
                                            data = data, 
-                                             weights = weights)
-                                  BIC = BIC(mdl)
-                                }, error = function(e){
-                                  BIC = NA
-                                })
-                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
-                                                BIC = BIC))
-                            }
+                                           weights = weight)
+
+                           data.frame(ensembl_gene_id = rownames(lcpm)[i], BIC = BIC(mdl))
+                         }, 
+                         lcpm, designMat, voomWeights, 
+                         .parallel = F,
+                         .paropts = list(.packages = c('dplyr')))
   return(modelBIC)
@@
 
+# vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
+#                                      COVARIATES, CQN.GENE_EXPRESSION$E, 
+#                                      NEW.COUNTS, baseModelBIC)
 # Function to select covariates in a step-wise regression fashion
@@
                           .packages = pkgs,
-                          .export = c('fitMixedEffectModel')) %dopar% {
+                          .export = c('fitFixedEffectModel')) %dopar% {
 
                             # Fit a mixed effect linear model and get all BIC
-                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
-                                                           covars[colnames(cnt),'Individual_ID'], 
+                            modelBIC = fitFixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                            lcpm, 
@@
 adjust.covars = c('Dx')
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
+baseModelBIC = fitFixedEffectModel(adjust.covars, 
                                    CQN.GENE_EXPRESSION$E, 
@@
 # Fit a base model with the following covariates
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
+baseModelBIC = fitFixedEffectModel(adjust.covars, 
                                    CQN.GENE_EXPRESSION$E, 

### 2020-01-31 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
verbose md filter

@@
   dplyr::select(Individual_ID, `SNP_report:Genotyping_Sample_ID`, `SNP_report:Exclude?`) %>% 
-  dplyr::inner_join(ANCESTRY)
+  dplyr::inner_join(ANCESTRY) %>% 
+  %>% 
+  dplyr::filter(is.na(`SNP_report:Exclude?`)) #Remove excluded samples here because there are duplicate individuals
 ```
@@
 }
-
-# vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
-#                                      COVARIATES, CQN.GENE_EXPRESSION$E, 
-#                                      NEW.COUNTS, baseModelBIC)
 # Function to select covariates in a step-wise regression fashion

### 2020-02-05 ampad-rnaseq-de-clean/msbb_cohort.R
blocked by unknown vars in msbb md

@@ -1,0 +1,33 @@
+# Get Access to Data: https://adknowledgeportal.synapse.org/#/DataAccess/Instructions
+source("../processFunctions/functions.R")
+
+# Set config file
+Sys.setenv(R_CONFIG_ACTIVE = "mssm")
+
+# Login to Synapse. Make a Synapse account: https://docs.synapse.org/articles/getting_started.html
+synapser::synLogin()
+
+# Get and join metadata
+# RNA-seq metadata
+#
+assay_metadata <- get_data(config::get("assay metadata")$synID, config::get("assay metadata")$version)
+
+#Get clarity on why individual identifier differs
+foo <- assay_metadata %>%
+  dplyr::mutate(ext = ifelse(grepl("bam", fileName), "bam", NA),
+         ext = ifelse(grepl("fastq", fileName), "fastq", ext)) %>%
+  dplyr::filter(ext == "bam") %>%
+  dplyr::select(individualIdentifier, individualIdentifier.inferred)
+  #dplyr::select(sampleIdentifier, BrodmannArea, barcode, individualIdentifier, batch, RIN)
+
+sequencing_metadata <- get_data(config::get("sequencing metadata")$synID)
+# Join synIDs and select sample identifiers to map sequencing metadata
+# sequencing_metadata <- get_data(config::get("sequencing metadata")$synID,
+#                                 config::get("sequencing metadata")$version) %>%
+#   full_join(get_data(config::get("synID mapping")$synID), by = c("sample" = "id")) %>%
+#   dplyr::select(specimenID, everything(), -sample, -versionNumber) %>%
+#   distinct()
+
+# Import counts
+# Remove unmapped, multimapped, noFeature and ambiguous counts
+counts <- get_data(config::get("counts")$synID)[-c(1:4),]

### 2020-02-05 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd

@@
 # Fit contrast
+# Fit contrast
 contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
@@
 # Get differnetial expression
-DE = lapply(1:3, function(i, FIT){
+DE = lapply(1:6, function(i, FIT){
   topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
@@
                 Direction = as.character(Direction)) %>%
+  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
+  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
+  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
+  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
   left_join(Ensemble2HGNC) %>%
-  left_join(gene_lengths %>%
-              rownameToFirstColumn('gene_id'))
+  left_join(gene_lengths %>% rownameToFirstColumn('gene_id'))
 DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison) %>%
-  group_by(Comparison) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
+  group_by(Comparison, Tissue) %>%
   dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
@@
   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison) %>%
-  group_by(Comparison) %>%
+  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
+  group_by(Comparison, Tissue) %>%
   dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
@@
 p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
-p = p + facet_grid(.~Comparison, scales = 'fixed')
+p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
 p

### 2020-02-14 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
specify dplyr due to variancePartition conflict

--- 2020-01-27 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
 knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
-                                 parentId ="syn21435176",
+                                 parentId ="syn9872607",
                                  folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model")
@@
 # Github link
-thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='compute-tissue-cor')
+thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='master')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))
@@
 ```
+Following sample are removed `r paste(removedIDs, collapse = ',')`
 
@@
   tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
-  mutate(value = log(value)) %>%
+  dplyr::mutate(value = log(value)) %>%
   dplyr::rename(`counts(log)`= value) %>% 
@@
 
-``` {r cor.repeat.measures}
-repeated <- COVARIATES %>%
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>% 
-  dplyr::group_by(Individual_ID, Tissue) %>% 
-  dplyr::summarize(Samples = n()) %>% 
-  dplyr::filter(Samples == 1) %>% 
-  dplyr::group_by(Individual_ID) %>% 
-  dplyr::summarize(multiSample_perTissue = n()) %>% 
-  dplyr::filter(multiSample_perTissue == 2)
-
-glue::glue("{length(unique(repeated$Individual_ID))} of {length(unique(COVARIATES$Individual_ID))} have repeat measurements")
-
-quant_cor <- function(count_matrix, metadata, repeated_vec){
-  correlations <- tribble(~ iid, ~ cor)
-  for (i in repeated_vec){
-     samps <- metadata$SampleID[metadata$Individual_ID %in% i]
-     cors <- cor(count_matrix[,samps[1]], count_matrix[,samps[2]])
-     entry <- tribble(~ iid, ~ cor,
-                      as.character(i), cors)
-     correlations <- bind_rows(correlations,entry)
-  }
-  correlations
-}
-
-tmp = NEW.COUNTS
-tmp[is.na(tmp)] = 0
-
-correlation_tissue <- quant_cor(tmp, METADATA, repeated$Individual_ID)
-```
-
 ### Sanity check
@@
 ```{r diffExp, fig.height=10, fig.width=15}
+
 formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
@@
 # Visualize contrast matrix
-variancePartition::plotContrasts(L)
+plotContrasts(L)
 
@@
 }, FIT.CONTR) 
-```
-
-```{r testchunk}
 names(DE) <- contrast
@@
 # Store design matrix
-FIT.CONTR$design %>%
+DM1$design %>%
   rownameToFirstColumn('SampleID') %>%
@@
 synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
-
-# Store correlation between tissues
-write.table(correlation_tissue, file = 'CMC_ACC_DLPFC_correlationTissue.tsv', sep = '\t', row.names = FALSE, quote = FALSE)
-COR_OBJ = File('CMC_ACC_DLPC_correlationTissue.tsv',
-               name = 'Correlation between Tissues, Same Individual',
-               parentId = CODE$properties$id)
-COR_OBJ = synStore(COR_OBJ, used = ALL_USED_IDs, activityName = activityName,
-                   executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(COR_OBJ, annotations = all.annotations)
 ```

### 2020-02-19 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
add DREAM across all cohorts

--- 2020-01-06 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
 ---
-title: "HBCC - ACC - STG - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
+title: "HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
 author: "Kelsey Montgomery and Thanneer Perumal"
@@
                                           parentId ="syn21435176",
-                                          folderName = 'HBCC - ACC - STG - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
+                                          folderName = 'HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
 ```
@@
 
+library(variancePartition)
 library(synapser)
@@
                 TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
-  dplyr::select(SampleID, Individual_ID, Institution, Tissue, Reported_Gender, Sex, Ethnicity, ageOfDeath, PMI, Dx, 
+  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
                 RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
@@
   tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
-  mutate(value = log(value)) %>%
-  rename(`counts(log)`= value) %>% 
-  rename(SampleID = item) %>%
+  dplyr::mutate(value = log(value)) %>%
+  dplyr::rename(`counts(log)`= value) %>% 
+  dplyr::rename(SampleID = item) %>%
   left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
-  rename(`Reported Gender` = Reported_Gender)
+  dplyr::rename(`Reported Gender` = Reported_Gender)
 
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, "SampleID")) %>%
-    tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
   dplyr::mutate(label = SampleID) %>% 
@@
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Tissue, size=ageOfDeath))
-p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
+p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
+p <- p + theme_bw() %+replace% theme(legend.position="right")
 p <- p + geom_text(aes(label= label), size=4, hjust=0)
 p
-
 # Plot abberent distribution of logcpm counts
@@
   left_join(COVARIATES %>%
-              rownameToFirstColumn('SampleID') %>%
-              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))
-
-p = ggplot(tmp1 %>%
-             dplyr::filter(SampleID %in% outliers),
-           aes(x = logCPM, color = SampleID)) + geom_density() 
-p = p + theme(legend.position = 'top') + facet_grid(Dx+.~Tissue, scale = 'free')
-p
+              rownameToFirstColumn('SampleID'))
+# p = ggplot(tmp1 %>%
+#              dplyr::filter(SampleID %in% indToRemove),
+#            aes(x = logCPM, color = SampleID)) + geom_density() 
+# p = p + theme_bw() %+replace% theme(legend.position = 'none') + facet_grid(Dx~., scale = 'free')
+# p
 
@@
 p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA HBCC ACC-STG Cohort") + 
+p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA MSSM/Penn/Pitt Cohort") + 
   guides(colour = guide_legend(nrow = 2),
@@
 p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("MDS HBCC ACC-STG Cohort") + 
+p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("MDS MSSM/Penn/Pitt Cohort") + 
   guides(colour = guide_legend(nrow = 2),
@@
 
-# Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
-DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
-# Estimate voom weights
+# Post adjusted formula
+random_effect <- "Individual_ID"
+formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+
+# Estimate voom weights with DREAM
 tmp = NEW.COUNTS
 tmp[is.na(tmp)] = 0
-VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)
-
-# Estimate consensus correlation coefficient
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION = voom(tmp, 
-                            design=DM1$design, plot=F,
-                            block = COVARIATES$Individual_ID, 
-                            correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
-
+geneExpr = DGEList(tmp)
+geneExpr = edgeR::calcNormFactors(geneExpr)
+
+VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
+                                                               formula = formula,
+                                                               data = COVARIATES)
+
+# CQN NA counts to 0 counts for input to dream()
+tmp = CQN.GENE_EXPRESSION$E
+tmp[is.na(tmp)] = 0
+VOOM.GENE_EXPRESSION$E = tmp
+ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
+                                        formula = formula,
+                                        data = COVARIATES,
+                                        computeResiduals = TRUE)
 # Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
 
@@
 # Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
+varsToAddIn = grep("Dx.Tissue", colnames(ADJUSTED.FIT$design), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+   ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
 ```
@@
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Tissue, size=ageOfDeath))
-p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
+p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
+p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx.Tissue, scales = 'free_y')
 p
@@
 ```{r diffExp, fig.height=10, fig.width=15}
+formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueOther_ACC", "Dx.TissueControl_ACC"))
+L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueControl_ACC"))
+L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueOther_ACC"))
+L4 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueOther_STG", "Dx.TissueControl_STG"))
+L5 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueControl_STG"))
+L6 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueOther_STG"))
+L = cbind(L1, L2, L3, L4, L5, L6)
+contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
+              glue::glue_collapse(names(L2[L2 != 0]), "-"),
+              glue::glue_collapse(names(L3[L3 != 0]), "-"),
+              glue::glue_collapse(names(L4[L4 != 0]), "-"),
+              glue::glue_collapse(names(L5[L5 != 0]), "-"),
+              glue::glue_collapse(names(L6[L6 != 0]), "-"))
+# Visualize contrast matrix
+plotContrasts(L)
+
 # Fit contrast
-contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
-                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
-                                     "Dx.TissueOther_STG-Dx.TissueControl_STG",
-                                     "Dx.TissueSCZ_STG-Dx.TissueControl_STG",
-                                     "Dx.TissueSCZ_STG-Dx.TissueOther_STG"),
-                         levels = colnames(ADJUSTED.FIT$coefficients))
-FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
-FIT.CONTR = eBayes(FIT.CONTR)
-
-# Get differnetial expression
-DE = lapply(1:3, function(i, FIT){
-  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
+                                     formula = formula,
+                                     data = COVARIATES,
+                                     L = L)
+# Get differential expression
+DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
+  topTable(FIT, coef=i, number = Inf) %>%
     rownameToFirstColumn('gene_id') %>%
@@
 # Code
-CODE <- Folder(name = "HBCC - ACC - STG - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
 CODE <- synStore(CODE)
@@
 ```
+
 |  *Results*                                  |  *SynapseID*                     |

### 2020-02-19 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
add DREAM across all cohorts

@@
                                           parentId ="syn21435176",
-                                          folderName = 'HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
+                                          folderName = 'HBCC - DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
 ```
@@
 
-# Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
-DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
-# Estimate voom weights
+# Post adjusted formula
+random_effect <- "Individual_ID"
+formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+
+# Estimate voom weights with DREAM
 tmp = NEW.COUNTS
 tmp[is.na(tmp)] = 0
-VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)
-
-# Estimate consensus correlation coefficient
-correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION = voom(tmp, 
-                            design=DM1$design, plot=F,
-                            block = COVARIATES$Individual_ID, 
-                            correlation = correlation$cor)
+geneExpr = DGEList(tmp)
+geneExpr = edgeR::calcNormFactors(geneExpr)
+
+VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
+                                                               formula = formula,
+                                                               data = COVARIATES)
 
 # Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
+# CQN NA counts to 0 counts for input to dream()
+tmp = CQN.GENE_EXPRESSION$E
+tmp[is.na(tmp)] = 0
+VOOM.GENE_EXPRESSION$E = tmp
+ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
+                                        formula = formula,
+                                        data = COVARIATES,
+                                        computeResiduals = TRUE)
 
 # Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
 
@@
 # Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
+ varsToAddIn = grep("Dx.Tissue", colnames(ADJUSTED.FIT$design), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+   ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
 ```
@@
 ```{r diffExp, fig.height=10, fig.width=15}
-# Fit contrast
-contrast = makeContrasts(contrasts=c("DxOther-DxControl",
-                                     "DxSCZ-DxControl",
-                                     "DxSCZ-DxOther"),
-                         levels = colnames(ADJUSTED.FIT$coefficients))
-FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
-FIT.CONTR = eBayes(FIT.CONTR)
-
-# Get differnetial expression
-DE = lapply(1:3, function(i, FIT){
-  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+
+L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueOther_DLPFC", "Dx.TissueControl_DLPFC"))
+L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueControl_DLPFC"))
+L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueOther_DLPFC"))
+L = cbind(L1, L2, L3)
+contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
+              glue::glue_collapse(names(L2[L2 != 0]), "-"),
+              glue::glue_collapse(names(L3[L3 != 0]), "-"))
+
+# Visualize contrast matrix
+plotContrasts(L)
+
+#Fit contrast
+FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
+                                     formula = formula,
+                                     data = COVARIATES,
+                                     L = L)
+# Get differential expression
+DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
+  topTable(FIT, coef=i, number = Inf) %>%
     rownameToFirstColumn('gene_id') %>%
@@
 }, FIT.CONTR) 
-names(DE) = colnames(contrast)
-
+ names(DE) <- contrast
 DE = DE %>% 
@@
 # Code
-CODE <- Folder(name = "HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "HBCC - DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
 CODE <- synStore(CODE)

### 2020-02-19 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
add DREAM across all cohorts

@@
                                   paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')
-                              writeLines(paste("This is the formula:",formula))
+
                                 weights = voomWeights[i,]
@@
 ```{r phase1}
-# Fit a base model with the following covariates varsToFit, groupingVar, lcpm, cnt, covars
+# Fit a base model with the following covariates
 adjust.covars = c('Dx.Tissue')
@@
 
-# Residuals after normalisation - residuals.MArrayLM() does not take MArrayLM2 objects
+# Residuals after normalisation
 RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
@@
 ```
-```
+
 |  *Results*                                  |  *SynapseID*                     |

### 2020-02-19 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
cp

@@
                                           parentId ="syn21435176",
-                                          folderName = 'HBCC - ACC-Nirmala - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
+                                          folderName = 'HBCC - ACC-Nirmala - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
 ```
@@
   dplyr::inner_join(ANCESTRY) %>%
-  %>% 
   dplyr::filter(is.na(`SNP_report:Exclude?`)) #Remove excluded samples here because there are duplicate individuals
@@
 
-# Post adjusted design matrix
-DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
-DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
-
-# Estimate voom weights
+# Post adjusted formula
+random_effect <- "Individual_ID"
+formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+
+# Estimate voom weights with DREAM
 tmp = NEW.COUNTS
 tmp[is.na(tmp)] = 0
-VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)
-
-# Estimate consensus correlation coefficient
-correlation <- CovariateAnalysis::parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
-
-# Re-calculate voom weights with correlation of random effects
-VOOM.GENE_EXPRESSION = voom(tmp, 
-                            design=DM1$design, plot=F,
-                            block = COVARIATES$Individual_ID, 
-                            correlation = correlation$cor)
-
-# Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
+geneExpr = DGEList(tmp)
+geneExpr = edgeR::calcNormFactors(geneExpr)
+
+VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
+                                                               formula = formula,
+                                                               data = COVARIATES)
+# CQN NA counts to 0 counts for input to dream()
+tmp = CQN.GENE_EXPRESSION$E
+tmp[is.na(tmp)] = 0
+VOOM.GENE_EXPRESSION$E = tmp
+ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
+                                        formula = formula,
+                                        data = COVARIATES,
+                                        computeResiduals = TRUE)
 
 # Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+ RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
 
@@
 # Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
+varsToAddIn = grep("Dx", colnames(ADJUSTED.FIT$design), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
+    ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
 ```
@@
 ```{r diffExp, fig.height=10, fig.width=15}
-# Fit contrast
-contrast = makeContrasts(contrasts=c("DxOther-DxControl",
-                                     "DxSCZ-DxControl",
-                                     "DxSCZ-DxOther"),
-                         levels = colnames(ADJUSTED.FIT$coefficients))
-FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
-FIT.CONTR = eBayes(FIT.CONTR)
-
-# Get differnetial expression
-DE = lapply(1:3, function(i, FIT){
-  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
+formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueOther", "Dx.TissueControl"))
+L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ", "Dx.TissueControl"))
+L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ", "Dx.TissueOther"))
+L = cbind(L1, L2, L3)
+contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
+              glue::glue_collapse(names(L2[L2 != 0]), "-"),
+              glue::glue_collapse(names(L3[L3 != 0]), "-"))
+# Visualize contrast matrix
+plotContrasts(L) 
+#Fit contrast
+FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
+                                     formula = formula,
+                                     data = COVARIATES,
+                                     L = L)
+# Get differential expression
+DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
+  topTable(FIT, coef=i, number = Inf) %>%
     rownameToFirstColumn('gene_id') %>%
@@
 # Code
-CODE <- Folder(name = "HBCC - ACC-Nirmala - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "HBCC - ACC-Nirmala - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
 CODE <- synStore(CODE)
@@
 DEXP_OBJ = File('CMC_HBCC_ACC-Nirmala_DiffExpression.tsv', 
-                name = 'Differential Expression Results (Dx.Tissue)', 
+                name = 'Differential Expression Results (Dx)', 
                 parentId = CODE$properties$id)
@@
 ENRICH_OBJ = File('CMC_HBCC_ACC-Nirmala_EnrichResults.tsv', 
-                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
+                  name = 'Enrichment Analysis Results (Dx)', 
                   parentId = CODE$properties$id)

### 2020-02-19 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd

@@
                                                                formula = formula,
+                                                               data = COVARIATES,
+                                                               save.plot = TRUE)
+plot(VOOM.GENE_EXPRESSION)
+
+violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
+                                                         formula = formula,
                                                          data = COVARIATES)
+plotVarPart(sortCols(violin_plot))
+
 # CQN NA counts to 0 counts for input to dream()

### 2020-02-19 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
add voom mean-variance plot

@@
                                                                formula = formula,
+                                                               data = COVARIATES,
+                                                               save.plot = TRUE)
+plot(VOOM.GENE_EXPRESSION)
+violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
+                                                         formula = formula,
                                                          data = COVARIATES)
+plotVarPart(sortCols(violin_plot))
 
@@
                                         computeResiduals = TRUE)
+
 # Residuals after normalisation

### 2020-02-19 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
add voom mean-variance plot

@@
                                                                formula = formula,
+                                                               data = COVARIATES,
+                                                               save.plot = TRUE)
+plot(VOOM.GENE_EXPRESSION)
+violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
+                                                         formula = formula,
                                                          data = COVARIATES)
-
-# Fit linear model using new weights and new design
+plotVarPart(sortCols(violin_plot))
+
 # CQN NA counts to 0 counts for input to dream()

### 2020-02-19 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
add voom mean-variance plot

@@
                                         data = COVARIATES,
-                                        computeResiduals = TRUE)
+                                        computeResiduals = TRUE,
+                                        save.plot = TRUE)
+plot(VOOM.GENE_EXPRESSION)
+violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
+                                                         formula = formula,
+                                                         data = COVARIATES)
+plotVarPart(sortCols(violin_plot))
 

### 2020-02-19 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
store adjusted fit dream output obj

@@
 synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
+
+# Store adjusted fit linear mixed model object as RDS
+saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_dream-output.rds')
+ADJ_OBJ = File('CMC_ACC_DLPFC_dream-output.rds',
+               name = 'Dream Output object', 
+               parentId = CODE$properties$id)
+ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
+                   executed = thisFile, activityDescription = activityDescription)
+synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
 ```

### 2020-02-19 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
plot tissue correlation

@@
 residualSigCovars[["PC_res"]][[2]]$plotData
+```
+
+### Tissue Correlation
+```{r tissue_correlation}
+hasReps = COVARIATES %>%
+  dplyr::select(Individual_ID) %>% 
+  table
+
+df_dlpfc = COVARIATES %>% 
+  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
+  filter(grepl("DLPFC", Dx.Tissue)) %>% 
+  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
+df_acc = COVARIATES %>%
+  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
+  filter(grepl("ACC", Dx.Tissue)) %>% 
+  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
+
+df_merge = merge(df_dlpfc, df_acc, by="Individual_ID")
+
+corRes = sapply(res_std$ensembl_gene_id, function(ensGene){
+
+  x = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.x,drop=FALSE]
+  y = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.y,drop=FALSE]
+
+  cor(t(x)[,1], t(y)[,1], use="pairwise.complete.obs")
+})
+
+hist(corRes, xlab="Correlation", main="Correlation between ACC and DLPFC sample from the sample individual for each gene from ")
+
+pca = prcomp(t(VOOM.GENE_EXPRESSION$E))
+
+df_pca = merge(pca$x, as.data.frame(cov_dream), by.x="row.names", by.y="SampleID" )
+
+ggplot(df_pca, aes(PC1, PC2, color=Tissue, shape=Institution)) + geom_point() + theme_bw(17) + theme(aspect.ratio=1)
 ```

### 2020-02-19 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
plot tissue correlation

@@
 
+### Tissue Correlation
+```{r tissue_correlation}
+hasReps = COVARIATES %>%
+  dplyr::select(Individual_ID) %>% 
+  table
+
+df_dlpfc = COVARIATES %>% 
+  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
+  filter(grepl("DLPFC", Dx.Tissue)) %>% 
+  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
+df_acc = COVARIATES %>%
+  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
+  filter(grepl("ACC", Dx.Tissue)) %>% 
+  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
+
+df_merge = merge(df_dlpfc, df_acc, by="Individual_ID")
+
+corRes = sapply(res_std$ensembl_gene_id, function(ensGene){
+
+  x = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.x,drop=FALSE]
+  y = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.y,drop=FALSE]
+
+  cor(t(x)[,1], t(y)[,1], use="pairwise.complete.obs")
+})
+
+hist(corRes, xlab="Correlation", main="Correlation between ACC and DLPFC sample from the sample individual for each gene from ")
+
+pca = prcomp(t(VOOM.GENE_EXPRESSION$E))
+
+df_pca = merge(pca$x, as.data.frame(cov_dream), by.x="row.names", by.y="SampleID" )
+
+ggplot(df_pca, aes(PC1, PC2, color=Tissue, shape=Institution)) + geom_point() + theme_bw(17) + theme(aspect.ratio=1)
+```
+
 ### Residual calculation

### 2020-02-20 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove count winsorization step

@@
   dplyr::summarise(count = n())
+
+NEW.COUNTS = PROCESSED_COUNTS$filteredExprMatrix$counts
 ```
@@
 `r kable(tmp)`
-
-### Winsorise counts (Gene outliers in samples)
-```{r winsorise.data}
-LOG.CPM = cpm(PROCESSED_COUNTS$filteredExprMatrix$counts, log = T)
-
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
-LOG.CPM = apply(LOG.CPM, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-
-  ind.low = which(x == max(x[which(x <= mn-3*std.dev)], na.rm = T))
-  ind.high = which(x == min(x[which(x >= mn+3*std.dev)], na.rm = T))
-
-  x[x < (mn-3*std.dev)] = NA
-  x[x > (mn+3*std.dev)] = NA
-  return(x)
-}) %>% t
-
-# Back calculate counts for further analysis
-NEW.COUNTS = PROCESSED_COUNTS$filteredExprMatrix$counts
-NEW.COUNTS[is.na(LOG.CPM)] = 0
-```
 

### 2020-02-20 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove count winsorization step

@@
 Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue and Reported Gender. Also remove genes with missing gene length and percentage GC content
-
 ```{r cpmnormalisation}
@@
   spread(Tissue, count)
+
+NEW.COUNTS = PROCESSED_COUNTS$filteredExprMatrix$counts
 ```

### 2020-02-20 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
rm mean variance plot

@@
 # Estimate voom weights with DREAM
-tmp = NEW.COUNTS
-tmp[is.na(tmp)] = 0
-geneExpr = DGEList(tmp)
+geneExpr = DGEList(NEW.COUNTS)
 geneExpr = edgeR::calcNormFactors(geneExpr)
@@
                                                                formula = formula,
-                                                               data = COVARIATES)
+                                                               data = COVARIATES,
+                                                               save.plot = TRUE)
+#plot(VOOM.GENE_EXPRESSION)
 
 # Fit linear model using new weights and new design
-# CQN NA counts to 0 counts for input to dream()
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-VOOM.GENE_EXPRESSION$E = tmp
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
 ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
@@
                                         data = COVARIATES,
-                                        computeResiduals = TRUE,
-                                        save.plot = TRUE)
-plot(VOOM.GENE_EXPRESSION)
+                                        computeResiduals = TRUE)
+
 violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,

### 2020-02-20 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove winsorization step in cqn

@@
 CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-
-# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
-log.mat = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
-  mn = mean(x, na.rm = T)
-  std.dev = sd(x, na.rm = T)
-  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
-}) %>% t
-NEW.COUNTS[log.mat] = NA
-CQN.GENE_EXPRESSION$E[log.mat] = NA
 ```

### 2020-02-20 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
note correct branch

@@
 # Github link
-thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='master')
+thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='add-dream')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))

### 2020-02-28 ampad-rnaseq-de-clean/mayo_cohort.R
patch, subset dropped gene features

@@ -119,7 +119,7 @@ PMI <- metadata[,c("pmi", "sampleid", "donorid")]
 
 # Filter for variables to be used in downstream analysis as covariates
 covariates <- c("sampleid", "donorid", "source", "diagnosis", "tissue_diagnosis", "source_tissue_diagnosis",
-                "tissue_APOE4", "APOE4", "sex", "rin", "rin2", "age_death", "pmi",
+                "tissue_APOE4", "APOE4", "sex", "rin", "rin2", "age_death", "pmi", "flowcell",
                 "pct_pf_reads_aligned", "pct_coding_bases", "pct_intergenic_bases", "pct_intronic_bases",
                 "pct_ribosomal_bases")
 

### 2020-02-28 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd

--- 2020-02-19 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
 ---
-title: "HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
+title: "HBCC - ACC - STG - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
 author: "Kelsey Montgomery and Thanneer Perumal"
@@
                                           parentId ="syn21435176",
-                                          folderName = 'HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
+                                          folderName = 'HBCC - ACC - STG - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
 ```
@@
 
-library(variancePartition)
 library(synapser)
@@
 ALL_USED_IDs = METADATA_QC_ID
-metadata = downloadFile_version(METADATA_QC_ID, version = 15) %>%
+metadata = downloadFile_version(METADATA_QC_ID, version = 16) %>%
   dplyr::select(`Individual_ID`, Institution, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
-                Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
+                Sample_RNA_ID, one_of('Cohort', 'rnaSeq_isolation:RIN',
                                       'rnaSeq_dissection:Brain_Region',
@@
                                       'rnaSeq_report:Transcripts_Detected')) %>% 
-  dplyr::filter(Institution %in% c("NIMH-HBCC")) %>% 
+  dplyr::filter(Cohort %in% c("NIMH-HBCC-4")) %>% 
   dplyr::filter(`rnaSeq_dissection:Brain_Region` %in% c("ACC","STG"))
@@
   dplyr::select(Individual_ID, `SNP_report:Genotyping_Sample_ID`, `SNP_report:Exclude?`) %>% 
-  dplyr::inner_join(ANCESTRY)
+  dplyr::inner_join(ANCESTRY) %>% 
+  dplyr::filter(is.na(`SNP_report:Exclude?`)) #Remove excluded samples here because there are duplicate individuals
 ```
@@
 
-ind = METADATA$SampleID[which(METADATA$ReportExclude == 1 | METADATA$GenotypeExclude)]
+ind = METADATA$SampleID[which(METADATA$ReportExclude == 1)]
 writeLines(paste('Following',length(ind),'samples are marked exclude'))
@@
   group_by(Dx) %>% 
-  dplyr::summarise(count = n()) %>% 
+  summarise(count = n()) %>% 
   kable()
@@
   tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
-  dplyr::mutate(value = log(value)) %>%
-  dplyr::rename(`counts(log)`= value) %>% 
-  dplyr::rename(SampleID = item) %>%
+  mutate(value = log(value)) %>%
+  rename(`counts(log)`= value) %>% 
+  rename(SampleID = item) %>%
   left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
-  dplyr::rename(`Reported Gender` = Reported_Gender)
+  rename(`Reported Gender` = Reported_Gender)
 
@@
   group_by(gene_biotype) %>%
-  dplyr::summarise(fraction  = n()) %>%
+  summarise(fraction  = n()) %>%
   filter(fraction > 100) %>%
@@
   group_by(Dx, Tissue) %>%
-  dplyr::summarise(count = n()) %>%
+  summarise(count = n()) %>%
   spread(Tissue, count)
@@
 
-# Post adjusted formula
-random_effect <- "Individual_ID"
-formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-
-# Estimate voom weights with DREAM
+# Post adjusted design matrix
+DM1 = getDesignMatrix(COVARIATES[,adjust.covars,drop=F],Intercept = F)
+DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]
+
+# Estimate voom weights
 tmp = NEW.COUNTS
 tmp[is.na(tmp)] = 0
-geneExpr = DGEList(tmp)
-geneExpr = edgeR::calcNormFactors(geneExpr)
-
-VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
-                                                               formula = formula,
-                                                               data = COVARIATES,
-                                                               save.plot = TRUE)
-plot(VOOM.GENE_EXPRESSION)
-violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
-                                                         formula = formula,
-                                                         data = COVARIATES)
-plotVarPart(sortCols(violin_plot))
-
-# CQN NA counts to 0 counts for input to dream()
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-VOOM.GENE_EXPRESSION$E = tmp
-ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
-                                        formula = formula,
-                                        data = COVARIATES,
-                                        computeResiduals = TRUE)
+VOOM.GENE_EXPRESSION = voom(tmp, design=DM1$design, plot=F)
+
+# Estimate consensus correlation coefficient
+correlation <- parallelDuplicateCorrelation(VOOM.GENE_EXPRESSION, block = COVARIATES$Individual_ID)
+
+# Re-calculate voom weights with correlation of random effects
+VOOM.GENE_EXPRESSION = voom(tmp, 
+                            design=DM1$design, plot=F,
+                            block = COVARIATES$Individual_ID, 
+                            correlation = correlation$cor)
+
+# Fit linear model using new weights and new design
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)
 
 # Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
 
@@
 
-### Tissue Correlation
-```{r tissue_correlation}
-hasReps = COVARIATES %>%
-  dplyr::select(Individual_ID) %>% 
-  table
-
-df_dlpfc = COVARIATES %>% 
-  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
-  filter(grepl("DLPFC", Dx.Tissue)) %>% 
-  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
-df_acc = COVARIATES %>%
-  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
-  filter(grepl("ACC", Dx.Tissue)) %>% 
-  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
-
-df_merge = merge(df_dlpfc, df_acc, by="Individual_ID")
-
-corRes = sapply(res_std$ensembl_gene_id, function(ensGene){
-
-  x = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.x,drop=FALSE]
-  y = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.y,drop=FALSE]
-
-  cor(t(x)[,1], t(y)[,1], use="pairwise.complete.obs")
-})
-
-hist(corRes, xlab="Correlation", main="Correlation between ACC and DLPFC sample from the sample individual for each gene from ")
-
-pca = prcomp(t(VOOM.GENE_EXPRESSION$E))
-
-df_pca = merge(pca$x, as.data.frame(cov_dream), by.x="row.names", by.y="SampleID" )
-
-ggplot(df_pca, aes(PC1, PC2, color=Tissue, shape=Institution)) + geom_point() + theme_bw(17) + theme(aspect.ratio=1)
-```
-
 ### Residual calculation
@@
 # Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(ADJUSTED.FIT$design), value = T)
+varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-   ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
+  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
 ```
@@
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx.Tissue, scales = 'free_y')
+p <- p + geom_point(aes(shape=Tissue, size=ageOfDeath))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
 p
@@
 ```{r diffExp, fig.height=10, fig.width=15}
-formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueOther_ACC", "Dx.TissueControl_ACC"))
-L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueControl_ACC"))
-L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueOther_ACC"))
-L4 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueOther_STG", "Dx.TissueControl_STG"))
-L5 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueControl_STG"))
-L6 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueOther_STG"))
-L = cbind(L1, L2, L3, L4, L5, L6)
-contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
-              glue::glue_collapse(names(L2[L2 != 0]), "-"),
-              glue::glue_collapse(names(L3[L3 != 0]), "-"),
-              glue::glue_collapse(names(L4[L4 != 0]), "-"),
-              glue::glue_collapse(names(L5[L5 != 0]), "-"),
-              glue::glue_collapse(names(L6[L6 != 0]), "-"))
-# Visualize contrast matrix
-plotContrasts(L)
-
 # Fit contrast
-FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
-                                     formula = formula,
-                                     data = COVARIATES,
-                                     L = L)
-# Get differential expression
-DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
-  topTable(FIT, coef=i, number = Inf) %>%
+# Fit contrast
+contrast = makeContrasts(contrasts=c("Dx.TissueOther_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueControl_ACC",
+                                     "Dx.TissueSCZ_ACC-Dx.TissueOther_ACC",
+                                     "Dx.TissueOther_STG-Dx.TissueControl_STG",
+                                     "Dx.TissueSCZ_STG-Dx.TissueControl_STG",
+                                     "Dx.TissueSCZ_STG-Dx.TissueOther_STG"),
+                         levels = colnames(ADJUSTED.FIT$coefficients))
+FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
+FIT.CONTR = eBayes(FIT.CONTR)
+
+# Get differnetial expression
+DE = lapply(1:6, function(i, FIT){
+  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
     rownameToFirstColumn('gene_id') %>% 
@@
 }, FIT.CONTR) 
-names(DE) <- contrast
+names(DE) = colnames(contrast)
 
@@
 # Code
-CODE <- Folder(name = "HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "HBCC - ACC - STG - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
 CODE <- synStore(CODE)
@@
 synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
-
-# Store adjusted fit linear mixed model object as RDS
-saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_dream-output.rds')
-ADJ_OBJ = File('CMC_ACC_DLPFC_dream-output.rds',
-               name = 'Dream Output object', 
-               parentId = CODE$properties$id)
-ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
-                   executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
-```
-
+```
 |  *Results*                                  |  *SynapseID*                     |

### 2020-02-28 ampad-rnaseq-de-clean/mayo_cohort.R
patch, subset dropped gene features

@@ -126,6 +126,7 @@ covariates <- c("sampleid", "donorid", "source", "diagnosis", "tissue_diagnosis"
 metadata <- dplyr::select(metadata, covariates)
 
 # Remove samples from counts matrix
-counts <- counts[,colnames(counts) %in% metadata$sampleid]
+to_parse <- c("feature", colnames(counts)[colnames(counts) %in% metadata$sampleid])
+foo <- counts[,to_parse]
 
 # TODO: appropriately parse all the synIDs and version to add to provenance with config::get()

### 2020-02-28 ampad-rnaseq-de-clean/mayo_cohort.R
patch, subset dropped gene features

@@ -127,6 +127,6 @@ metadata <- dplyr::select(metadata, covariates)
 
 # Remove samples from counts matrix
 to_parse <- c("feature", colnames(counts)[colnames(counts) %in% metadata$sampleid])
-foo <- counts[,to_parse]
+counts <- counts[,to_parse]
 
 # TODO: appropriately parse all the synIDs and version to add to provenance with config::get()

### 2020-03-02 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove assignment of 0 counts, add plot dependency

@@
 ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-cr = cor(t(tmp))
+cr = cor(t(CQN.GENE_EXPRESSION$E))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-PC <- prcomp(tmp, scale.=T, center = T)
+PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)
 
@@
 
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-tree = hclust(as.dist(t(tmp)))
+tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 # Find correlation between PC's of gene expression with covariates
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-preAdjustedSigCovars = runPCAandPlotCorrelations(tmp, 
+preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
                                                  COVARIATES,
@@
                                                                save.plot = TRUE)
-#plot(VOOM.GENE_EXPRESSION)
+
+# How to plot a voom object if save.plot=TRUE
+# Set plot generic to voom result EList
+#' @export
+setMethod("plot", signature(x="EList", y="missing"),
+  function(x,  y, ...){
+    if( is.null(x$voom.xy) || is.null(x$voom.line)){
+        stop("x does not contain the trend information.\nvoom() must be run with save.plot=TRUE")
+    }
+    # points
+    df = data.frame(x = x$voom.xy$x,
+                    y = x$voom.xy$y)
+    # trend line
+    df.line = data.frame(x = x$voom.line$x,
+                        y = x$voom.line$y)
+    ggplot(df, aes(x,y)) + geom_point(size=.1) + theme_bw(15) + 
+      theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + 
+      geom_line(data=df.line, aes(x,y), color="red") + xlab(bquote(log[2](count + 0.5))) + 
+      ylab(expression( sqrt("standard deviation"))) + ggtitle("Voom: Mean-variance trend") + 
+      ylim(0, max(df$y))
+  }
+)
+
+plot(VOOM.GENE_EXPRESSION)
 
@@
 
+col_type <- dplyr::select(COVARIATES, c(adjust.covars, "Individual_ID")) %>%
+  dplyr::summarise_all(class) %>% 
+  tidyr::gather(variable, class)
+
+# Model categorical variables as random to report variance fractions
+formula_for_variance_fractions = sapply(1:length(col_type$class), function(i){
+  switch(col_type$class[i], 
+    "factor" = paste0('(1|', col_type$variable[i], ')'),
+    "numeric" = col_type$variable[i])
+  })
+
+formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(formula_for_variance_fractions, sep = " + "))
+
 violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
-                                                         formula = formula,
+                                                         formula = formula_for_variance_fractions,
                                                          data = COVARIATES)
@@
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-tmp = RESIDUAL.GENE_EXPRESSION
-tmp[is.na(tmp)] = 0
-residualSigCovars = runPCAandPlotCorrelations(tmp, COVARIATES,
+residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, COVARIATES,
                                               'adjusted design(voom-normalized)',

### 2020-03-02 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix tissue correlation plot, fix contrasts

@@
 ```{r tissue_correlation}
-hasReps = COVARIATES %>%
+hasReps = rownameToFirstColumn(COVARIATES, "SampleID") %>%
   dplyr::select(Individual_ID) %>% 
@@
 
-df_dlpfc = COVARIATES %>% 
+df_dlpfc = rownameToFirstColumn(COVARIATES, "SampleID") %>% 
   filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
@@
   dplyr::select(SampleID, Individual_ID, Dx.Tissue)
-df_acc = COVARIATES %>%
+df_acc = rownameToFirstColumn(COVARIATES, "SampleID") %>%
   filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
@@
 
+res_std <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id")
+
 corRes = sapply(res_std$ensembl_gene_id, function(ensGene){
@@
 
-df_pca = merge(pca$x, as.data.frame(cov_dream), by.x="row.names", by.y="SampleID" )
+df_pca = merge(pca$x, as.data.frame(rownameToFirstColumn(COVARIATES, "SampleID")), by.x="row.names", by.y="SampleID" ) %>% 
+  tidyr::separate(Dx.Tissue, into = c("Dx", "Tissue"), sep = "_")
 
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-tmp = RESIDUAL.GENE_EXPRESSION
-tmp[is.na(tmp)] = 0
-cr = cor(t(tmp))
+cr = cor(t(RESIDUAL.GENE_EXPRESSION))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Institution')])
-COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
 
@@
 ### Differential expression analysis
-Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=10, fig.width=15}
-
+Genes that are differentially expressed at an FDR <= 0.05 are:
+```{r set-formula}
 formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
   glue::glue(" + (1|{random_effect})")
-
+```
+`r formula`
+```{r diffExp, fig.height=10, fig.width=15}
 L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
@@
                                    coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueControl_DLPFC"))
-
 L6 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
@@
                                    coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueOther_DLPFC"))
-L = cbind(L1, L2, L3, L4, L5, L6)
+L7 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueSCZ_ACC"))
+L8 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueControl_DLPFC", "Dx.TissueControl_ACC"))
+L9 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.TissueOther_DLPFC", "Dx.TissueOther_ACC"))
+
+L = cbind(L1, L2, L3, L4, L5, L6, L7, L8, L9)
 
@@
               glue::glue_collapse(names(L5[L5 != 0]), "-"),
-              glue::glue_collapse(names(L6[L6 != 0]), "-"))
+              glue::glue_collapse(names(L6[L6 != 0]), "-"),
+              glue::glue_collapse(names(L7[L7 != 0]), "-"),
+              glue::glue_collapse(names(L8[L8 != 0]), "-"),
+              glue::glue_collapse(names(L9[L9 != 0]), "-")
+)
 
@@
 # Get differential expression
-DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
+DE = lapply(glue::glue("L{1:9}"), function(i, FIT){
   topTable(FIT, coef=i, number = Inf) %>%

### 2020-03-04 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
add contrasts for other conditions

@@
 title: "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
-author: "Kelsey Montgomery"
+author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
 date: "`r date()`"
@@
 
-### Normalisation
+### Normalisation - Dx.Tissue
 1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
@@
 VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
+Dx.Tissue_ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
                                         formula = formula,
@@
 # Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+Dx.Tissue_RESIDUAL.GENE_EXPRESSION = residuals(Dx.Tissue_ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
 
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, COVARIATES,
+residualSigCovars = runPCAandPlotCorrelations(Dx.Tissue_RESIDUAL.GENE_EXPRESSION, COVARIATES,
                                               'adjusted design(voom-normalized)',
@@
 
-res_std <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id")
+res_std <- rownameToFirstColumn(Dx.Tissue_RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id")
 
@@
 # Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(ADJUSTED.FIT$design), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
+varsToAddIn = grep("Dx.Tissue", colnames(Dx.Tissue_ADJUSTED.FIT$design), value = T)
+Dx.Tissue_RESIDUAL.GENE_EXPRESSION = Dx.Tissue_RESIDUAL.GENE_EXPRESSION +
+  Dx.Tissue_ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(Dx.Tissue_ADJUSTED.FIT$design[,varsToAddIn])
 ```
@@
 ```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(RESIDUAL.GENE_EXPRESSION))
+cr = cor(t(Dx.Tissue_RESIDUAL.GENE_EXPRESSION))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 
-### Differential expression analysis
+### Differential expression analysis (SCZ - Control - Other)
 Genes that are differentially expressed at an FDR <= 0.05 are:
@@
 ```
+
 `r formula`
+
 ```{r diffExp, fig.height=10, fig.width=15}
@@
                                    coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueOther_DLPFC"))
-L7 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueSCZ_ACC"))
-L8 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueControl_DLPFC", "Dx.TissueControl_ACC"))
-L9 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueOther_DLPFC", "Dx.TissueOther_ACC"))
-
-L = cbind(L1, L2, L3, L4, L5, L6, L7, L8, L9)
+
+L = cbind(L1, L2, L3, L4, L5, L6)
 
@@
               glue::glue_collapse(names(L5[L5 != 0]), "-"),
-              glue::glue_collapse(names(L6[L6 != 0]), "-"),
-              glue::glue_collapse(names(L7[L7 != 0]), "-"),
-              glue::glue_collapse(names(L8[L8 != 0]), "-"),
-              glue::glue_collapse(names(L9[L9 != 0]), "-")
+              glue::glue_collapse(names(L6[L6 != 0]), "-")
 )
@@
 # Get differential expression
-DE = lapply(glue::glue("L{1:9}"), function(i, FIT){
+DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
   topTable(FIT, coef=i, number = Inf) %>%
@@
 pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[1]] = pl[[1]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
 pl[[1]] = pl[[1]] + theme(legend.position = 'top')
@@
 pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gene_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[2]] = pl[[2]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
 pl[[2]] = pl[[2]] + theme(legend.position = 'top')
@@
 pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[3]] = pl[[3]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
 pl[[3]] = pl[[3]] + theme(legend.position = 'top')
@@
 multiplot(plotlist = pl, cols = 3)
+```
+
+### Normalisation - Dx.Tissue.Gender
+1. Dx.Tissue.Gender is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis, tissue and reported gender)
+2. Individual_ID is chosen as random effect
+```{r iterative.normalisation, results='asis'}
+COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Reported_Gender, sep = "_"))
+COVARIATES$Reported_Gender <- NULL
+COVARIATES$Dx.Tissue <- NULL
+adjust.covars[1] <- "Dx.Tissue.Gender"
+
+writeLines(paste('Using following covariates in the model:',
+                 paste(adjust.covars, collapse=', '),
+                 'as fixed effects and Individual_ID is chosen as random effect'))
+
+# Post adjusted formula
+random_effect <- "Individual_ID"
+formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+
+# Estimate voom weights with DREAM
+geneExpr = DGEList(NEW.COUNTS)
+geneExpr = edgeR::calcNormFactors(geneExpr)
+
+VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
+                                                               formula = formula,
+                                                               data = COVARIATES,
+                                                               save.plot = TRUE)
+
+plot(VOOM.GENE_EXPRESSION)
+
+# Fit linear model using new weights and new design
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+Dx.Tissue.Gender_ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
+                                        formula = formula,
+                                        data = COVARIATES,
+                                        computeResiduals = TRUE)
+
+col_type <- dplyr::select(COVARIATES, c(adjust.covars, "Individual_ID")) %>%
+  dplyr::summarise_all(class) %>% 
+  tidyr::gather(variable, class)
+
+# Model categorical variables as random to report variance fractions
+formula_for_variance_fractions = sapply(1:length(col_type$class), function(i){
+  switch(col_type$class[i], 
+    "factor" = paste0('(1|', col_type$variable[i], ')'),
+    "numeric" = col_type$variable[i])
+  })
+
+formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(formula_for_variance_fractions, sep = " + "))
+
+violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
+                                                         formula = formula_for_variance_fractions,
+                                                         data = COVARIATES)
+plotVarPart(sortCols(violin_plot))
+
+# Residuals after normalisation
+Dx.Tissue.Gender_RESIDUAL.GENE_EXPRESSION = residuals(Dx.Tissue.Gender_ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+
+# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+residualSigCovars = runPCAandPlotCorrelations(Dx.Tissue.Gender_RESIDUAL.GENE_EXPRESSION, COVARIATES,
+                                              'adjusted design(voom-normalized)',
+                                              isKeyPlot=TRUE)
+```
+
+### Sanity check
+```{r residualSigCovars.manual, fig.width=12, fig.height=8}
+residualSigCovars[["PC_res"]][[2]]$plotData
+```
+
+### Differential expression analysis (Gender-specific SCZ - Control - Other)
+Genes that are differentially expressed at an FDR <= 0.05 are:
+```{r set-formula}
+formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+```
+
+`r formula`
+
+```{r diffExp, fig.height=10, fig.width=15}
+L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.Tissue.GenderSCZ_ACC_Female", "Dx.Tissue.GenderControl_ACC_Female"))
+L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.Tissue.GenderSCZ_ACC_Male", "Dx.Tissue.GenderControl_ACC_Male"))
+L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.Tissue.GenderSCZ_DLPFC_Female", "Dx.Tissue.GenderControl_DLPFC_Female"))
+L4 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Dx.Tissue.GenderSCZ_DLPFC_Male", "Dx.Tissue.GenderControl_DLPFC_Male"))
+
+L = cbind(L1, L2, L3, L4)
+
+contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
+              glue::glue_collapse(names(L2[L2 != 0]), "-"),
+              glue::glue_collapse(names(L3[L3 != 0]), "-"),
+              glue::glue_collapse(names(L4[L4 != 0]), "-")
+)
+
+# Visualize contrast matrix
+plotContrasts(L)
+
+# Fit contrast
+FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
+                                     formula = formula,
+                                     data = COVARIATES,
+                                     L = L)
+
+# Get differential expression
+DE_2 = lapply(glue::glue("L{1:4}"), function(i, FIT){
+  topTable(FIT, coef=i, number = Inf) %>%
+    rownameToFirstColumn('gene_id') %>% 
+    left_join(backgroundGenes)
+}, FIT.CONTR) 
+names(DE_2) <- contrast
+
+DE_2 = DE_2%>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('Dx.Tissue.Gender','',Comparison),
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>% 
+  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
+  tidyr::separate(to.state, into = c('to.Dx', 'to.Tissue', 'to.Gender'), sep = '_') %>%
+  tidyr::separate(ref.state, into = c('ref.Dx', 'ref.Tissue', 'ref.Gender'), sep = '_') %>% 
+  left_join(Ensemble2HGNC) %>%
+  left_join(gene_lengths %>% rownameToFirstColumn('gene_id'))
+DE_2$Direction[DE_2$adj.P.Val > 0.05 | abs(DE_2$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE_2 %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
+  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
+  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
+
+tmp1 = DE_2 %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
+  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
+
+kable(full_join(tmp,tmp1))
+
+p = ggplot(DE_2, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
+p = p + facet_grid(ref.Tissue~.+ref.Gender, scales = 'fixed')
+p = p + ggtitle("Differential Expression between Schizophrenia and Control")
+p
+```
+
+### Normalisation - Tissue
+1. Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and reported gender)
+2. Individual_ID is chosen as random effect
+```{r iterative.normalisation, results='asis'}
+###we might not need this adjustment
+COVARIATES <- tidyr::separate(COVARIATES, Dx.Tissue.Gender, c("Dx", "Tissue", "Reported_Gender")) %>% 
+  dplyr::mutate(Dx = factor(Dx, levels = c("SCZ", "Control", "Other")),
+                Tissue = factor(Tissue), 
+                Reported_Gender = factor(Reported_Gender))
+
+adjust.covars[1] <- "Dx"
+
+adjust.covars <- c(adjust.covars, "Tissue", "Reported_Gender")
+
+writeLines(paste('Using following covariates in the model:',
+                 paste(adjust.covars, collapse=', '),
+                 'as fixed effects and Individual_ID is chosen as random effect'))
+
+# Post adjusted formula
+random_effect <- "Individual_ID"
+formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+
+# Estimate voom weights with DREAM
+geneExpr = DGEList(NEW.COUNTS)
+geneExpr = edgeR::calcNormFactors(geneExpr)
+
+VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
+                                                               formula = formula,
+                                                               data = COVARIATES,
+                                                               save.plot = TRUE)
+
+plot(VOOM.GENE_EXPRESSION)
+
+# Fit linear model using new weights and new design
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+Tissue_ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
+                                        formula = formula,
+                                        data = COVARIATES,
+                                        computeResiduals = TRUE)
+
+col_type <- dplyr::select(COVARIATES, c(adjust.covars, "Individual_ID")) %>%
+  dplyr::summarise_all(class) %>% 
+  tidyr::gather(variable, class)
+
+# Model categorical variables as random to report variance fractions
+formula_for_variance_fractions = sapply(1:length(col_type$class), function(i){
+  switch(col_type$class[i], 
+    "factor" = paste0('(1|', col_type$variable[i], ')'),
+    "numeric" = col_type$variable[i])
+  })
+
+formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(formula_for_variance_fractions, sep = " + "))
+
+violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
+                                                         formula = formula_for_variance_fractions,
+                                                         data = COVARIATES)
+plotVarPart(sortCols(violin_plot))
+
+# Residuals after normalisation
+Tissue_RESIDUAL.GENE_EXPRESSION = residuals(Tissue_ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+
+# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
+residualSigCovars = runPCAandPlotCorrelations(Tissue_RESIDUAL.GENE_EXPRESSION, COVARIATES,
+                                              'adjusted design(voom-normalized)',
+                                              isKeyPlot=TRUE)
+```
+
+### Sanity check
+```{r residualSigCovars.manual, fig.width=12, fig.height=8}
+residualSigCovars[["PC_res"]][[2]]$plotData
+```
+
+### Differential expression analysis (ACC - DLPFC)
+Genes that are differentially expressed at an FDR <= 0.05 are:
+```{r set-formula}
+formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+```
+
+`r formula`
+
+```{r diffExp, fig.height=10, fig.width=15}
+L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("TissueDLPFC"))
+L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("DxSCZ", "DxControl"))
+L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Reported_GenderMale"))
+
+L = cbind(L1, L2, L3)
+
+contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
+              glue::glue_collapse(names(L2[L2 != 0]), "-"),
+              glue::glue_collapse(names(L3[L3 != 0]), "-")
+)
+
+# Visualize contrast matrix
+plotContrasts(L)
+
+# Fit contrast
+FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
+                                     formula = formula,
+                                     data = COVARIATES,
+                                     L = L)
+
+# Get differential expression
+DE_3 = lapply(glue::glue("L{1:3}"), function(i, FIT){
+  topTable(FIT, coef=i, number = Inf) %>%
+    rownameToFirstColumn('gene_id') %>% 
+    left_join(backgroundGenes)
+}, FIT.CONTR) 
+names(DE_3) <- contrast
+
+DE_3 = DE_3%>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>% 
+  dplyr::mutate(Comparison = ifelse(Comparison %in% "Reported_GenderMale", "Female-Male", Comparison),
+                Comparison = ifelse(Comparison %in% "DxSCZ-DxControl", "SCZ-Control", Comparison),
+                Comparison = ifelse(Comparison %in% "TissueDLPFC", "ACC-DLPFC", Comparison)) %>% 
+  left_join(Ensemble2HGNC) %>%
+  left_join(gene_lengths %>% rownameToFirstColumn('gene_id'))
+DE_3$Direction[DE_3$adj.P.Val > 0.05 | abs(DE_3$logFC) < log2(1.2)] = 'NONE'
+
+tmp = DE_3 %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(ensembl_gene_id, Comparison) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
+
+tmp1 = DE_3 %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+  dplyr::select(ensembl_gene_id, Comparison) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
+
+kable(full_join(tmp,tmp1))
+
+p = ggplot(DE_3, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
+p = p + facet_grid(Comparison~., scales = 'fixed')
+p
 ```
@@
 
-# Store residual gene expression for network analysis
-RESIDUAL.GENE_EXPRESSION %>%
+# Store residual gene expression for network analysis - Dx.Tissue
+Dx.Tissue_RESIDUAL.GENE_EXPRESSION %>%
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
-                name = 'Normalised, covariates removed residual expression (for network analysis)', 
+  write.table(file = 'CMC_ACC_DLPFC_Dx.Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+nEXP_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue_netResidualExpression.tsv', 
+                name = 'Dx.Tissue Normalised, covariates removed residual expression (for network analysis)', 
                 parentId = CODE$properties$id)
@@
 
-# Store differential expression results
-write.table(DE, file = 'CMC_ACC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_ACC_DLPFC_DiffExpression.tsv', 
+# Store residual gene expression for network analysis - Dx.Tissue.Gender
+Dx.Tissue.Gender_RESIDUAL.GENE_EXPRESSION %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'CMC_ACC_DLPFC_Dx.Tissue.Gender_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+nEXP_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue.Gender_netResidualExpression.tsv', 
+                name = 'Dx.Tissue.Gender Normalised, covariates removed residual expression (for network analysis)', 
+                parentId = CODE$properties$id)
+nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                    executed = thisFile, activityDescription = activityDescription)
+synSetAnnotations(nEXP_OBJ, annotations = all.annotations)
+
+# Store residual gene expression for network analysis - Tissue
+Tissue_RESIDUAL.GENE_EXPRESSION %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'CMC_ACC_DLPFC_Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+nEXP_OBJ = File('CMC_ACC_DLPFC_Tissue_netResidualExpression.tsv', 
+                name = 'Tissue Normalised, covariates removed residual expression (for network analysis)', 
+                parentId = CODE$properties$id)
+nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                    executed = thisFile, activityDescription = activityDescription)
+synSetAnnotations(nEXP_OBJ, annotations = all.annotations)
+
+# Store differential expression results - Dx.Tissue
+write.table(DE, file = 'CMC_ACC_DLPFC_Dx.Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue_DiffExpression.tsv', 
                 name = 'Differential Expression Results (Dx.Tissue)', 
+                parentId = CODE$properties$id)
+DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                    executed = thisFile, activityDescription = activityDescription)
+synSetAnnotations(DEXP_OBJ, annotations = all.annotations)
+
+# Store differential expression results - Dx.Tissue.Gender
+write.table(DE_2, file = 'CMC_ACC_DLPFC_Dx.Tissue.Gender_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue.Gender_DiffExpression.tsv', 
+                name = 'Differential Expression Results (Dx.Tissue.Gender)', 
+                parentId = CODE$properties$id)
+DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                    executed = thisFile, activityDescription = activityDescription)
+synSetAnnotations(DEXP_OBJ, annotations = all.annotations)
+
+# Store differential expression results - Tissue
+write.table(DE_3, file = 'CMC_ACC_DLPFC_Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('CMC_ACC_DLPFC_Tissue_DiffExpression.tsv', 
+                name = 'Differential Expression Results (Tissue)', 
                 parentId = CODE$properties$id)
@@
 
-# Store adjusted fit linear mixed model object as RDS
-saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_dream-output.rds')
-ADJ_OBJ = File('CMC_ACC_DLPFC_dream-output.rds',
-               name = 'Dream Output object', 
+# Store adjusted fit linear mixed model object as RDS - Dx.Tissue
+saveRDS(Dx.Tissue_ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_Dx.Tissue_dream-output.rds')
+ADJ_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue_dream-output.rds',
+               name = 'Dream Output object - Dx.Tissue', 
+               parentId = CODE$properties$id)
+ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
+                   executed = thisFile, activityDescription = activityDescription)
+synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
+
+# Store adjusted fit linear mixed model object as RDS - Dx.Tissue.Gender
+saveRDS(Dx.Tissue.Gender_ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_Dx.Tissue.Gender_dream-output.rds')
+ADJ_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue.Gender_dream-output.rds',
+               name = 'Dream Output object - Dx.Tissue.Gender', 
+               parentId = CODE$properties$id)
+
+ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
+                   executed = thisFile, activityDescription = activityDescription)
+synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
+
+# Store adjusted fit linear mixed model object as RDS - Tissue
+saveRDS(Dx.Tissue.Gender_ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_Tissue_dream-output.rds')
+ADJ_OBJ = File('CMC_ACC_DLPFC_Tissue_dream-output.rds',
+               name = 'Dream Output object - Tissue', 
                parentId = CODE$properties$id)

### 2020-03-04 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove enrichment analysis

@@
 
-### Gene set enrichment analysis
-Using Fisher's exact test for gene set enrichment analysis
-```{r enrich.de, fig.height=16, fig.width=16}
-diffexp.genes = DE %>%
-  dlply(.(Comparison, Tissue), .fun = function(x){
-    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
-    tmp = tmp[!is.na(tmp)]
-  })
-diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
-backgroundGenes = unique(DE$hgnc_symbol)
-
-# Download all related genesets from synapse
-geneset.files = synGetChildren("syn3240583", includeTypes=list("file"))$asList()
-genesets.to.test = purrr::map(geneset.files, .f = function(fileList){
-  tmp = downloadFile(fileList$id) %>%
-    dlply(.(Name), .fun = function(x){
-      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
-    })
-  }) %>%
-  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
-  do.call(c,.)
-
-# Perform enrichment analysis
-enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
-   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
-     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
-    }, x, backgroundGenes, .parallel = TRUE) %>%
-    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
-}, genesets.to.test, backgroundGenes) %>%
-  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
-  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
-  dplyr::rename(Comparison = Name, Category = .id)
-```
-
 ### Store files in synapse
@@
 
-# Store enrichment results
-write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
-ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
-                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
-                  parentId = CODE$properties$id)
-ENRICH_OBJ = synStore(ENRICH_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                      executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
-
 # Store adjusted fit linear mixed model object as RDS - Dx.Tissue

### 2020-03-04 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove design matrix

@@
 
-# Store design matrix
-DM1$design %>%
-  rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'CMC_ACC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('CMC_ACC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
-DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                  executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(DM_OBJ, annotations = all.annotations)
-
 # Store residual gene expression for network analysis - Dx.Tissue

### 2020-03-04 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix typo

@@
                                  parentId ="syn21435176",
-                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model")
+                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN")
 ```
@@
 # Code
-CODE <- Folder(name = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
 CODE <- synStore(CODE)

### 2020-03-04 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix duplicate code chunk names

@@
 2. Individual_ID is chosen as random effect
-```{r iterative.normalisation, results='asis'}
+```{r iterative.normalisation-gender, results='asis'}
 COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Reported_Gender, sep = "_"))
@@
 ### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
+```{r residualSigCovars.manual-gender, fig.width=12, fig.height=8}
 residualSigCovars[["PC_res"]][[2]]$plotData
@@
 Genes that are differentially expressed at an FDR <= 0.05 are:
-```{r set-formula}
+```{r set-formula-gender}
 formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
@@
 
-```{r diffExp, fig.height=10, fig.width=15}
+```{r diffExp-gender, fig.height=10, fig.width=15}
 L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
@@
 2. Individual_ID is chosen as random effect
-```{r iterative.normalisation, results='asis'}
+```{r iterative.normalisation-tissue, results='asis'}
 ###we might not need this adjustment
@@
 ### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
+```{r residualSigCovars.manual-tissue, fig.width=12, fig.height=8}
 residualSigCovars[["PC_res"]][[2]]$plotData
@@
 Genes that are differentially expressed at an FDR <= 0.05 are:
-```{r set-formula}
+```{r set-formula-tissue}
 formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
@@
 
-```{r diffExp, fig.height=10, fig.width=15}
+```{r diffExp-tissue, fig.height=10, fig.width=15}
 L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 

### 2020-03-05 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
emphasize where vars are added

@@
 ```
+#### Variables included in the model 
+Covariates were added as fixed effects iteratively in three phases if model improvement by BIC criteria was observed in the majority of genes.
+Clinical, ancestry and sample-specific technical variables were tested for model improvement first.
+Covariates related to batch effects were next tested for model improvement and finally, RNASeq alignment-specific covariates.
+
 Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`
@@
                                                          data = COVARIATES)
+
 plotVarPart(sortCols(violin_plot))

### 2020-03-06 debug.Rmd
debug voom serialize error

---
title: "test-problems"
output: html_document
---

```{r knit2synapse, eval=FALSE}
library(synapser)
library(knit2synapse)

synLogin()

knit2synapse::createAndKnitToFolderEntity(file = "debug.Rmd",
                                 parentId ="syn21725610",
                                 folderName = "debugME")
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(tibble)
library(plyr)
library(dplyr)
library(stringr)
library(R.utils)
library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(lme4)
library(biomaRt)
#devtools::install_github("GabrielHoffman/variancePartition@613dfe9")
library(variancePartition)
library(synapser)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')
library(BiocParallel)
param = SnowParam(4, "SOCK", progressbar=TRUE)

synLogin()

# library(doParallel)
library(foreach)

# cl = makeCluster(detectCores(TRUE)-2)
# 
# registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r library}
# NEW.COUNTS <-

NEW.COUNTS <- read.table(synGet("syn21725766")$path, sep = "\t", header = TRUE)
rownames(NEW.COUNTS) <- NEW.COUNTS$genes
NEW.COUNTS$genes <- NULL
COVARIATES <- read.table(synGet("syn21725830")$path, sep = "\t", header = TRUE)
rownames(COVARIATES) <- COVARIATES$sampleid
COVARIATES$sampleid <- NULL
CQN.GENE_EXPRESSION$E <- read.table(synGet("syn21725967")$path, sep = "\t", header = TRUE)
rownames(CQN.GENE_EXPRESSION$E) <- CQN.GENE_EXPRESSION$E$genes
CQN.GENE_EXPRESSION$E$genes <- NULL
```

```{r iterative.normalisation}
adjust.covars <- c("Dx.Tissue", "RIN", "IntronicRate")

writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))

# Post adjusted formula
random_effect <- "Individual_ID"
formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
  glue::glue(" + (1|{random_effect})")

# Estimate voom weights with DREAM
geneExpr = DGEList(NEW.COUNTS)
geneExpr = edgeR::calcNormFactors(geneExpr)
message("pass")
VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
                                                               formula = formula,
                                                               data = COVARIATES,
                                                               save.plot = TRUE)
message("pass")
# How to plot a voom object if save.plot=TRUE
# Set plot generic to voom result EList
#' @export
setMethod("plot", signature(x="EList", y="missing"),
  function(x,  y, ...){
    if( is.null(x$voom.xy) || is.null(x$voom.line)){
        stop("x does not contain the trend information.\nvoom() must be run with save.plot=TRUE")
    }
    # points
    df = data.frame(x = x$voom.xy$x,
                    y = x$voom.xy$y)
    # trend line
    df.line = data.frame(x = x$voom.line$x,
                        y = x$voom.line$y)
    ggplot(df, aes(x,y)) + geom_point(size=.1) + theme_bw(15) + 
      theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + 
      geom_line(data=df.line, aes(x,y), color="red") + xlab(bquote(log[2](count + 0.5))) + 
      ylab(expression( sqrt("standard deviation"))) + ggtitle("Voom: Mean-variance trend") + 
      ylim(0, max(df$y))
  }
)
message("pass")  
plot(VOOM.GENE_EXPRESSION)

# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
Dx.Tissue_ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
                                        formula = formula,
                                        data = COVARIATES,
                                        computeResiduals = TRUE)
message("pass")
col_type <- dplyr::select(COVARIATES, c(adjust.covars, "Individual_ID")) %>%
  dplyr::summarise_all(class) %>% 
  tidyr::gather(variable, class)

# Model categorical variables as random to report variance fractions
formula_for_variance_fractions = sapply(1:length(col_type$class), function(i){
  switch(col_type$class[i], 
    "factor" = paste0('(1|', col_type$variable[i], ')'),
    "numeric" = col_type$variable[i])
  })

formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(formula_for_variance_fractions, sep = " + "))

violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
                                                         formula = formula_for_variance_fractions,
                                                         data = COVARIATES)
message("pass")
plotVarPart(sortCols(violin_plot))

# Residuals after normalisation
Dx.Tissue_RESIDUAL.GENE_EXPRESSION = residuals(Dx.Tissue_ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(Dx.Tissue_RESIDUAL.GENE_EXPRESSION, COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

```{r store}
CODE <- Folder(name = "debugME", parentId = "syn21725610")
CODE <- synStore(CODE)
```

### 2020-03-25 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix input

@@
 library(biomaRt)
-#devtools::install_github("GabrielHoffman/variancePartition@613dfe9")
 library(variancePartition)
@@
 # Download reprocessed counts
-counts = 'syn21071622'
+counts = 'syn21754352'
 ALL_USED_IDs = counts
@@
 metadata = downloadFile_version(metadata, version = 14)  %>% 
-  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx,
+  dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, 
+                Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx,
                 Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',

### 2020-03-25 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
add future pkg
add ENSEMBL mirror

@@
 library(foreach)
-
-cl = makeCluster(detectCores()-2)
+library(future)
+
+cl = makeCluster(availableCores()-2)
 
@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "sep2019.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
+                host = "sep2019.archive.ensembl.org", 
+                dataset = "hsapiens_gene_ensembl",
+                mirror = "uswest")
 

### 2020-03-25 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
archived versions of ensembl down, using current versions

@@
 mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
-                host = "sep2019.archive.ensembl.org", 
-                dataset = "hsapiens_gene_ensembl",
-                mirror = "uswest")
+                host = "uswest.ensembl.org", 
+                dataset = "hsapiens_gene_ensembl")
 

### 2020-03-25 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
archived versions of ensembl down, using current versions

@@
 mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
-                host = "uswest.ensembl.org", 
+                host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)
                 dataset = "hsapiens_gene_ensembl")

### 2020-03-26 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove comp. heavy model selection step

@@
 
-### Model Identification
-```{r fxn}
-# Function to fit a mixed effect model and return BIC
-fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){
-
-  # Get design matrix
-  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
-  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]
-
-  # Estimate voom weights
-  cnt[is.na(cnt)] = 0
-  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights
-
-  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
-                              .combine = rbind,
-                              .packages = c('lme4', 'dplyr'),
-                              .export = c('voomWeights')) %dopar% {
-
-                                # Perform weighted mixed linear models for every gene
-                                expr = data.frame(Gx = lcpm[i,])
-                                data = cbind(expr,
-                                             designMat[rownames(expr),],
-                                             Individual_ID = groupingVar)
-                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
-                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')
-
-                                weights = voomWeights[i,]
-                                BIC = NA
-                                tryCatch({
-                                  mdl = lmer(formula = formula, 
-                                             data = data, 
-                                             weights = weights)
-                                  BIC = BIC(mdl)
-                                }, error = function(e){
-                                  BIC = NA
-                                })
-                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
-                                                BIC = BIC))
-                            }
-  return(modelBIC)
-  }
-
-# Function to select covariates in a step-wise regression fashion
-selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){
-
-  varsSelected = c()
-  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
-           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
-  while(length(covarsToSelect) != 0){
-    allModelBIC = foreach(i = 1:length(covarsToSelect),
-                          .combine = c,
-                          .packages = pkgs,
-                          .export = c('fitMixedEffectModel')) %dopar% {
-
-                            # Fit a mixed effect linear model and get all BIC
-                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
-                                                           covars[colnames(cnt),'Individual_ID'], 
-                                                           lcpm, 
-                                                           cnt, 
-                                                           covars) %>%
-                            plyr::rename(c('BIC' = covarsToSelect[i]))
-
-                            return(list(modelBIC))
-                          }
-    allModelBIC = plyr::join_all(allModelBIC)
-
-    # Find number of genes that are changed based on BIC
-    nChange = allModelBIC %>%
-      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
-      dplyr::left_join(baseModelBIC) %>%
-      dplyr::mutate(diff = baseModel - Value) %>%
-      dplyr::group_by(ModelName) %>%
-      dplyr::summarise(better = sum(diff >= 5), 
-                       worse = sum(diff <= -5)) %>%
-      dplyr::mutate(nchange = better - worse)
-
-    print(nChange)
-
-    # Select variable
-    tmp = nChange %>%
-      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
-      dplyr::top_n(1, nchange)
-
-    if (dim(tmp)[1] == 0){
-      covarsToSelect = c()
-    } else {
-      baseModelBIC = allModelBIC %>%
-        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
-      colnames(baseModelBIC)[2] = 'baseModel'
-      varsSelected = c(varsSelected, tmp$ModelName)
-      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
-      print(paste('Variable chosen to be added:',tmp$ModelName))
-    }
-  }
-
-  return(varsSelected)
-}
-```
-#### Phase I (clinical, ancestry and sample specific technical variables)
-```{r phase1}
-# Fit a base model with the following covariates
-adjust.covars = c('Dx.Tissue')
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-
-# Iteratively identify covariates in phase I
-phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'ageOfDeath', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
-vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-
-adjust.covars = c(adjust.covars, vars.selected)
-```
-Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`
-
-#### Phase II (batch effects)
-```{r phase2}
-# Fit a base model with the following covariates
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-
-# Iteratively identify covariates in phase I
-phase2.covars = c('LibraryBatch', 'FlowcellBatch')
-vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-
-adjust.covars = c(adjust.covars, vars.selected)
-```
-Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`
-
-#### Phase III (RNASeq alignment specific covariates)
-```{r phase3}
-# Fit a base model with the following covariates
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-
-# Iteratively identify covariates in phase I
-phase3.covars = c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected", "rRNARate", "TotalReads")
-vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-
-adjust.covars = c(adjust.covars, vars.selected)
-```
+#### Model Selection
+```{r set_vars}
+adjust.covars <- c("Dx.Tissue", "RIN", "IntronicRate")
+```
+
+[Code](https://github.com/kelshmo/analysis/blob/f7345adde2052afbc258773f69930df300989bf3/cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd#L579-L737) for model selection and [BIC Criteria](https://www.synapse.org/#!Synapse:syn21543160)
+
 #### Variables included in the model 

### 2020-03-27 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
mitigate oom with less cores
parse DE to new markdown file

@@
 ---
-title: "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
+title: "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Model Selection"
 author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
@@
 
-cl = makeCluster(availableCores()-2)
+cl = makeCluster((availableCores()-2)/2)
 
@@
 
-### Normalisation - Dx.Tissue
-1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
-2. Individual_ID is chosen as random effect
-```{r iterative.normalisation, results='asis'}
-writeLines(paste('Using following covariates in the model:',
-                 paste(adjust.covars, collapse=', '),
-                 'as fixed effects and Individual_ID is chosen as random effect'))
-
-# Post adjusted formula
-random_effect <- "Individual_ID"
-formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-
-# Estimate voom weights with DREAM
-geneExpr = DGEList(NEW.COUNTS)
-geneExpr = edgeR::calcNormFactors(geneExpr)
-
-VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
-                                                               formula = formula,
-                                                               data = COVARIATES,
-                                                               save.plot = TRUE)
-
-# How to plot a voom object if save.plot=TRUE
-# Set plot generic to voom result EList
-#' @export
-setMethod("plot", signature(x="EList", y="missing"),
-  function(x,  y, ...){
-    if( is.null(x$voom.xy) || is.null(x$voom.line)){
-        stop("x does not contain the trend information.\nvoom() must be run with save.plot=TRUE")
-    }
-    # points
-    df = data.frame(x = x$voom.xy$x,
-                    y = x$voom.xy$y)
-    # trend line
-    df.line = data.frame(x = x$voom.line$x,
-                        y = x$voom.line$y)
-    ggplot(df, aes(x,y)) + geom_point(size=.1) + theme_bw(15) + 
-      theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + 
-      geom_line(data=df.line, aes(x,y), color="red") + xlab(bquote(log[2](count + 0.5))) + 
-      ylab(expression( sqrt("standard deviation"))) + ggtitle("Voom: Mean-variance trend") + 
-      ylim(0, max(df$y))
-  }
-)
-
-plot(VOOM.GENE_EXPRESSION)
-
-# Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-Dx.Tissue_ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
-                                        formula = formula,
-                                        data = COVARIATES,
-                                        computeResiduals = TRUE)
-
-col_type <- dplyr::select(COVARIATES, c(adjust.covars, "Individual_ID")) %>%
-  dplyr::summarise_all(class) %>% 
-  tidyr::gather(variable, class)
-
-# Model categorical variables as random to report variance fractions
-formula_for_variance_fractions = sapply(1:length(col_type$class), function(i){
-  switch(col_type$class[i], 
-    "factor" = paste0('(1|', col_type$variable[i], ')'),
-    "numeric" = col_type$variable[i])
-  })
-
-formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(formula_for_variance_fractions, sep = " + "))
-
-violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
-                                                         formula = formula_for_variance_fractions,
-                                                         data = COVARIATES)
-
-plotVarPart(sortCols(violin_plot))
-
-# Residuals after normalisation
-Dx.Tissue_RESIDUAL.GENE_EXPRESSION = residuals(Dx.Tissue_ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(Dx.Tissue_RESIDUAL.GENE_EXPRESSION, COVARIATES,
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-```
-
-### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
-
-### Tissue Correlation
-```{r tissue_correlation}
-hasReps = rownameToFirstColumn(COVARIATES, "SampleID") %>%
-  dplyr::select(Individual_ID) %>% 
-  table
-
-df_dlpfc = rownameToFirstColumn(COVARIATES, "SampleID") %>% 
-  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
-  filter(grepl("DLPFC", Dx.Tissue)) %>% 
-  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
-df_acc = rownameToFirstColumn(COVARIATES, "SampleID") %>%
-  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
-  filter(grepl("ACC", Dx.Tissue)) %>% 
-  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
-
-df_merge = merge(df_dlpfc, df_acc, by="Individual_ID")
-
-res_std <- rownameToFirstColumn(Dx.Tissue_RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id")
-
-corRes = sapply(res_std$ensembl_gene_id, function(ensGene){
-
-  x = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.x,drop=FALSE]
-  y = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.y,drop=FALSE]
-
-  cor(t(x)[,1], t(y)[,1], use="pairwise.complete.obs")
-})
-
-hist(corRes, xlab="Correlation", main="Correlation between ACC and DLPFC sample from the sample individual for each gene from ")
-
-pca = prcomp(t(VOOM.GENE_EXPRESSION$E))
-
-df_pca = merge(pca$x, as.data.frame(rownameToFirstColumn(COVARIATES, "SampleID")), by.x="row.names", by.y="SampleID" ) %>% 
-  tidyr::separate(Dx.Tissue, into = c("Dx", "Tissue"), sep = "_")
-
-ggplot(df_pca, aes(PC1, PC2, color=Tissue, shape=Institution)) + geom_point() + theme_bw(17) + theme(aspect.ratio=1)
-```
-
-### Residual calculation
-Calculate weighted residuals and add back "Dx.Tissue" to the residuals
-```{r varsToAddBack}
-# Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(Dx.Tissue_ADJUSTED.FIT$design), value = T)
-Dx.Tissue_RESIDUAL.GENE_EXPRESSION = Dx.Tissue_RESIDUAL.GENE_EXPRESSION +
-  Dx.Tissue_ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(Dx.Tissue_ADJUSTED.FIT$design[,varsToAddIn])
-```
-
-### Coexpression of residual expression 
-```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(Dx.Tissue_RESIDUAL.GENE_EXPRESSION))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-
-### Clustering residual data
-```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
-# Find principal components of expression to plot
-PC <- prcomp(tmp, scale.=T, center = T)
-
-# Plot first 4 PCs
-plotdata <- data.frame(SampleID=rownames(PC$rotation), 
-                       PC1=PC$rotation[,1], 
-                       PC2=PC$rotation[,2])
-
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
-
-p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
-p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
-p
-```
-
-```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
-# Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Institution')])
-
-tree = hclust(as.dist(t(tmp)))
-cols = WGCNA::labels2colors(COVARIATES.tmp);
-
-WGCNA::plotDendroAndColors(tree, 
-                           colors = cols, 
-                           dendroLabels = FALSE, 
-                           abHeight = 0.80, 
-                           main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Institution')]))
-```
-
-```{r temp1, include=F}
-dev.off()
-gc()
-```
-
-### Differential expression analysis (SCZ - Control - Other)
-Genes that are differentially expressed at an FDR <= 0.05 are:
-```{r set-formula}
-formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-```
-
-`r formula`
-
-```{r diffExp, fig.height=10, fig.width=15}
-L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueOther_ACC", "Dx.TissueControl_ACC"))
-L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueControl_ACC"))
-L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueOther_ACC"))
-L4 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueOther_DLPFC", "Dx.TissueControl_DLPFC"))
-L5 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueControl_DLPFC"))
-L6 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueOther_DLPFC"))
-
-L = cbind(L1, L2, L3, L4, L5, L6)
-
-contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
-              glue::glue_collapse(names(L2[L2 != 0]), "-"),
-              glue::glue_collapse(names(L3[L3 != 0]), "-"),
-              glue::glue_collapse(names(L4[L4 != 0]), "-"),
-              glue::glue_collapse(names(L5[L5 != 0]), "-"),
-              glue::glue_collapse(names(L6[L6 != 0]), "-")
-)
-
-# Visualize contrast matrix
-plotContrasts(L)
-
-# Fit contrast
-FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
-                                     formula = formula,
-                                     data = COVARIATES,
-                                     L = L)
-
-# Get differential expression
-DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
-  topTable(FIT, coef=i, number = Inf) %>%
-    rownameToFirstColumn('gene_id') %>% 
-    left_join(backgroundGenes)
-}, FIT.CONTR) 
-names(DE) <- contrast
-
-DE = DE %>% 
-  rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
-                Direction = logFC/abs(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
-                Direction = as.character(Direction)) %>%
-  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
-  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
-  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
-  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
-  left_join(Ensemble2HGNC) %>%
-  left_join(gene_lengths %>% rownameToFirstColumn('gene_id'))
-DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
-
-tmp = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tissue) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
-
-tmp1 = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tissue) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
-
-kable(full_join(tmp,tmp1))
-
-p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
-p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
-p
-```
-
-### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de, fig.height=10, fig.width=15}
-pl = list()
-pl[[1]] = ggplot(DE %>% 
-                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
-pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
-pl[[1]] = pl[[1]] + theme(legend.position = 'top')
-
-pl[[2]] = ggplot(DE %>% 
-                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = percentage_gene_gc_content, y = logFC, color = Direction)) + geom_point()
-pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gene_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
-pl[[2]] = pl[[2]] + theme(legend.position = 'top')
-
-pl[[3]] = ggplot(DE %>% 
-                    arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
-pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
-pl[[3]] = pl[[3]] + theme(legend.position = 'top')
-
-multiplot(plotlist = pl, cols = 3)
-```
-
-### Normalisation - Dx.Tissue.Gender
-1. Dx.Tissue.Gender is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis, tissue and reported gender)
-2. Individual_ID is chosen as random effect
-```{r iterative.normalisation-gender, results='asis'}
-COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Reported_Gender, sep = "_"))
-COVARIATES$Reported_Gender <- NULL
-COVARIATES$Dx.Tissue <- NULL
-adjust.covars[1] <- "Dx.Tissue.Gender"
-
-writeLines(paste('Using following covariates in the model:',
-                 paste(adjust.covars, collapse=', '),
-                 'as fixed effects and Individual_ID is chosen as random effect'))
-
-# Post adjusted formula
-random_effect <- "Individual_ID"
-formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-
-# Estimate voom weights with DREAM
-geneExpr = DGEList(NEW.COUNTS)
-geneExpr = edgeR::calcNormFactors(geneExpr)
-
-VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
-                                                               formula = formula,
-                                                               data = COVARIATES,
-                                                               save.plot = TRUE)
-
-plot(VOOM.GENE_EXPRESSION)
-
-# Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-Dx.Tissue.Gender_ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
-                                        formula = formula,
-                                        data = COVARIATES,
-                                        computeResiduals = TRUE)
-
-col_type <- dplyr::select(COVARIATES, c(adjust.covars, "Individual_ID")) %>%
-  dplyr::summarise_all(class) %>% 
-  tidyr::gather(variable, class)
-
-# Model categorical variables as random to report variance fractions
-formula_for_variance_fractions = sapply(1:length(col_type$class), function(i){
-  switch(col_type$class[i], 
-    "factor" = paste0('(1|', col_type$variable[i], ')'),
-    "numeric" = col_type$variable[i])
-  })
-
-formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(formula_for_variance_fractions, sep = " + "))
-
-violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
-                                                         formula = formula_for_variance_fractions,
-                                                         data = COVARIATES)
-plotVarPart(sortCols(violin_plot))
-
-# Residuals after normalisation
-Dx.Tissue.Gender_RESIDUAL.GENE_EXPRESSION = residuals(Dx.Tissue.Gender_ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(Dx.Tissue.Gender_RESIDUAL.GENE_EXPRESSION, COVARIATES,
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-```
-
-### Sanity check
-```{r residualSigCovars.manual-gender, fig.width=12, fig.height=8}
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
-
-### Differential expression analysis (Gender-specific SCZ - Control - Other)
-Genes that are differentially expressed at an FDR <= 0.05 are:
-```{r set-formula-gender}
-formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-```
-
-`r formula`
-
-```{r diffExp-gender, fig.height=10, fig.width=15}
-L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.Tissue.GenderSCZ_ACC_Female", "Dx.Tissue.GenderControl_ACC_Female"))
-L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.Tissue.GenderSCZ_ACC_Male", "Dx.Tissue.GenderControl_ACC_Male"))
-L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.Tissue.GenderSCZ_DLPFC_Female", "Dx.Tissue.GenderControl_DLPFC_Female"))
-L4 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.Tissue.GenderSCZ_DLPFC_Male", "Dx.Tissue.GenderControl_DLPFC_Male"))
-
-L = cbind(L1, L2, L3, L4)
-
-contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
-              glue::glue_collapse(names(L2[L2 != 0]), "-"),
-              glue::glue_collapse(names(L3[L3 != 0]), "-"),
-              glue::glue_collapse(names(L4[L4 != 0]), "-")
-)
-
-# Visualize contrast matrix
-plotContrasts(L)
-
-# Fit contrast
-FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
-                                     formula = formula,
-                                     data = COVARIATES,
-                                     L = L)
-
-# Get differential expression
-DE_2 = lapply(glue::glue("L{1:4}"), function(i, FIT){
-  topTable(FIT, coef=i, number = Inf) %>%
-    rownameToFirstColumn('gene_id') %>% 
-    left_join(backgroundGenes)
-}, FIT.CONTR) 
-names(DE_2) <- contrast
-
-DE_2 = DE_2%>% 
-  rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('Dx.Tissue.Gender','',Comparison),
-                Direction = logFC/abs(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
-                Direction = as.character(Direction)) %>% 
-  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
-  tidyr::separate(to.state, into = c('to.Dx', 'to.Tissue', 'to.Gender'), sep = '_') %>%
-  tidyr::separate(ref.state, into = c('ref.Dx', 'ref.Tissue', 'ref.Gender'), sep = '_') %>% 
-  left_join(Ensemble2HGNC) %>%
-  left_join(gene_lengths %>% rownameToFirstColumn('gene_id'))
-DE_2$Direction[DE_2$adj.P.Val > 0.05 | abs(DE_2$logFC) < log2(1.2)] = 'NONE'
-
-tmp = DE_2 %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
-  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
-
-tmp1 = DE_2 %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
-  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
-
-kable(full_join(tmp,tmp1))
-
-p = ggplot(DE_2, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
-p = p + facet_grid(ref.Tissue~.+ref.Gender, scales = 'fixed')
-p = p + ggtitle("Differential Expression between Schizophrenia and Control")
-p
-```
-
-### Normalisation - Tissue
-1. Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and reported gender)
-2. Individual_ID is chosen as random effect
-```{r iterative.normalisation-tissue, results='asis'}
-###we might not need this adjustment
-COVARIATES <- tidyr::separate(COVARIATES, Dx.Tissue.Gender, c("Dx", "Tissue", "Reported_Gender")) %>% 
-  dplyr::mutate(Dx = factor(Dx, levels = c("SCZ", "Control", "Other")),
-                Tissue = factor(Tissue), 
-                Reported_Gender = factor(Reported_Gender))
-
-adjust.covars[1] <- "Dx"
-
-adjust.covars <- c(adjust.covars, "Tissue", "Reported_Gender")
-
-writeLines(paste('Using following covariates in the model:',
-                 paste(adjust.covars, collapse=', '),
-                 'as fixed effects and Individual_ID is chosen as random effect'))
-
-# Post adjusted formula
-random_effect <- "Individual_ID"
-formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-
-# Estimate voom weights with DREAM
-geneExpr = DGEList(NEW.COUNTS)
-geneExpr = edgeR::calcNormFactors(geneExpr)
-
-VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
-                                                               formula = formula,
-                                                               data = COVARIATES,
-                                                               save.plot = TRUE)
-
-plot(VOOM.GENE_EXPRESSION)
-
-# Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
-Tissue_ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
-                                        formula = formula,
-                                        data = COVARIATES,
-                                        computeResiduals = TRUE)
-
-col_type <- dplyr::select(COVARIATES, c(adjust.covars, "Individual_ID")) %>%
-  dplyr::summarise_all(class) %>% 
-  tidyr::gather(variable, class)
-
-# Model categorical variables as random to report variance fractions
-formula_for_variance_fractions = sapply(1:length(col_type$class), function(i){
-  switch(col_type$class[i], 
-    "factor" = paste0('(1|', col_type$variable[i], ')'),
-    "numeric" = col_type$variable[i])
-  })
-
-formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(formula_for_variance_fractions, sep = " + "))
-
-violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
-                                                         formula = formula_for_variance_fractions,
-                                                         data = COVARIATES)
-plotVarPart(sortCols(violin_plot))
-
-# Residuals after normalisation
-Tissue_RESIDUAL.GENE_EXPRESSION = residuals(Tissue_ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(Tissue_RESIDUAL.GENE_EXPRESSION, COVARIATES,
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-```
-
-### Sanity check
-```{r residualSigCovars.manual-tissue, fig.width=12, fig.height=8}
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
-
-### Differential expression analysis (ACC - DLPFC)
-Genes that are differentially expressed at an FDR <= 0.05 are:
-```{r set-formula-tissue}
-formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-```
-
-`r formula`
-
-```{r diffExp-tissue, fig.height=10, fig.width=15}
-L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("TissueDLPFC"))
-L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("DxSCZ", "DxControl"))
-L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Reported_GenderMale"))
-
-L = cbind(L1, L2, L3)
-
-contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
-              glue::glue_collapse(names(L2[L2 != 0]), "-"),
-              glue::glue_collapse(names(L3[L3 != 0]), "-")
-)
-
-# Visualize contrast matrix
-plotContrasts(L)
-
-# Fit contrast
-FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
-                                     formula = formula,
-                                     data = COVARIATES,
-                                     L = L)
-
-# Get differential expression
-DE_3 = lapply(glue::glue("L{1:3}"), function(i, FIT){
-  topTable(FIT, coef=i, number = Inf) %>%
-    rownameToFirstColumn('gene_id') %>% 
-    left_join(backgroundGenes)
-}, FIT.CONTR) 
-names(DE_3) <- contrast
-
-DE_3 = DE_3%>% 
-  rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
-                Direction = logFC/abs(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
-                Direction = as.character(Direction)) %>% 
-  dplyr::mutate(Comparison = ifelse(Comparison %in% "Reported_GenderMale", "Female-Male", Comparison),
-                Comparison = ifelse(Comparison %in% "DxSCZ-DxControl", "SCZ-Control", Comparison),
-                Comparison = ifelse(Comparison %in% "TissueDLPFC", "ACC-DLPFC", Comparison)) %>% 
-  left_join(Ensemble2HGNC) %>%
-  left_join(gene_lengths %>% rownameToFirstColumn('gene_id'))
-DE_3$Direction[DE_3$adj.P.Val > 0.05 | abs(DE_3$logFC) < log2(1.2)] = 'NONE'
-
-tmp = DE_3 %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
-
-tmp1 = DE_3 %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
-
-kable(full_join(tmp,tmp1))
-
-p = ggplot(DE_3, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
-p = p + facet_grid(Comparison~., scales = 'fixed')
-p
-```
-
 ### Store files in synapse
@@
 synSetAnnotations(OFFSET_OBJ, annotations = all.annotations)
-
-# Store residual gene expression for network analysis - Dx.Tissue
-Dx.Tissue_RESIDUAL.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_Dx.Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue_netResidualExpression.tsv', 
-                name = 'Dx.Tissue Normalised, covariates removed residual expression (for network analysis)', 
-                parentId = CODE$properties$id)
-nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(nEXP_OBJ, annotations = all.annotations)
-
-# Store residual gene expression for network analysis - Dx.Tissue.Gender
-Dx.Tissue.Gender_RESIDUAL.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_Dx.Tissue.Gender_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue.Gender_netResidualExpression.tsv', 
-                name = 'Dx.Tissue.Gender Normalised, covariates removed residual expression (for network analysis)', 
-                parentId = CODE$properties$id)
-nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(nEXP_OBJ, annotations = all.annotations)
-
-# Store residual gene expression for network analysis - Tissue
-Tissue_RESIDUAL.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_DLPFC_Tissue_netResidualExpression.tsv', 
-                name = 'Tissue Normalised, covariates removed residual expression (for network analysis)', 
-                parentId = CODE$properties$id)
-nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(nEXP_OBJ, annotations = all.annotations)
-
-# Store differential expression results - Dx.Tissue
-write.table(DE, file = 'CMC_ACC_DLPFC_Dx.Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue_DiffExpression.tsv', 
-                name = 'Differential Expression Results (Dx.Tissue)', 
-                parentId = CODE$properties$id)
-DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(DEXP_OBJ, annotations = all.annotations)
-
-# Store differential expression results - Dx.Tissue.Gender
-write.table(DE_2, file = 'CMC_ACC_DLPFC_Dx.Tissue.Gender_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue.Gender_DiffExpression.tsv', 
-                name = 'Differential Expression Results (Dx.Tissue.Gender)', 
-                parentId = CODE$properties$id)
-DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(DEXP_OBJ, annotations = all.annotations)
-
-# Store differential expression results - Tissue
-write.table(DE_3, file = 'CMC_ACC_DLPFC_Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_ACC_DLPFC_Tissue_DiffExpression.tsv', 
-                name = 'Differential Expression Results (Tissue)', 
-                parentId = CODE$properties$id)
-DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(DEXP_OBJ, annotations = all.annotations)
-
-# Store adjusted fit linear mixed model object as RDS - Dx.Tissue
-saveRDS(Dx.Tissue_ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_Dx.Tissue_dream-output.rds')
-ADJ_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue_dream-output.rds',
-               name = 'Dream Output object - Dx.Tissue', 
-               parentId = CODE$properties$id)
-ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
-                   executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
-
-# Store adjusted fit linear mixed model object as RDS - Dx.Tissue.Gender
-saveRDS(Dx.Tissue.Gender_ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_Dx.Tissue.Gender_dream-output.rds')
-ADJ_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue.Gender_dream-output.rds',
-               name = 'Dream Output object - Dx.Tissue.Gender', 
-               parentId = CODE$properties$id)
-
-ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
-                   executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
-
-# Store adjusted fit linear mixed model object as RDS - Tissue
-saveRDS(Dx.Tissue.Gender_ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_Tissue_dream-output.rds')
-ADJ_OBJ = File('CMC_ACC_DLPFC_Tissue_dream-output.rds',
-               name = 'Dream Output object - Tissue', 
-               parentId = CODE$properties$id)
-ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
-                   executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
 ```
@@
 |  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
-|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
-|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
-|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
-|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |
 

### 2020-03-27 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
add model selection

@@
 
-#### Model Selection
-```{r set_vars}
-adjust.covars <- c("Dx.Tissue", "RIN", "IntronicRate")
+### Model Identification
+```{r fxn}
+# Function to fit a mixed effect model and return BIC
+fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){
+
+  # Get design matrix
+  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
+  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]
+
+  # Estimate voom weights
+  cnt[is.na(cnt)] = 0
+  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights
+
+  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
+                              .combine = rbind,
+                              .packages = c('lme4', 'dplyr'),
+                              .export = c('voomWeights')) %dopar% {
+
+                                # Perform weighted mixed linear models for every gene
+                                expr = data.frame(Gx = lcpm[i,])
+                                data = cbind(expr,
+                                             designMat[rownames(expr),],
+                                             Individual_ID = groupingVar)
+                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
+                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')
+
+                                weights = voomWeights[i,]
+                                BIC = NA
+                                tryCatch({
+                                  mdl = lmer(formula = formula, 
+                                             data = data, 
+                                             weights = weights)
+                                  BIC = BIC(mdl)
+                                }, error = function(e){
+                                  BIC = NA
+                                })
+                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
+                                                BIC = BIC))
+                            }
+  return(modelBIC)
+  }
+# Function to select covariates in a step-wise regression fashion
+selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){
+
+  varsSelected = c()
+  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
+           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
+  while(length(covarsToSelect) != 0){
+    allModelBIC = foreach(i = 1:length(covarsToSelect),
+                          .combine = c,
+                          .packages = pkgs,
+                          .export = c('fitMixedEffectModel')) %dopar% {
+
+                            # Fit a mixed effect linear model and get all BIC
+                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
+                                                           covars[colnames(cnt),'Individual_ID'], 
+                                                           lcpm, 
+                                                           cnt, 
+                                                           covars) %>%
+                            plyr::rename(c('BIC' = covarsToSelect[i]))
+
+                            return(list(modelBIC))
+                          }
+    allModelBIC = plyr::join_all(allModelBIC)
+
+    # Find number of genes that are changed based on BIC
+    nChange = allModelBIC %>%
+      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
+      dplyr::left_join(baseModelBIC) %>%
+      dplyr::mutate(diff = baseModel - Value) %>%
+      dplyr::group_by(ModelName) %>%
+      dplyr::summarise(better = sum(diff >= 5), 
+                       worse = sum(diff <= -5)) %>%
+      dplyr::mutate(nchange = better - worse)
+
+    print(nChange)
+
+    # Select variable
+    tmp = nChange %>%
+      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
+      dplyr::top_n(1, nchange)
+
+    if (dim(tmp)[1] == 0){
+      covarsToSelect = c()
+    } else {
+      baseModelBIC = allModelBIC %>%
+        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
+      colnames(baseModelBIC)[2] = 'baseModel'
+      varsSelected = c(varsSelected, tmp$ModelName)
+      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
+      print(paste('Variable chosen to be added:',tmp$ModelName))
+    }
+  }
+
+  return(varsSelected)
+}
 ```
+#### Phase I (clinical, ancestry and sample specific technical variables)
+```{r phase1}
+# Fit a base model with the following covariates
+adjust.covars = c('Dx.Tissue')
+baseModelBIC = fitMixedEffectModel(adjust.covars, 
+                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
+                                   CQN.GENE_EXPRESSION$E, 
+                                   NEW.COUNTS, 
+                                   COVARIATES) %>%
+  dplyr::rename(baseModel = BIC)
+# Iteratively identify covariates in phase I
+phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'ageOfDeath', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
+vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
+                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
+                                     NEW.COUNTS, baseModelBIC)
+adjust.covars = c(adjust.covars, vars.selected)
+```
+Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`
 
-[Code](https://github.com/kelshmo/analysis/blob/f7345adde2052afbc258773f69930df300989bf3/cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd#L579-L737) for model selection and [BIC Criteria](https://www.synapse.org/#!Synapse:syn21543160)
+#### Phase II (batch effects)
+```{r phase2}
+# Fit a base model with the following covariates
+baseModelBIC = fitMixedEffectModel(adjust.covars, 
+                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
+                                   CQN.GENE_EXPRESSION$E, 
+                                   NEW.COUNTS, 
+                                   COVARIATES) %>%
+  dplyr::rename(baseModel = BIC)
+# Iteratively identify covariates in phase I
+phase2.covars = c('LibraryBatch', 'FlowcellBatch')
+vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
+                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
+                                     NEW.COUNTS, baseModelBIC)
+adjust.covars = c(adjust.covars, vars.selected)
+```
+Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`
+
+#### Phase III (RNASeq alignment specific covariates)
+```{r phase3}
+# Fit a base model with the following covariates
+baseModelBIC = fitMixedEffectModel(adjust.covars, 
+                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
+                                   CQN.GENE_EXPRESSION$E, 
+                                   NEW.COUNTS, 
+                                   COVARIATES) %>%
+  dplyr::rename(baseModel = BIC)
+# Iteratively identify covariates in phase I
+phase3.covars = c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected", "rRNARate", "TotalReads")
+vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
+                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
+                                     NEW.COUNTS, baseModelBIC)
+adjust.covars = c(adjust.covars, vars.selected)
+```
 

### 2020-03-27 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix reference to synapse objects

@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = 'syn9872607';
+parentId = 'syn21435176';
 activityName = 'Covariate adjustments';
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = metadata
-metadata = downloadFile_version(metadata, version = 14)  %>% 
+metadata = downloadFile_version(metadata, version = 16)  %>% 
   dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, 

### 2020-03-27 cmc-rnaseq-de/Differential-Expression.Rmd
markdown specific to DE

---
title: "Differential Expression"
author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
date: "`r date()`"
output: html_document
---

```{r knit2synapse, eval=FALSE}
library(synapser)
library(knit2synapse)

synLogin()

knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/Differential-Expression.Rmd",
                                 parentId ="syn21824667",
                                 folderName = "Differential Expression - Dx.Tissue")
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)
library(R.utils)
library(ggplot2)
library(limma)
library(lme4)
library(biomaRt)
library(variancePartition)
library(synapser)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

synLogin()

library(doParallel)
library(foreach)
library(future)

cl = makeCluster((availableCores()-2)/2)

registerDoParallel(cl)

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE)
```

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn21824667';
activityName = 'Differential Expression';
activityDescription = 'Differential Expression - Dx.Tissue';

thisFileName <- 'Differential-Expression.Rmd'

# Github link
thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='add-dream')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))
```

### Data download
#### Obtain filtered and CQN normalized counts
```{r download.data, cache=TRUE}
downloadFile <- function(id){
  fread(synGet(id)$path, data.table = F)
}
downloadFile_version <- function(id , version){
  fread(synGet(id, version = version)$path, data.table = F)
}
# Download CQN counts
CQN_ID = "syn21825509"
ALL_USED_IDs = c(ALL_USED_IDs, CQN_ID)
CQN.GENE_EXPRESSION$E = downloadFile(CQN_ID) %>% 
  tibble::column_to_rownames(var = "ensembl_gene_id")

# Download filtered counts
COUNTS_ID = "syn21825508"
ALL_USED_IDs = c(ALL_USED_IDs, COUNTS_ID)
NEW.COUNTS = downloadFile(COUNTS_ID) %>% 
  tibble::column_to_rownames(var = "ensembl_gene_id")

# Get covariates
COVAR_ID = "syn21825507"
ALL_USED_IDs = c(ALL_USED_IDs, COVAR_ID)
COVARIATES = downloadFile(COVAR_ID)

# Set covariates to adjust for from model selection
adjust.covars <- c("Dx.Tissue", "RIN", "IntronicRate")
```

### Normalisation - Dx.Tissue
1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
2. Individual_ID is chosen as random effect

```{r iterative.normalisation, results='asis'}
writeLines(paste('Using following covariates in the model:',
                 paste(adjust.covars, collapse=', '),
                 'as fixed effects and Individual_ID is chosen as random effect'))

# Post adjusted formula
random_effect <- "Individual_ID"
formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
  glue::glue(" + (1|{random_effect})")

# Estimate voom weights with DREAM
geneExpr = DGEList(NEW.COUNTS)
geneExpr = edgeR::calcNormFactors(geneExpr)

VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
                                                               formula = formula,
                                                               data = COVARIATES,
                                                               save.plot = TRUE)

# How to plot a voom object if save.plot=TRUE
# Set plot generic to voom result EList
#' @export
setMethod("plot", signature(x="EList", y="missing"),
  function(x,  y, ...){
    if( is.null(x$voom.xy) || is.null(x$voom.line)){
        stop("x does not contain the trend information.\nvoom() must be run with save.plot=TRUE")
    }
    # points
    df = data.frame(x = x$voom.xy$x,
                    y = x$voom.xy$y)
    # trend line
    df.line = data.frame(x = x$voom.line$x,
                        y = x$voom.line$y)
    ggplot(df, aes(x,y)) + geom_point(size=.1) + theme_bw(15) + 
      theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + 
      geom_line(data=df.line, aes(x,y), color="red") + xlab(bquote(log[2](count + 0.5))) + 
      ylab(expression( sqrt("standard deviation"))) + ggtitle("Voom: Mean-variance trend") + 
      ylim(0, max(df$y))
  }
)

plot(VOOM.GENE_EXPRESSION)

# Fit linear model using new weights and new design
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
                                        formula = formula,
                                        data = COVARIATES,
                                        computeResiduals = TRUE)

col_type <- dplyr::select(COVARIATES, c(adjust.covars, "Individual_ID")) %>%
  dplyr::summarise_all(class) %>% 
  tidyr::gather(variable, class)

# Model categorical variables as random to report variance fractions
formula_for_variance_fractions = sapply(1:length(col_type$class), function(i){
  switch(col_type$class[i], 
    "factor" = paste0('(1|', col_type$variable[i], ')'),
    "numeric" = col_type$variable[i])
  })

formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(formula_for_variance_fractions, sep = " + "))

violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
                                                         formula = formula_for_variance_fractions,
                                                         data = COVARIATES)

plotVarPart(sortCols(violin_plot))

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Tissue Correlation
```{r tissue_correlation}
hasReps = rownameToFirstColumn(COVARIATES, "SampleID") %>%
  dplyr::select(Individual_ID) %>% 
  table

df_dlpfc = rownameToFirstColumn(COVARIATES, "SampleID") %>% 
  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
  filter(grepl("DLPFC", Dx.Tissue)) %>% 
  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
df_acc = rownameToFirstColumn(COVARIATES, "SampleID") %>%
  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
  filter(grepl("ACC", Dx.Tissue)) %>% 
  dplyr::select(SampleID, Individual_ID, Dx.Tissue)

df_merge = merge(df_dlpfc, df_acc, by="Individual_ID")

res_std <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id")

corRes = sapply(res_std$ensembl_gene_id, function(ensGene){

  x = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.x,drop=FALSE]
  y = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.y,drop=FALSE]

  cor(t(x)[,1], t(y)[,1], use="pairwise.complete.obs")
})

hist(corRes, xlab="Correlation", main="Correlation between ACC and DLPFC sample from the sample individual for each gene from ")

pca = prcomp(t(VOOM.GENE_EXPRESSION$E))

df_pca = merge(pca$x, as.data.frame(rownameToFirstColumn(COVARIATES, "SampleID")), 
               by.x="row.names", 
               by.y="SampleID" ) %>% 
  tidyr::separate(Dx.Tissue, into = c("Dx", "Tissue"), sep = "_")

ggplot(df_pca, aes(PC1, PC2, color=Tissue, shape=Institution)) + geom_point() + theme_bw(17) + theme(aspect.ratio=1)
```

### Residual calculation
Calculate weighted residuals and add back "Dx.Tissue" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(ADJUSTED.FIT$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
```

### Coexpression of residual expression 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(tmp, scale.=T, center = T)

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
p
```

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Institution')])

tree = hclust(as.dist(t(tmp)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Institution')]))
```

```{r temp1, include=F}
dev.off()
gc()
```

### Differential expression analysis (SCZ - Control - Other)
Genes that are differentially expressed at an FDR <= 0.05 are:
```{r set-formula}
formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
  glue::glue(" + (1|{random_effect})")
```

`r formula`

```{r diffExp, fig.height=10, fig.width=15}
L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueOther_ACC", "Dx.TissueControl_ACC"))
L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueControl_ACC"))
L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueOther_ACC"))
L4 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueOther_DLPFC", "Dx.TissueControl_DLPFC"))
L5 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueControl_DLPFC"))
L6 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueOther_DLPFC"))

L = cbind(L1, L2, L3, L4, L5, L6)

contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
              glue::glue_collapse(names(L2[L2 != 0]), "-"),
              glue::glue_collapse(names(L3[L3 != 0]), "-"),
              glue::glue_collapse(names(L4[L4 != 0]), "-"),
              glue::glue_collapse(names(L5[L5 != 0]), "-"),
              glue::glue_collapse(names(L6[L6 != 0]), "-")
)

# Visualize contrast matrix
plotContrasts(L)

# Fit contrast
FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
                                     formula = formula,
                                     data = COVARIATES,
                                     L = L)

# Get differential expression
DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
  topTable(FIT, coef=i, number = Inf) %>%
    rownameToFirstColumn('gene_id') %>% 
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) <- contrast

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(gene_lengths %>% rownameToFirstColumn('gene_id'))
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
  group_by(Comparison, Tissue) %>%
  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
  group_by(Comparison, Tissue) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))

kable(full_join(tmp,tmp1))

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE %>% 
                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE %>% 
                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = percentage_gene_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gene_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE %>% 
                    arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

multiplot(plotlist = pl, cols = 3)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Set annotations
all.annotations = list(
  dataType = 'geneExpression',
  dataSubtype = 'processed',
  assay	 = 'rnaSeq',

  tissue	= 'dorsolateral prefrontal cortex, anterior cingulate cortex', 
  study = 'CMC',

  species = 'Human',
  consortium	= 'CMC',

  analysisType	= "data normalization",

  transcriptNormalizationMethod	= 'CQN',
  transcriptQuantificationMethod = 'featureCounts',
  referenceSet = 'GRCh38'
)

# Code
CODE <- Folder(name = "Differential Expression - Dx.Tissue", parentId = parentId)
CODE <- synStore(CODE)

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Dx.Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue_netResidualExpression.tsv', 
                name = 'Dx.Tissue Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(nEXP_OBJ, annotations = all.annotations)

# Store differential expression results
write.table(DE, file = 'CMC_ACC_DLPFC_Dx.Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue_DiffExpression.tsv', 
                name = 'Differential Expression Results (Dx.Tissue)', 
                parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(DEXP_OBJ, annotations = all.annotations)

# Store adjusted fit linear mixed model object as RDS - Dx.Tissue
saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_Dx.Tissue_dream-output.rds')
ADJ_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue_dream-output.rds',
               name = 'Dream Output object - Dx.Tissue', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
```

### 2020-03-27 cmc-rnaseq-de/Differential-Expression.Rmd
error with inputs

@@
 }
-downloadFile_version <- function(id , version){
-  fread(synGet(id, version = version)$path, data.table = F)
-}
+
 # Download CQN counts
 CQN_ID = "syn21825509"
-ALL_USED_IDs = c(ALL_USED_IDs, CQN_ID)
-CQN.GENE_EXPRESSION$E = downloadFile(CQN_ID) %>% 
+ALL_USED_IDs = c(CQN_ID)
+CQN.GENE_EXPRESSION = downloadFile(CQN_ID) %>% 
   tibble::column_to_rownames(var = "ensembl_gene_id")
@@
 # Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION
 ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
@@
 # Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
+RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION)
 

### 2020-03-30 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
parse out DE

@@
 ---
-title: "HBCC - ACC-Nirmala - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
-author: "Kelsey Montgomery and Thanneer Perumal"
+title: "HBCC - ACC-Nirmala - FEATURECOUNTS - GRCh38 - CQN"
+author: "Kelsey Montgomery, Gabriel Hoffman and Thanneer Perumal"
 date: "`r date()`"
@@
                                           parentId ="syn21435176",
-                                          folderName = 'HBCC - ACC-Nirmala - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
+                                          folderName = 'HBCC - ACC-Nirmala - DREAM - FEATURECOUNTS - GRCh38 - CQN')
 ```
@@
 library(foreach)
+library(future)
 
-cl = makeCluster(detectCores()-2)
+cl = makeCluster((availableCores()-2)/2)
 
@@
 ## Download reprocessed counts (ACC)
-counts = 'syn21444441'
+counts = 'syn21754352'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = counts
@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "sep2019.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
+                host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)
+                dataset = "hsapiens_gene_ensembl")
 
@@
 ```
+
 ```{r temp, include = F}
@@
 ```
+
+#### Variables included in the model 
+Covariates were added as fixed effects iteratively in three phases if model improvement by BIC criteria was observed in the majority of genes.
+Clinical, ancestry and sample-specific technical variables were tested for model improvement first.
+Covariates related to batch effects were next tested for model improvement and finally, RNASeq alignment-specific covariates.
+
 Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`
-
-### Normalisation
-1. Dx is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
-2. Individual_ID is chosen as random effect
-```{r iterative.normalisation, results='asis'}
-writeLines(paste('Using following covariates in the model:',
-                 paste(adjust.covars, collapse=', '),
-                 'as fixed effects and Individual_ID is chosen as random effect'))
-
-# Post adjusted formula
-random_effect <- "Individual_ID"
-formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-
-# Estimate voom weights with DREAM
-tmp = NEW.COUNTS
-tmp[is.na(tmp)] = 0
-geneExpr = DGEList(tmp)
-geneExpr = edgeR::calcNormFactors(geneExpr)
-
-VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
-                                                               formula = formula,
-                                                               data = COVARIATES,
-                                                               save.plot = TRUE)
-plot(VOOM.GENE_EXPRESSION)
-
-violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
-                                                         formula = formula,
-                                                         data = COVARIATES)
-plotVarPart(sortCols(violin_plot))
-
-# CQN NA counts to 0 counts for input to dream()
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-VOOM.GENE_EXPRESSION$E = tmp
-ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
-                                        formula = formula,
-                                        data = COVARIATES,
-                                        computeResiduals = TRUE)
-
-# Residuals after normalisation
- RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-tmp = RESIDUAL.GENE_EXPRESSION
-tmp[is.na(tmp)] = 0
-residualSigCovars = runPCAandPlotCorrelations(tmp, COVARIATES,
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-```
-
-### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
-
-### Residual calculation
-Calculate weighted residuals and add back "Dx" to the residuals
-```{r varsToAddBack}
-# Add variable of interest back to the residuals
-varsToAddIn = grep("Dx", colnames(ADJUSTED.FIT$design), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-    ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
-```
-
-### Coexpression of residual expression 
-```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-tmp = RESIDUAL.GENE_EXPRESSION
-tmp[is.na(tmp)] = 0
-cr = cor(t(tmp))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-
-### Clustering residual data
-```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
-# Find principal components of expression to plot
-PC <- prcomp(tmp, scale.=T, center = T)
-
-# Plot first 4 PCs
-plotdata <- data.frame(SampleID=rownames(PC$rotation), 
-                       PC1=PC$rotation[,1], 
-                       PC2=PC$rotation[,2])
-
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
-
-p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx, scales = 'free_y')
-p
-```
-
-```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
-# Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx')])
-COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
-
-tree = hclust(as.dist(t(tmp)))
-cols = WGCNA::labels2colors(COVARIATES.tmp);
-
-WGCNA::plotDendroAndColors(tree, 
-                           colors = cols, 
-                           dendroLabels = FALSE, 
-                           abHeight = 0.80, 
-                           main = "Sample dendrogram",
-                           groupLabels = "Dx")
-```
-
-```{r temp1, include=F}
-dev.off()
-gc()
-```
-
-### Differential expression analysis
-Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=10, fig.width=15}
-formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueOther", "Dx.TissueControl"))
-L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ", "Dx.TissueControl"))
-L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ", "Dx.TissueOther"))
-L = cbind(L1, L2, L3)
-contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
-              glue::glue_collapse(names(L2[L2 != 0]), "-"),
-              glue::glue_collapse(names(L3[L3 != 0]), "-"))
-# Visualize contrast matrix
-plotContrasts(L) 
-#Fit contrast
-FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
-                                     formula = formula,
-                                     data = COVARIATES,
-                                     L = L)
-# Get differential expression
-DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
-  topTable(FIT, coef=i, number = Inf) %>%
-    rownameToFirstColumn('gene_id') %>%
-    left_join(backgroundGenes)
-}, FIT.CONTR) 
-names(DE) <- contrast
-
-DE = DE %>% 
-  rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('Dx','',Comparison),
-                Direction = logFC/abs(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
-                Direction = as.character(Direction)) %>%
-  left_join(Ensemble2HGNC) %>%
-  left_join(gene_lengths %>%
-              rownameToFirstColumn('gene_id'))
-DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
-
-tmp = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
-
-tmp1 = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
-
-kable(full_join(tmp,tmp1))
-
-p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
-p = p + facet_grid(.~Comparison, scales = 'fixed')
-p
-```
-
-### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de, fig.height=10, fig.width=15}
-pl = list()
-pl[[1]] = ggplot(DE %>% 
-                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
-pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('red','grey','green'))
-pl[[1]] = pl[[1]] + theme_bw() %+replace% theme(legend.position = 'top')
-
-pl[[2]] = ggplot(DE %>% 
-                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = percentage_gene_gc_content, y = logFC, color = Direction)) + geom_point()
-pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gene_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('red','grey','green'))
-pl[[2]] = pl[[2]] + theme_bw() %+replace% theme(legend.position = 'top')
-
-pl[[3]] = ggplot(DE %>% 
-                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
-pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('red','grey','green'))
-pl[[3]] = pl[[3]] + theme_bw() %+replace% theme(legend.position = 'top')
-
-multiplot(plotlist = pl, cols = 3)
-```
-
-### Gene set enrichment analysis
-Using Fisher's exact test for gene set enrichment analysis
-```{r enrich.de, fig.height=16, fig.width=16}
-diffexp.genes = DE %>%
-  dlply(.(Comparison), .fun = function(x){
-    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
-    tmp = tmp[!is.na(tmp)]
-  })
-diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
-backgroundGenes = unique(DE$hgnc_symbol)
-
-# Download all related genesets from synapse
-geneset.files = synGetChildren("syn3240583", includeTypes=list("file"))$asList()
-genesets.to.test = purrr::map(geneset.files, .f = function(fileList){
-  tmp = downloadFile(fileList$id) %>%
-    dlply(.(Name), .fun = function(x){
-      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
-    })
-  }) %>%
-  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
-  do.call(c,.)
-
-# Perform enrichment analysis
-enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
-   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
-     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
-    }, x, backgroundGenes, .parallel = TRUE) %>%
-    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
-}, genesets.to.test, backgroundGenes) %>%
-  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
-  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
-  dplyr::rename(Comparison = Name, Category = .id)
-```
 
@@
 # Code
-CODE <- Folder(name = "HBCC - ACC-Nirmala - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "HBCC - ACC-Nirmala - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
 CODE <- synStore(CODE)
@@
 synSetAnnotations(OFFSET_OBJ, annotations = all.annotations)
-
-# Store design matrix
-DM1$design %>%
-  rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'CMC_HBCC_ACC-Nirmala_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('CMC_HBCC_ACC-Nirmala_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
-DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                  executed = thisFile, activityDescription = activityDescription)
-all.annotations$dataSubType = 'designMatrix'
-synSetAnnotations(DM_OBJ, annotations = all.annotations)
-
-# Store residual gene expression for network analysis
-RESIDUAL.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_HBCC_ACC-Nirmala_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_HBCC_ACC-Nirmala_netResidualExpression.tsv', 
-                name = 'Normalised, covariates removed residual expression (for network analysis)', 
-                parentId = CODE$properties$id)
-nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-all.annotations$dataSubType = 'residualGeneExpForNetAnlz'
-synSetAnnotations(nEXP_OBJ, annotations = all.annotations)
-
-# Store differential expression results
-write.table(DE, file = 'CMC_HBCC_ACC-Nirmala_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_HBCC_ACC-Nirmala_DiffExpression.tsv', 
-                name = 'Differential Expression Results (Dx)', 
-                parentId = CODE$properties$id)
-DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-all.annotations$dataSubType = 'diffExp'
-synSetAnnotations(DEXP_OBJ, annotations = all.annotations)
-
-# Store enrichment results
-write.table(enrichment.results, file = 'CMC_HBCC_ACC-Nirmala_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
-ENRICH_OBJ = File('CMC_HBCC_ACC-Nirmala_EnrichResults.tsv', 
-                  name = 'Enrichment Analysis Results (Dx)', 
-                  parentId = CODE$properties$id)
-ENRICH_OBJ = synStore(ENRICH_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                      executed = thisFile, activityDescription = activityDescription)
-all.annotations$dataSubType = 'enrichResults'
-synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
-
-# Store adjusted fit linear mixed model object as RDS
-saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_dream-output.rds')
-ADJ_OBJ = File('CMC_ACC_DLPFC_dream-output.rds',
-               name = 'Dream Output object', 
-               parentId = CODE$properties$id)
-ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
-                   executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
 ```

### 2020-03-30 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
parse out DE

--- 2020-02-20 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
 ---
-title: "HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
-author: "Kelsey Montgomery and Thanneer Perumal"
+title: "HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN"
+author: "Kelsey Montgomery, Gabriel Hoffman and Thanneer Perumal"
 date: "`r date()`"
@@
                                           parentId ="syn21435176",
-                                          folderName = 'HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
+                                          folderName = 'HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN')
 ```
@@
 ALL_USED_IDs = METADATA_QC_ID
-metadata = downloadFile_version(METADATA_QC_ID, version = 15) %>%
+metadata = downloadFile_version(METADATA_QC_ID, version = 16) %>%
   dplyr::select(`Individual_ID`, Institution, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
@@
 ## Download reprocessed counts (ACC and STG)
-counts = 'syn21420322'
+counts = 'syn21754352'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = counts
@@
 COUNT$Strand <- NULL
-# Fix counts column header - remove "X" in sample name
-new_names <- gsub("X6", "6", colnames(COUNT))
-colnames(COUNT) <- new_names
 
@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "sep2019.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
+                host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020), 
+                dataset = "hsapiens_gene_ensembl")
 
@@
 ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-cr = cor(t(tmp))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+hist(CQN.GENE_EXPRESSION$E, main = 'Distribution of correlation between genes', xlab = 'Correlation')
 ```
@@
 ```
+#### Variables included in the model 
+Covariates were added as fixed effects iteratively in three phases if model improvement by BIC criteria was observed in the majority of genes.
+Clinical, ancestry and sample-specific technical variables were tested for model improvement first.
+Covariates related to batch effects were next tested for model improvement and finally, RNASeq alignment-specific covariates.
+
 Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`
-
-### Normalisation
-1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
-2. Individual_ID is chosen as random effect
-```{r iterative.normalisation, results='asis'}
-writeLines(paste('Using following covariates in the model:',
-                 paste(adjust.covars, collapse=', '),
-                 'as fixed effects and Individual_ID is chosen as random effect'))
-
-# Post adjusted formula
-random_effect <- "Individual_ID"
-formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-
-# Estimate voom weights with DREAM
-tmp = NEW.COUNTS
-tmp[is.na(tmp)] = 0
-geneExpr = DGEList(tmp)
-geneExpr = edgeR::calcNormFactors(geneExpr)
-
-VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
-                                                               formula = formula,
-                                                               data = COVARIATES,
-                                                               save.plot = TRUE)
-plot(VOOM.GENE_EXPRESSION)
-violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
-                                                         formula = formula,
-                                                         data = COVARIATES)
-plotVarPart(sortCols(violin_plot))
-
-# CQN NA counts to 0 counts for input to dream()
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-VOOM.GENE_EXPRESSION$E = tmp
-ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
-                                        formula = formula,
-                                        data = COVARIATES,
-                                        computeResiduals = TRUE)
-
-# Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-tmp = RESIDUAL.GENE_EXPRESSION
-tmp[is.na(tmp)] = 0
-residualSigCovars = runPCAandPlotCorrelations(tmp, COVARIATES,
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-```
-
-### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
-
-### Tissue Correlation
-```{r tissue_correlation}
-hasReps = COVARIATES %>%
-  dplyr::select(Individual_ID) %>% 
-  table
-
-df_dlpfc = COVARIATES %>% 
-  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
-  filter(grepl("DLPFC", Dx.Tissue)) %>% 
-  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
-df_acc = COVARIATES %>%
-  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
-  filter(grepl("ACC", Dx.Tissue)) %>% 
-  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
-
-df_merge = merge(df_dlpfc, df_acc, by="Individual_ID")
-
-corRes = sapply(res_std$ensembl_gene_id, function(ensGene){
-
-  x = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.x,drop=FALSE]
-  y = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.y,drop=FALSE]
-
-  cor(t(x)[,1], t(y)[,1], use="pairwise.complete.obs")
-})
-
-hist(corRes, xlab="Correlation", main="Correlation between ACC and DLPFC sample from the sample individual for each gene from ")
-
-pca = prcomp(t(VOOM.GENE_EXPRESSION$E))
-
-df_pca = merge(pca$x, as.data.frame(cov_dream), by.x="row.names", by.y="SampleID" )
-
-ggplot(df_pca, aes(PC1, PC2, color=Tissue, shape=Institution)) + geom_point() + theme_bw(17) + theme(aspect.ratio=1)
-```
-
-### Residual calculation
-Calculate weighted residuals and add back "Dx.Tissue" to the residuals
-```{r varsToAddBack}
-# Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(ADJUSTED.FIT$design), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-   ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
-```
-
-### Coexpression of residual expression 
-```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-tmp = RESIDUAL.GENE_EXPRESSION
-tmp[is.na(tmp)] = 0
-cr = cor(t(tmp))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-
-### Clustering residual data
-```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
-# Find principal components of expression to plot
-PC <- prcomp(tmp, scale.=T, center = T)
-
-# Plot first 4 PCs
-plotdata <- data.frame(SampleID=rownames(PC$rotation), 
-                       PC1=PC$rotation[,1], 
-                       PC2=PC$rotation[,2])
-
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
-
-p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx.Tissue, scales = 'free_y')
-p
-```
-
-```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
-# Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue')])
-COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
-
-tree = hclust(as.dist(t(tmp)))
-cols = WGCNA::labels2colors(COVARIATES.tmp);
-
-WGCNA::plotDendroAndColors(tree, 
-                           colors = cols, 
-                           dendroLabels = FALSE, 
-                           abHeight = 0.80, 
-                           main = "Sample dendrogram",
-                           groupLabels = "Dx.Tissue")
-```
-
-```{r temp1, include=F}
-dev.off()
-gc()
-```
-
-### Differential expression analysis
-Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=10, fig.width=15}
-formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueOther_ACC", "Dx.TissueControl_ACC"))
-L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueControl_ACC"))
-L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueOther_ACC"))
-L4 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueOther_STG", "Dx.TissueControl_STG"))
-L5 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueControl_STG"))
-L6 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueOther_STG"))
-L = cbind(L1, L2, L3, L4, L5, L6)
-contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
-              glue::glue_collapse(names(L2[L2 != 0]), "-"),
-              glue::glue_collapse(names(L3[L3 != 0]), "-"),
-              glue::glue_collapse(names(L4[L4 != 0]), "-"),
-              glue::glue_collapse(names(L5[L5 != 0]), "-"),
-              glue::glue_collapse(names(L6[L6 != 0]), "-"))
-# Visualize contrast matrix
-plotContrasts(L)
-
-# Fit contrast
-FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
-                                     formula = formula,
-                                     data = COVARIATES,
-                                     L = L)
-# Get differential expression
-DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
-  topTable(FIT, coef=i, number = Inf) %>%
-    rownameToFirstColumn('gene_id') %>%
-    left_join(backgroundGenes)
-}, FIT.CONTR) 
-names(DE) <- contrast
-
-DE = DE %>% 
-  rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
-                Direction = logFC/abs(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
-                Direction = as.character(Direction)) %>%
-  left_join(Ensemble2HGNC) %>%
-  left_join(gene_lengths %>%
-              rownameToFirstColumn('gene_id'))
-DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
-
-tmp = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
-
-tmp1 = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
-
-kable(full_join(tmp,tmp1))
-
-p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
-p = p + facet_grid(.~Comparison, scales = 'fixed')
-p
-```
-
-### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de, fig.height=10, fig.width=15}
-pl = list()
-pl[[1]] = ggplot(DE %>% 
-                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
-pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('red','grey','green'))
-pl[[1]] = pl[[1]] + theme_bw() %+replace% theme(legend.position = 'top')
-
-pl[[2]] = ggplot(DE %>% 
-                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = percentage_gene_gc_content, y = logFC, color = Direction)) + geom_point()
-pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gene_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('red','grey','green'))
-pl[[2]] = pl[[2]] + theme_bw() %+replace% theme(legend.position = 'top')
-
-pl[[3]] = ggplot(DE %>% 
-                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
-pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('red','grey','green'))
-pl[[3]] = pl[[3]] + theme_bw() %+replace% theme(legend.position = 'top')
-
-multiplot(plotlist = pl, cols = 3)
-```
-
-### Gene set enrichment analysis
-Using Fisher's exact test for gene set enrichment analysis
-```{r enrich.de, fig.height=16, fig.width=16}
-diffexp.genes = DE %>%
-  dlply(.(Comparison), .fun = function(x){
-    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
-    tmp = tmp[!is.na(tmp)]
-  })
-diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
-backgroundGenes = unique(DE$hgnc_symbol)
-
-# Download all related genesets from synapse
-geneset.files = synGetChildren("syn3240583", includeTypes=list("file"))$asList()
-genesets.to.test = purrr::map(geneset.files, .f = function(fileList){
-  tmp = downloadFile(fileList$id) %>%
-    dlply(.(Name), .fun = function(x){
-      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
-    })
-  }) %>%
-  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
-  do.call(c,.)
-
-# Perform enrichment analysis
-enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
-   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
-     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
-    }, x, backgroundGenes, .parallel = TRUE) %>%
-    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
-}, genesets.to.test, backgroundGenes) %>%
-  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
-  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
-  dplyr::rename(Comparison = Name, Category = .id)
-```
 
@@
 # Code
-CODE <- Folder(name = "HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
 CODE <- synStore(CODE)
@@
 synSetAnnotations(OFFSET_OBJ, annotations = all.annotations)
-
-# Store design matrix
-DM1$design %>%
-  rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'CMC_HBCC_ACC-STG_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('CMC_HBCC_ACC-STG_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
-DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                  executed = thisFile, activityDescription = activityDescription)
-all.annotations$dataSubType = 'designMatrix'
-synSetAnnotations(DM_OBJ, annotations = all.annotations)
-
-# Store residual gene expression for network analysis
-RESIDUAL.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_HBCC_ACC-STG_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_HBCC_ACC-STG_netResidualExpression.tsv', 
-                name = 'Normalised, covariates removed residual expression (for network analysis)', 
-                parentId = CODE$properties$id)
-nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-all.annotations$dataSubType = 'residualGeneExpForNetAnlz'
-synSetAnnotations(nEXP_OBJ, annotations = all.annotations)
-
-# Store differential expression results
-write.table(DE, file = 'CMC_HBCC_ACC-STG_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_HBCC_ACC-STG_DiffExpression.tsv', 
-                name = 'Differential Expression Results (Dx.Tissue)', 
-                parentId = CODE$properties$id)
-DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-all.annotations$dataSubType = 'diffExp'
-synSetAnnotations(DEXP_OBJ, annotations = all.annotations)
-
-# Store enrichment results
-write.table(enrichment.results, file = 'CMC_HBCC_ACC-STG_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
-ENRICH_OBJ = File('CMC_HBCC_ACC-STG_EnrichResults.tsv', 
-                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
-                  parentId = CODE$properties$id)
-ENRICH_OBJ = synStore(ENRICH_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                      executed = thisFile, activityDescription = activityDescription)
-all.annotations$dataSubType = 'enrichResults'
-synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
-
-# Store adjusted fit linear mixed model object as RDS
-saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_dream-output.rds')
-ADJ_OBJ = File('CMC_ACC_DLPFC_dream-output.rds',
-               name = 'Dream Output object', 
-               parentId = CODE$properties$id)
-ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
-                   executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
 ```

### 2020-03-30 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
parse out DE

@@
 ---
-title: "HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN - Sparse Model"
-author: "Kelsey Montgomery and Thanneer Perumal"
+title: "HBCC - DLPFC - FEATURECOUNTS - GRCh38 - CQN"
+author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
 date: "`r date()`"
@@
                                           parentId ="syn21435176",
-                                          folderName = 'HBCC - DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model')
+                                          folderName = 'HBCC - DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN')
 ```
@@
 ALL_USED_IDs = METADATA_QC_DLPFC_ID
-metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 14) %>%
-  dplyr::select(`Individual_ID`, Institution, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
+metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 16) %>%
+  dplyr::select(`Individual_ID`, Institution, Cohort, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
                 Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
@@
                                       'rnaSeq_report:Transcripts_Detected')) %>% 
-  dplyr::filter(Institution %in% c("NIMH-HBCC")) %>% 
   dplyr::filter(`rnaSeq_dissection:Brain_Region` %in% c("DLPFC"))
@@
 ## Download reprocessed counts (DLPFC)
-counts = 'syn21382250'
+counts = 'syn21754352'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = counts
@@
                 TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
-  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
+  dplyr::select(SampleID, Individual_ID, Institution, Cohort, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
                 RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
@@
                 IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
+  dplyr::filter(Cohort %in% "NIMH-HBCC") %>% 
+  dplyr::filter(SampleID %in% colnames(COUNT), !is.na(SampleID)) %>% 
   dplyr::mutate(Dx = forcats::fct_recode(Dx, Other = "BP", Other = "undetermined", Control = "Control", SCZ = "SCZ"))
-
-# Limit HBCC DLPFC Counts
-rownames(COUNT) = COUNT$Geneid
-COUNT <- COUNT[,colnames(COUNT) %in% METADATA$SampleID]
 
@@
 
+rownames(COUNT) <- COUNT$Geneid
 COUNT = COUNT[,indToRetain]
@@
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "sep2019.archive.ensembl.org", dataset = "hsapiens_gene_ensembl")
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL",
+                host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)",
+                dataset = "hsapiens_gene_ensembl")
 
@@
 ```
+
+#### Variables included in the model 
+Covariates were added as fixed effects iteratively in three phases if model improvement by BIC criteria was observed in the majority of genes.
+Clinical, ancestry and sample-specific technical variables were tested for model improvement first.
+Covariates related to batch effects were next tested for model improvement and finally, RNASeq alignment-specific covariates.
+
 Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`
-
-### Normalisation
-1. Dx is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
-2. Individual_ID is chosen as random effect
-```{r iterative.normalisation, results='asis'}
-writeLines(paste('Using following covariates in the model:',
-                 paste(adjust.covars, collapse=', '),
-                 'as fixed effects and Individual_ID is chosen as random effect'))
-
-# Post adjusted formula
-random_effect <- "Individual_ID"
-formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-
-# Estimate voom weights with DREAM
-tmp = NEW.COUNTS
-tmp[is.na(tmp)] = 0
-geneExpr = DGEList(tmp)
-geneExpr = edgeR::calcNormFactors(geneExpr)
-
-VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
-                                                               formula = formula,
-                                                               data = COVARIATES,
-                                                               save.plot = TRUE)
-plot(VOOM.GENE_EXPRESSION)
-violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
-                                                         formula = formula,
-                                                         data = COVARIATES)
-plotVarPart(sortCols(violin_plot))
-
-# CQN NA counts to 0 counts for input to dream()
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-VOOM.GENE_EXPRESSION$E = tmp
-ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
-                                        formula = formula,
-                                        data = COVARIATES,
-                                        computeResiduals = TRUE)
-
-# Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
-
-# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-tmp = RESIDUAL.GENE_EXPRESSION
-tmp[is.na(tmp)] = 0
-residualSigCovars = runPCAandPlotCorrelations(tmp, COVARIATES,
-                                              'adjusted design(voom-normalized)',
-                                              isKeyPlot=TRUE)
-```
-
-### Sanity check
-```{r residualSigCovars.manual, fig.width=12, fig.height=8}
-residualSigCovars[["PC_res"]][[2]]$plotData
-```
-
-### Residual calculation
-Calculate weighted residuals and add back "Dx" to the residuals
-```{r varsToAddBack}
-# Add variable of interest back to the residuals
- varsToAddIn = grep("Dx.Tissue", colnames(ADJUSTED.FIT$design), value = T)
-RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
-   ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
-```
-
-### Coexpression of residual expression 
-```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
-tmp = RESIDUAL.GENE_EXPRESSION
-tmp[is.na(tmp)] = 0
-cr = cor(t(tmp))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
-
-### Clustering residual data
-```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
-# Find principal components of expression to plot
-PC <- prcomp(tmp, scale.=T, center = T)
-
-# Plot first 4 PCs
-plotdata <- data.frame(SampleID=rownames(PC$rotation), 
-                       PC1=PC$rotation[,1], 
-                       PC2=PC$rotation[,2])
-
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
-
-p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top") + facet_grid(.~Dx, scales = 'free_y')
-p
-```
-
-```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
-# Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx')])
-COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
-
-tree = hclust(as.dist(t(tmp)))
-cols = WGCNA::labels2colors(COVARIATES.tmp);
-
-WGCNA::plotDendroAndColors(tree, 
-                           colors = cols, 
-                           dendroLabels = FALSE, 
-                           abHeight = 0.80, 
-                           main = "Sample dendrogram",
-                           groupLabels = "Dx")
-```
-
-```{r temp1, include=F}
-dev.off()
-gc()
-```
-
-### Differential expression analysis
-Genes that are differentially expressed at an FDR <= 0.05 are
-```{r diffExp, fig.height=10, fig.width=15}
-formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-
-L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueOther_DLPFC", "Dx.TissueControl_DLPFC"))
-L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueControl_DLPFC"))
-L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueOther_DLPFC"))
-L = cbind(L1, L2, L3)
-contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
-              glue::glue_collapse(names(L2[L2 != 0]), "-"),
-              glue::glue_collapse(names(L3[L3 != 0]), "-"))
-
-# Visualize contrast matrix
-plotContrasts(L)
-
-#Fit contrast
-FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
-                                     formula = formula,
-                                     data = COVARIATES,
-                                     L = L)
-# Get differential expression
-DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
-  topTable(FIT, coef=i, number = Inf) %>%
-    rownameToFirstColumn('gene_id') %>%
-    left_join(backgroundGenes)
-}, FIT.CONTR) 
- names(DE) <- contrast
-DE = DE %>% 
-  rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('Dx','',Comparison),
-                Direction = logFC/abs(logFC),
-                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
-                Direction = as.character(Direction)) %>%
-  left_join(Ensemble2HGNC) %>%
-  left_join(gene_lengths %>%
-              rownameToFirstColumn('gene_id'))
-DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
-
-tmp = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
-
-tmp1 = DE %>%
-  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison) %>%
-  group_by(Comparison) %>%
-  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
-
-kable(full_join(tmp,tmp1))
-
-p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
-p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
-p = p + facet_grid(.~Comparison, scales = 'fixed')
-p
-```
-
-### Associate differential expression results with gc content, gene length and average expression
-```{r associate.de, fig.height=10, fig.width=15}
-pl = list()
-pl[[1]] = ggplot(DE %>% 
-                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
-pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('red','grey','green'))
-pl[[1]] = pl[[1]] + theme_bw() %+replace% theme(legend.position = 'top')
-
-pl[[2]] = ggplot(DE %>% 
-                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = percentage_gene_gc_content, y = logFC, color = Direction)) + geom_point()
-pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gene_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('red','grey','green'))
-pl[[2]] = pl[[2]] + theme_bw() %+replace% theme(legend.position = 'top')
-
-pl[[3]] = ggplot(DE %>% 
-                   arrange(match(Direction, c("NONE", "UP", "DOWN"))), aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
-pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('red','grey','green'))
-pl[[3]] = pl[[3]] + theme_bw() %+replace% theme(legend.position = 'top')
-
-multiplot(plotlist = pl, cols = 3)
-```
-
-### Gene set enrichment analysis
-Using Fisher's exact test for gene set enrichment analysis
-```{r enrich.de, fig.height=16, fig.width=16}
-diffexp.genes = DE %>%
-  dlply(.(Comparison), .fun = function(x){
-    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
-    tmp = tmp[!is.na(tmp)]
-  })
-diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
-backgroundGenes = unique(DE$hgnc_symbol)
-
-# Download all related genesets from synapse
-geneset.files = synGetChildren("syn3240583", includeTypes=list("file"))$asList()
-genesets.to.test = purrr::map(geneset.files, .f = function(fileList){
-  tmp = downloadFile(fileList$id) %>%
-    dlply(.(Name), .fun = function(x){
-      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
-    })
-  }) %>%
-  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
-  do.call(c,.)
-
-# Perform enrichment analysis
-enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
-   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
-     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
-    }, x, backgroundGenes, .parallel = TRUE) %>%
-    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
-}, genesets.to.test, backgroundGenes) %>%
-  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
-  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
-  dplyr::rename(Comparison = Name, Category = .id)
-```
 
@@
 # Code
-CODE <- Folder(name = "HBCC - DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
+CODE <- Folder(name = "HBCC - DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
 CODE <- synStore(CODE)
@@
 synSetAnnotations(OFFSET_OBJ, annotations = all.annotations)
-
-# Store design matrix
-DM1$design %>%
-  rownameToFirstColumn('SampleID') %>%
-  write.table(file = 'CMC_HBCC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
-DM_OBJ = File('CMC_HBCC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
-DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                  executed = thisFile, activityDescription = activityDescription)
-all.annotations$dataSubType = 'designMatrix'
-synSetAnnotations(DM_OBJ, annotations = all.annotations)
-
-# Store residual gene expression for network analysis
-RESIDUAL.GENE_EXPRESSION %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_HBCC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_HBCC_DLPFC_netResidualExpression.tsv', 
-                name = 'Normalised, covariates removed residual expression (for network analysis)', 
-                parentId = CODE$properties$id)
-nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-all.annotations$dataSubType = 'residualGeneExpForNetAnlz'
-synSetAnnotations(nEXP_OBJ, annotations = all.annotations)
-
-# Store differential expression results
-write.table(DE, file = 'CMC_HBCC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_HBCC_DLPFC_DiffExpression.tsv', 
-                name = 'Differential Expression Results (Dx.Tissue)', 
-                parentId = CODE$properties$id)
-DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                    executed = thisFile, activityDescription = activityDescription)
-all.annotations$dataSubType = 'diffExp'
-synSetAnnotations(DEXP_OBJ, annotations = all.annotations)
-
-# Store enrichment results
-write.table(enrichment.results, file = 'CMC_HBCC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
-ENRICH_OBJ = File('CMC_HBCC_DLPFC_EnrichResults.tsv', 
-                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
-                  parentId = CODE$properties$id)
-ENRICH_OBJ = synStore(ENRICH_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                      executed = thisFile, activityDescription = activityDescription)
-all.annotations$dataSubType = 'enrichResults'
-synSetAnnotations(ENRICH_OBJ, annotations = all.annotations)
-
-# Store adjusted fit linear mixed model object as RDS
-saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_dream-output.rds')
-ADJ_OBJ = File('CMC_ACC_DLPFC_dream-output.rds',
-               name = 'Dream Output object', 
-               parentId = CODE$properties$id)
-ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
-                   executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
 ```

### 2020-03-30 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove mds plot

@@
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA MSSM/Penn/Pitt Cohort") + 
-  guides(colour = guide_legend(nrow = 2),
-         shape = guide_legend(nrow = 2),
-         size = guide_legend(nrow = 2))
-p
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
 # p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
-
-#MDS
-tmp.mds <- dist(t(tmp))
-mds <- cmdscale(tmp.mds)
-#Plot MDS
-plotdata <- data.frame(SampleID = rownames(mds),
-                       MDS1= mds[,1],
-                       MDS2= mds[,2])
-
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
-
-p <- ggplot(plotdata, aes(MDS1,MDS2))
-p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("MDS MSSM/Penn/Pitt Cohort") + 
-  guides(colour = guide_legend(nrow = 2),
-         shape = guide_legend(nrow = 2),
-         size = guide_legend(nrow = 2))
 p

### 2020-03-31 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove legacy code of printing outliers and removed samples manually

@@
 indToRetain = intersect(METADATA$SampleID, colnames(COUNT))
-removedIDs = setdiff(colnames(COUNT), METADATA$SampleID)
 
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>% 
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+  dplyr::mutate(Institution = "NIMH-HBCC") %>% 
+  dplyr::mutate(Tissue = "DLPFC")
 

### 2020-03-31 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove legacy code of printing outliers and removed samples manually

@@
 ```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
-indToRemove = c('MSSM_RNA_ACC_BP_24', 'MSSM_RNA_ACC_375', 'MSSM_RNA_ACC_170', 'MSSM_RNA_ACC_315', 'MSSM_RNA_ACC_250',
-                'MSSM_RNA_ACC_21', 'MSSM_RNA_ACC_356', 'MSSM_RNA_ACC_318', 'MSSM_RNA_ACC_222', 'MSSM_RNA_ACC_220',
-                'MSSM_RNA_ACC_379', 'MSSM_RNA_PFC_133', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_163', 'MSSM_RNA_PFC_260',
-                'MSSM_RNA_PFC_299', 'MSSM_RNA_PFC_361', 'MSSM_RNA_PFC_89', 'MSSM_RNA_PFC_319', 'MSSM_RNA_PFC_348',
-                'MSSM_RNA_PFC_354', 'MSSM_RNA_PFC_60', 'PENN_RNA_PFC_71', 'PENN_RNA_PFC_56',  "MSSM_RNA_ACC_310",
-                "MSSM_RNA_ACC_51",  "MSSM_RNA_ACC_254", "MSSM_RNA_ACC_28", "MSSM_RNA_ACC_293", "MSSM_RNA_ACC_233",
-                "MSSM_RNA_ACC_167", "MSSM_RNA_ACC_353", "MSSM_RNA_ACC_355", "MSSM_RNA_ACC_211", "PENN_RNA_ACC_122",
-                "PENN_RNA_ACC_049", "PENN_RNA_ACC_052", "PENN_RNA_ACC_068", "PENN_RNA_ACC_070", "PITT_RNA_ACC_10020",
-                "PITT_RNA_ACC_1240", "PITT_RNA_ACC_1284", "PITT_RNA_ACC_700", "PITT_RNA_ACC_857", "PITT_RNA_ACC_BP_1589",
-                "MSSM_RNA_PFC_230", "MSSM_RNA_PFC_304", "MSSM_RNA_PFC_309", "MSSM_RNA_PFC_329", "MSSM_RNA_PFC_82",
-                "PENN_RNA_PFC_5", "PENN_RNA_PFC_58", "PENN_RNA_PFC_99", "PITT_RNA_PFC_1284")
-
-indToRemove = c(indToRemove, c("MSSM_RNA_ACC_224", "MSSM_RNA_ACC_240", "MSSM_RNA_ACC_BP_11", "MSSM_RNA_ACC_42", "MSSM_RNA_ACC_BP_15", "PENN_RNA_ACC_003", "MSSM_RNA_ACC_374"))
-
-indToRemove = c(indToRemove, "PITT_RNA_BP_PFC_1770")
-
 # Find principal components of expression to plot
@@
 Based on the expression pattern following samples were tagged as outliers: `r paste(outliers, collapse = ', ')`
-
-Based on prior evidence, investigators elected to remove: `r setdiff(indToRemove, outliers)[setdiff(indToRemove,outliers) %in% colnames(counts)]`
 

### 2020-03-31 cmc-rnaseq-de/Differential-Expression.Rmd
error with inputs
fix factor vars and df to matrix

@@
 CQN.GENE_EXPRESSION = downloadFile(CQN_ID) %>% 
-  tibble::column_to_rownames(var = "ensembl_gene_id")
+  tibble::column_to_rownames(var = "ensembl_gene_id") %>% 
+  data.matrix()
 
@@
 ALL_USED_IDs = c(ALL_USED_IDs, COVAR_ID)
-COVARIATES = downloadFile(COVAR_ID)
+COVARIATES = downloadFile(COVAR_ID) %>% 
+  tibble::column_to_rownames(var = "SampleID")
+
+FactorCovariates <- c("Individual_ID", "Reported_Gender", "Dx.Tissue")
+COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
 

### 2020-03-31 cmc-rnaseq-de/Differential-Expression.Rmd
set factor and numeric class on covars

@@
 
-FactorCovariates <- c("Individual_ID", "Reported_Gender", "Dx.Tissue")
+# Set variable class
+FactorCovariates <- c("Individual_ID", "Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
 COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
+
+ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
+                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads","AlignmentRate", 
+                    "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")
+COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], function(x){
+  x = as.numeric(as.character(gsub('[\\,\\%]','',x)))
+})
 

### 2020-04-01 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd
remove CQN normalization from salmon model sel

--- 2020-03-31 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
 ---
-title: "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Model Selection"
+title: "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - SALMON - GRCh38 - Model Selection"
 author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
@@
 
-knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd",
+knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd",
                                  parentId ="syn21435176",
-                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN")
+                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - SALMON - GRCh38")
 ```
@@
 library(edgeR)
-library(cqn)
 library(lme4)
@@
 activityName = 'Covariate adjustments';
-activityDescription = 'Covariate analysis of GRCh38 feature counts with CQN normalisation (ACC and DLPFC)';
+activityDescription = 'Covariate analysis of GRCh38 salmon counts (ACC and DLPFC)';
 
-thisFileName <- 'MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd'
+thisFileName <- 'MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd'
 
@@
 # Download reprocessed counts
-counts = 'syn21754352'
+counts = 'syn21869686'
 ALL_USED_IDs = counts
-counts = downloadFile(counts) %>% data.frame()
-counts$Chr <- NULL
-counts$Start <- NULL
-counts$End <- NULL
-counts$Strand <- NULL
-
-# Parse gene lengths
-gene_lengths = counts[,c("Geneid", "Length")]
-counts$Length <- NULL
-gene_lengths <- dplyr::rename(gene_lengths, gene_id = Geneid)
+counts = readRDS(synGet(counts)$path)
+counts = counts$counts
 
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = metadata
-metadata = downloadFile_version(metadata, version = 16)  %>% 
+metadata = downloadFile_version(metadata, version = 18)  %>% 
   dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, 
@@
 ### Filter genes
-Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
+Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Remove genes with percentage GC content
 ```{r cpmnormalisation}
@@
   unique() %>%
-  intersect(GENE.GC.CONT$gene_id[!is.na(GENE.GC.CONT$percentage_gene_gc_content)]) %>%
-  intersect(gene_lengths$gene_id[!is.na(gene_lengths$Length)])
+  intersect(GENE.GC.CONT$gene_id[!is.na(GENE.GC.CONT$percentage_gene_gc_content)])
 
@@
 ### Library Normalisation
-Initial library normalisation usign cqn
-```{r cqn}
-# Compute offset for gene length and gc content
-gene_lengths = tibble::column_to_rownames(gene_lengths, var = 'gene_id') %>% as.data.frame()
-CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
-                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gene_gc_content'],
-                          lengths = gene_lengths[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
-                          lengthMethod = "smooth", 
-                          verbose = FALSE)
-CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset
-```
+Salmon utilizes sample-specific bias models for transcriptome-wide abundance estimation.
 
@@
 ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$E))
+cr = cor(t(NEW.COUNTS))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
@@
 # Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)
+PC <- prcomp(NEW.COUNTS, scale.=T, center = T)
 
@@
 
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
+tree = hclust(as.dist(t(NEW.COUNTS)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 
-### Distribution of samples (log cpm)
-```{r lcpm.dist, cache=FALSE}
-# Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$E) %>%
-  rownameToFirstColumn('gene_id') %>%
-  tidyr::gather(SampleID, logCPM, -gene_id) %>%
-  left_join(COVARIATES %>%
-              rownameToFirstColumn('SampleID'))
-
-p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
-p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx.Tissue, scale = 'free')
-p
-```
-
 ### Significant Covariates
@@
 # Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
+preAdjustedSigCovars = runPCAandPlotCorrelations(NEW.COUNTS, 
                                                  COVARIATES,
@@
 # Function to fit a mixed effect model and return BIC
-fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){
+fitMixedEffectModel <- function(varsToFit, groupingVar, cnt, covars){
 
   # Get design matrix
-  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
+  designMat = getDesignMatrix(covars[colnames(cnt), varsToFit, drop=F], Intercept = F)
   designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]
@@
 
-  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
+  modelBIC = foreach::foreach(i = 1:dim(cnt)[1],
                               .combine = rbind,
@@
                                 # Perform weighted mixed linear models for every gene
-                                expr = data.frame(Gx = lcpm[i,])
+                                expr = data.frame(Gx = cnt[i,])
                                 data = cbind(expr,
@@
                                 })
-                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
+                              return(data.frame(ensembl_gene_id = rownames(cnt)[i],
                                                 BIC = BIC))
@@
 # Function to select covariates in a step-wise regression fashion
-selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){
+selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, cnt, baseModelBIC){
 
@@
                                                            covars[colnames(cnt),'Individual_ID'], 
-                                                           lcpm, 
                                                            cnt, 
@@
                                    COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
-                                   CQN.GENE_EXPRESSION$E, 
                                    NEW.COUNTS, 
@@
 vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
+                                     COVARIATES,
                                      NEW.COUNTS, baseModelBIC)
@@
 vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
+                                     COVARIATES,
                                      NEW.COUNTS, baseModelBIC)
@@
 vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
+                                     COVARIATES,
                                      NEW.COUNTS, baseModelBIC)
@@
 
-  transcriptNormalizationMethod	= 'CQN',
-  transcriptQuantificationMethod = 'featureCounts',
+  transcriptNormalizationMethod	= 'salmon',
+  transcriptQuantificationMethod = 'salmon',
   referenceSet = 'GRCh38'
@@
 # Code
-CODE <- Folder(name = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
+CODE <- Folder(name = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - SALMON - GRCh38", parentId = parentId)
 CODE <- synStore(CODE)
@@
 synSetAnnotations(COUNT_OBJ, annotations = all.annotations)
-
-# Store logCPM
-CQN.GENE_EXPRESSION$E %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
-LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                      executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(LCOUNT_OBJ, annotations = all.annotations)
-
-# Store cqn offsets
-CQN.GENE_EXPRESSION$offset %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
-OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                      executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(OFFSET_OBJ, annotations = all.annotations)
 ```
@@
 |  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
-|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
-|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
 

### 2020-04-01 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd
modify inputs for salmon transcripts

@@
 
-rownames(counts) <- counts$Geneid
 counts = counts[,indToRetain]
@@
 ## Get GC content from biomart
-backgroundGenes = data.frame(gene_id = rownames(counts)) %>%
-  dplyr::mutate(id = gene_id) %>%
-  tidyr::separate(id, c('ensembl_gene_id','position'), sep = '\\.')
+backgroundGenes = data.frame(transcript_id = rownames(counts)) %>%
+  dplyr::mutate(id = transcript_id) %>%
+  tidyr::separate(id, c('ensembl_transcript_id','position'), sep = '\\.')
 
@@
 # Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
-                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
+Ensemble2HGNC <- getBM(attributes = c("ensembl_transcript_id", "hgnc_trans_name", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
+                       filters = "ensembl_transcript_id", values = backgroundGenes$ensembl_transcript_id,
                        mart = mart)
@@
   dplyr::left_join(backgroundGenes) %>% 
-  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
+  dplyr::select(transcript_id, percentage_gene_gc_content, chromosome_name) %>%
   unique
-rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
+rownames(GENE.GC.CONT) = GENE.GC.CONT$transcript_id
 ```
@@
 ```{r Reported_Gender_plots}
-counts$gene_id = rownames(counts) 
+counts$transcript_id = rownames(counts) 
 REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
@@
   filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
-  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
+  tidyr::gather(key = item, value = value, -c(transcript_id, chromosome_name)) %>%
   dplyr::mutate(value = log(value)) %>%
@@
 FILT <- REPORTED.GENDER.COUNTS %>% 
-  filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>% 
+  filter(transcript_id == "ENSG00000229807.11" | transcript_id == "ENSG00000183878.15") %>% 
   dplyr::select(-one_of("chromosome_name")) %>% 
-  tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
+  tidyr::spread(key = transcript_id, value = `counts(log)`) %>% 
   mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>% 
@@
 ```{r cpmnormalisation}
-counts$gene_id = NULL
+counts$transcript_id = NULL
 
@@
   unique() %>%
-  intersect(GENE.GC.CONT$gene_id[!is.na(GENE.GC.CONT$percentage_gene_gc_content)])
+  intersect(GENE.GC.CONT$transcript_id[!is.na(GENE.GC.CONT$percentage_gene_gc_content)])
 
@@
 pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
-  dplyr::rename(gene_id = genes) %>%
+  dplyr::rename(transcript_id = genes) %>%
   left_join(backgroundGenes) %>%
@@
 tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
+  rownameToFirstColumn('ensembl_transcript_id') %>%
+  tidyr::gather(SampleID, logCPM, -ensembl_transcript_id) %>%
   left_join(COVARIATES %>%
@@
                                 })
-                              return(data.frame(ensembl_gene_id = rownames(cnt)[i],
+                              return(data.frame(ensembl_transcript_id = rownames(cnt)[i],
                                                 BIC = BIC))
@@
     nChange = allModelBIC %>%
-      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
+      tidyr::gather(ModelName, Value, -ensembl_transcript_id) %>%
       dplyr::left_join(baseModelBIC) %>%
@@
       baseModelBIC = allModelBIC %>%
-        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
+        dplyr::select(ensembl_transcript_id, one_of(tmp$ModelName))
       colnames(baseModelBIC)[2] = 'baseModel'
@@
 PROCESSED_COUNTS$filteredExprMatrix$counts %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
+  rownameToFirstColumn('ensembl_transcript_id') %>%
   write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)

### 2020-04-01 cmc-rnaseq-de/Differential-Expression.Rmd
fix provenance MPP cohort

@@
 title: "Differential Expression"
-author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
+author: "Kelsey Montgomery, Gabriel Hoffman, Thaneer Perumal"
 date: "`r date()`"
@@
 library(limma)
+library(psych)
 library(lme4)
 library(biomaRt)
-library(variancePartition)
+library(variancePartition) #1.17.9
+library(edgeR)
 library(synapser)
@@
 
-library(doParallel)
 library(foreach)
 library(future)
+library(BiocParallel)
 
-cl = makeCluster((availableCores()-2)/2)
+# See dream documentation to customize parallel processing       http://bioconductor.org/packages/release/bioc/vignettes/variancePartition/inst/doc/dream.html#parallel-processing
+param = SnowParam((availableCores()-2)/2, "SOCK", progressbar=TRUE)
 
-registerDoParallel(cl)
+register(param)
 
@@
 # Find principal components of expression to plot
-PC <- prcomp(tmp, scale.=T, center = T)
+PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)
 
@@
 
-```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
+```{r decompose.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
 # Eucledian tree based analysis
@@
 
-tree = hclust(as.dist(t(tmp)))
+tree = hclust(as.dist(t(RESIDUAL.GENE_EXPRESSION)))
 cols = WGCNA::labels2colors(COVARIATES.tmp);
@@
 
+# Get background genes 
+backgroundGenes = data.frame(gene_id = rownames(NEW.COUNTS)) %>%
+  dplyr::mutate(id = gene_id) %>%
+  tidyr::separate(id, c('ensembl_gene_id','position'), sep = '\\.')
+
+# Get gene lengths
+counts = 'syn21867938'
+ALL_USED_IDs = counts
+counts = downloadFile(counts) %>% data.frame()
+gene_lengths = counts[,c("Geneid", "Length")] %>% 
+  dplyr::rename(gene_id = Geneid)
+rm(counts)
+
+# Define biomart object
+mart <- biomaRt::useMart(biomart = "ENSEMBL_MART_ENSEMBL",
+                host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)
+                dataset = "hsapiens_gene_ensembl")
+
+# Query biomart
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", 
+                                      "percentage_gene_gc_content", "gene_biotype", 
+                                      "chromosome_name"),
+                       filters = "ensembl_gene_id", 
+                       values = backgroundGenes$ensembl_gene_id,
+                       mart = mart)
+
 # Get differential expression
@@
   left_join(Ensemble2HGNC) %>%
-  left_join(gene_lengths %>% rownameToFirstColumn('gene_id'))
+  left_join(gene_lengths)
 DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'
@@
 pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[1]] = pl[[1]] + facet_grid(Tissue~.) + scale_color_manual(values = c('green','grey','red'))
 pl[[1]] = pl[[1]] + theme(legend.position = 'top')
@@
 pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gene_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[2]] = pl[[2]] + facet_grid(Tissue~.) + scale_color_manual(values = c('green','grey','red'))
 pl[[2]] = pl[[2]] + theme(legend.position = 'top')
@@
 pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[3]] = pl[[3]] + facet_grid(Tissue~.) + scale_color_manual(values = c('green','grey','red'))
 pl[[3]] = pl[[3]] + theme(legend.position = 'top')

### 2020-04-01 cmc-rnaseq-de/reproduce_dream.R
remove test files

@@ -1,44 +1,0 @@
-devtools::install_github("GabrielHoffman/variancePartition@3287d7b")
-library(variancePartition)
-library(tibble)
-library(synapser)
-synLogin()
-param = BiocParallel::SnowParam(4, "SOCK", progressbar=TRUE)
-BiocParallel::register(param)
-
-COVARIATES <- readr::read_csv(synGet("syn21488871")$path) %>% 
-  tibble::column_to_rownames(var = "SampleID")
-
-NEW.COUNTS <- readRDS(synGet("syn21535143")$path)
-
-CQN.GENE_EXPRESSION <- readRDS(synGet("syn21535146")$path)
-
-# Post adjusted formula
-random_effect <- "Individual_ID"
-adjust.covars <- c("Dx.Tissue", "RIN")
-formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
-
-tmp = NEW.COUNTS
-tmp[is.na(tmp)] = 0
-geneExpr = DGEList(tmp)
-geneExpr = edgeR::calcNormFactors(geneExpr)
-
-VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
-                                                               formula = formula,
-                                                               data = COVARIATES)
-
-# Fit linear model using new weights and new design
-# CQN NA counts to 0 counts for input to dream()
-tmp = CQN.GENE_EXPRESSION$E
-tmp[is.na(tmp)] = 0
-VOOM.GENE_EXPRESSION$E = tmp
-ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
-                                        formula = formula,
-                                        data = COVARIATES,
-                                        computeResiduals = TRUE)
-
-RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
-
-
-

### 2020-04-01 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
update feature counts input

@@
 ALL_USED_IDs = METADATA_QC_ACC_ID
-metadata = downloadFile_version(METADATA_QC_ACC_ID, version = 16) %>%
+metadata = downloadFile_version(METADATA_QC_ACC_ID, version = 18) %>%
   dplyr::select(`Individual_ID`, Institution, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
@@
 ## Download reprocessed counts (ACC)
-counts = 'syn21754352'
+counts = 'syn21867762'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = counts

### 2020-04-01 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
update feature counts input

@@
 title: "HBCC - ACC - STG - DREAM - FEATURECOUNTS - GRCh38 - CQN"
-author: "Kelsey Montgomery, Gabriel Hoffman and Thanneer Perumal"
+author: "Kelsey Montgomery, Gabriel Hoffman and Thaneer Perumal"
 date: "`r date()`"
@@
 ALL_USED_IDs = METADATA_QC_ID
-metadata = downloadFile_version(METADATA_QC_ID, version = 16) %>%
+metadata = downloadFile_version(METADATA_QC_ID, version = 18) %>%
   dplyr::select(`Individual_ID`, Institution, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
@@
 ## Download reprocessed counts (ACC and STG)
-counts = 'syn21754352'
+counts = 'syn21867866'
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = counts

### 2020-04-01 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove mds plot

@@
 # p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
-
-#MDS
-tmp.mds <- dist(t(tmp))
-mds <- cmdscale(tmp.mds)
-#Plot MDS
-plotdata <- data.frame(SampleID = rownames(mds),
-                       MDS1= mds[,1],
-                       MDS2= mds[,2])
-
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
-
-p <- ggplot(plotdata, aes(MDS1,MDS2))
-p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("HBCC Cohort") + 
-  guides(colour = guide_legend(nrow = 2),
-         shape = guide_legend(nrow = 2),
-         size = guide_legend(nrow = 2))
-p
 ```

### 2020-04-01 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove mds plot

@@
 # p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
-
-#MDS
-tmp.mds <- dist(t(tmp))
-mds <- cmdscale(tmp.mds)
-#Plot MDS
-plotdata <- data.frame(SampleID = rownames(mds),
-                       MDS1= mds[,1],
-                       MDS2= mds[,2])
-
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
-
-p <- ggplot(plotdata, aes(MDS1,MDS2))
-p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("MDS MSSM/Penn/Pitt Cohort") + 
-  guides(colour = guide_legend(nrow = 2),
-         shape = guide_legend(nrow = 2),
-         size = guide_legend(nrow = 2))
-p
 ```

### 2020-04-02 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
debug debug

@@
 
-```{r synapse.parameters, include=FALSE, cache=TRUE}
+```{r synapse.parameters, include=FALSE, cache = FALSE}
 parentId = 'syn21435176';
@@
 metadata = downloadFile_version(METADATA_QC_ID, version = 18) %>%
-  dplyr::select(`Individual_ID`, Institution, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
+  dplyr::select(`Individual_ID`, Institution, Cohort, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
                 Sample_RNA_ID, one_of('rnaSeq_isolation:RIN',
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = counts
-COUNT = downloadFile(counts) %>% data.frame()
+COUNT = downloadFile(counts)
 COUNT$Chr <- NULL
@@
                 TotalReads = `rnaSeq_report:Total_Reads`, AlignmentRate = `rnaSeq_report:Percent_Aligned`, TranscriptsDetected = `rnaSeq_report:Transcripts_Detected`) %>%
-  dplyr::select(SampleID, Individual_ID, Institution, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
+  dplyr::select(SampleID, Individual_ID, Tissue, Institution, Reported_Gender,Sex, Ethnicity, ageOfDeath, PMI, Dx, 
                 RIN, ReportExclude, GenotypeExclude, EV.1, EV.2, EV.3, EV.4, EV.5,
@@
   dplyr::mutate(Dx = forcats::fct_recode(Dx, Other = "MDD", Other = "BP", Control = "Control", SCZ = "SCZ"))
-
-# Limit HBCC ACC and STG Counts
-rownames(COUNT) = COUNT$Geneid
-COUNT <- COUNT[,colnames(COUNT) %in% METADATA$SampleID]
 

### 2020-04-02 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
debug debug
fix pca plots facets and legend

@@
   dplyr::mutate(label = SampleID) %>% 
-  dplyr::mutate(label = ifelse((label %in% outliers), label, NA))
+  dplyr::mutate(label = ifelse((label %in% outliers), label, NA)) %>% 
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
-p <- p + theme_bw() %+replace% theme(legend.position="right")
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
 p <- p + geom_text(aes(label= label), size=4, hjust=0)
@@
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>% 
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx.Tissue, size=ageOfDeath))
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
 p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA MSSM/Penn/Pitt Cohort") + 
@@
          shape = guide_legend(nrow = 2),
-         size = guide_legend(nrow = 2))
+         size = guide_legend(nrow = 2)) +
+  facet_grid(Tissue~.)
 p

### 2020-04-02 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix pca plots facets and legend

@@
   dplyr::mutate(label = ifelse((label %in% outliers), label, NA)) %>% 
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>% 
+  dplyr::mutate(Institution = "NIMH-HBCC")
 
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>% 
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>% 
+  dplyr::mutate(Institution = "NIMH-HBCC")
 

### 2020-04-02 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix pca plots facets and legend

@@
 ALL_USED_IDs = METADATA_QC_DLPFC_ID
-metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 16) %>%
+metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 18) %>%
   dplyr::select(`Individual_ID`, Institution, Cohort, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
@@
   dplyr::mutate(label = SampleID) %>% 
-  dplyr::mutate(label = ifelse((label %in% outliers), label, NA))
+  dplyr::mutate(label = ifelse((label %in% outliers), label, NA)) %>% 
+  dplyr::mutate(Institution = "NIMH-HBCC") %>% 
+  dplyr::mutate(Tissue = "DLPFC")
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
+p <- p + geom_point(aes(shape= Tissue, color = Institution, size=ageOfDeath))
 p <- p + theme_bw() %+replace% theme(legend.position="right")

### 2020-04-02 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd

@@
 # Github link
-thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='add-dream')
+thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='master')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))
@@
 ```
-
 #### Variables included in the model 

### 2020-04-02 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd

@@
 ALL_USED_IDs = METADATA_QC_ID
+
 metadata = downloadFile_version(METADATA_QC_ID, version = 18) %>%
@@
 p
+
 # Plot abberent distribution of logcpm counts

### 2020-04-02 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd

@@
 # Download reprocessed counts
-counts = 'syn21754352'
+counts = 'syn21867938'
 ALL_USED_IDs = counts
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = metadata
-metadata = downloadFile_version(metadata, version = 16)  %>% 
+metadata = downloadFile_version(metadata, version = 17)  %>% 
   dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, 

### 2020-04-03 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
ggplot legend

@@
 ALL_USED_IDs = METADATA_QC_ACC_ID
-metadata = downloadFile_version(METADATA_QC_ACC_ID, version = 18) %>%
+metadata = downloadFile_version(METADATA_QC_ACC_ID, version = 19) %>%
   dplyr::select(`Individual_ID`, Institution, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 
@@
   dplyr::mutate(label = SampleID) %>% 
-  dplyr::mutate(label = ifelse((label %in% outliers), label, NA))
+  dplyr::mutate(label = ifelse((label %in% outliers), label, NA)) %>% 
+  dplyr::mutate(Tissue = "DLPFC") %>% 
+  dplyr::mutate(Institution = "NIMH-HBCC")
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
+p <- p + geom_point(aes(shape= Tissue, color = Institution, size = ageOfDeath))
 p <- p + theme_bw() %+replace% theme(legend.position="right")
@@
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID'))
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  dplyr::mutate(Tissue = "DLPFC") %>% 
+  dplyr::mutate(Institution = "NIMH-HBCC")
 
 p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(shape=Dx, size=ageOfDeath))
+p <- p + geom_point(aes(shape= Tissue, color = Institution, size = ageOfDeath))
 p <- p + theme_bw() %+replace% theme(legend.position="top", plot.title = element_text(hjust=0.5, size = 15)) + ggtitle ("PCA HBCC Cohort") + 

### 2020-04-03 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix forcats clinical var normalization

@@
                 IntragenicRate, IntergenicRate, IntronicRate, ExpProfEfficiency, rRNARate, TranscriptsDetected) %>%
-  dplyr::mutate(Dx = forcats::fct_recode(Dx, Other = "BP", Control = "Control", SCZ = "SCZ"))
+  dplyr::mutate(Dx = forcats::fct_recode(Dx, Other = "BP", Other = "MDD", Control = "Control", SCZ = "SCZ"))
 
@@
 
-  tissue	= 'dorsolateral prefrontal cortex, anterior cingulate cortex', 
+  tissue	= 'anterior cingulate cortex', 
   study = 'CMC',

### 2020-04-04 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd
transcript ids XIST UTY

@@
 # Github link
-thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='add-dream')
+thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='compute-salmon')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))
@@
 ALL_USED_IDs[length(ALL_USED_IDs)+1] = metadata
-metadata = downloadFile_version(metadata, version = 18)  %>% 
+metadata = downloadFile_version(metadata, version = 19)  %>% 
   dplyr::select(Individual_ID, Institution, Reported_Gender, Sex, 
@@
 ##XIST and UTY expression 
-#ENSG00000229807.11 and ENSG00000183878.15 
+#ENST00000416330.2 and ENST00000329134.9 
 FILT <- REPORTED.GENDER.COUNTS %>% 
-  filter(transcript_id == "ENSG00000229807.11" | transcript_id == "ENSG00000183878.15") %>% 
+  filter(transcript_id == "ENST00000416330.2" | transcript_id == "ENST00000329134.9") %>% 
   dplyr::select(-one_of("chromosome_name")) %>% 
   tidyr::spread(key = transcript_id, value = `counts(log)`) %>% 
-  mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>% 
-  mutate(UTY = as.numeric(`ENSG00000183878.15`))
+  mutate(XIST = as.numeric(`ENST00000416330.2`)) %>% 
+  mutate(UTY = as.numeric(`ENST00000329134.9`))
 
@@
 ```
+
 ### Covariates clustering

### 2020-04-05 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd
provenance for salmon counts

@@
 title: "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - SALMON - GRCh38 - Model Selection"
-author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
+author: "Kelsey Montgomery, Gabriel Hoffman, Thaneer Perumal"
 date: "`r date()`"

### 2020-04-05 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd
counts as df

@@
 
-```{r synapse.parameters, include=FALSE, cache=TRUE}
+```{r synapse.parameters, include=FALSE, cache=FALSE}
 parentId = 'syn21435176';
@@
 counts = readRDS(synGet(counts)$path)
-counts = counts$counts
+counts = counts$counts %>% 
+  as.data.frame()
 

### 2020-04-05 cmc-rnaseq-de/Differential-Expression.Rmd
ACC STG HBCC Dx.Tissue

@@
 knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/Differential-Expression.Rmd",
-                                 parentId ="syn21824667",
+                                 parentId ="syn21887806",
                                  folderName = "Differential Expression - Dx.Tissue")
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = 'syn21824667';
+parentId = 'syn21887806';
 activityName = 'Differential Expression';
@@
 # Github link
-thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='add-dream')
+thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='differential-expression')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))
@@
 # Download CQN counts
-CQN_ID = "syn21825509"
+CQN_ID = "syn21888972"
 ALL_USED_IDs = c(CQN_ID)
@@
 # Download filtered counts
-COUNTS_ID = "syn21825508"
+COUNTS_ID = "syn21888969"
 ALL_USED_IDs = c(ALL_USED_IDs, COUNTS_ID)
@@
 # Get covariates
-COVAR_ID = "syn21825507"
+COVAR_ID = "syn21888968"
 ALL_USED_IDs = c(ALL_USED_IDs, COVAR_ID)
@@
 # Set covariates to adjust for from model selection
-adjust.covars <- c("Dx.Tissue", "RIN", "IntronicRate")
+adjust.covars <- c("Dx.Tissue", "RIN", "IntragenicRate")
 ```
@@
 
-df_dlpfc = rownameToFirstColumn(COVARIATES, "SampleID") %>% 
+df_stg = rownameToFirstColumn(COVARIATES, "SampleID") %>% 
   filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
-  filter(grepl("DLPFC", Dx.Tissue)) %>% 
+  filter(grepl("STG", Dx.Tissue)) %>% 
   dplyr::select(SampleID, Individual_ID, Dx.Tissue)
@@
 
-df_merge = merge(df_dlpfc, df_acc, by="Individual_ID")
+df_merge = merge(df_stg, df_acc, by="Individual_ID")
 
@@
 
-hist(corRes, xlab="Correlation", main="Correlation between ACC and DLPFC sample from the sample individual for each gene from ")
+hist(corRes, xlab="Correlation", main="Correlation between ACC and STG sample from the sample individual for each gene")
 
@@
                                    data = COVARIATES, 
-                                   coefficient = c("Dx.TissueOther_DLPFC", "Dx.TissueControl_DLPFC"))
+                                   coefficient = c("Dx.TissueOther_STG", "Dx.TissueControl_STG"))
 L5 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
@@
                                    data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueControl_DLPFC"))
+                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueControl_STG"))
 L6 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
@@
                                    data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_DLPFC", "Dx.TissueOther_DLPFC"))
+                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueOther_STG"))
 
@@
 # Get gene lengths
-counts = 'syn21867938'
+counts = 'syn21867866'
 ALL_USED_IDs = counts
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_Dx.Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue_netResidualExpression.tsv', 
+  write.table(file = 'CMC_ACC_STG_Dx.Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+nEXP_OBJ = File('CMC_ACC_STG_Dx.Tissue_netResidualExpression.tsv', 
                 name = 'Dx.Tissue Normalised, covariates removed residual expression (for network analysis)', 
@@
 # Store differential expression results
-write.table(DE, file = 'CMC_ACC_DLPFC_Dx.Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue_DiffExpression.tsv', 
+write.table(DE, file = 'CMC_ACC_STG_Dx.Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('CMC_ACC_STG_Dx.Tissue_DiffExpression.tsv', 
                 name = 'Differential Expression Results (Dx.Tissue)', 
@@
 # Store adjusted fit linear mixed model object as RDS - Dx.Tissue
-saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_DLPFC_Dx.Tissue_dream-output.rds')
-ADJ_OBJ = File('CMC_ACC_DLPFC_Dx.Tissue_dream-output.rds',
+saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_STG_Dx.Tissue_dream-output.rds')
+ADJ_OBJ = File('CMC_ACC_STG_Dx.Tissue_dream-output.rds',
                name = 'Dream Output object - Dx.Tissue', 

### 2020-04-05 cmc-rnaseq-de/Differential-Expression.Rmd
header = T in fread

@@
 downloadFile <- function(id){
-  fread(synGet(id)$path, data.table = F)
+  fread(synGet(id)$path, data.table = F, header = T)
 }

### 2020-04-05 cmc-rnaseq-de/Differential-Expression.Rmd
remove missing factor covariates for this cohort

@@
 # Set variable class
-FactorCovariates <- c("Individual_ID", "Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue", "RibozeroBatch", "FlowcellBatch")
+FactorCovariates <- c("Individual_ID", "Reported_Gender", "Dx.Tissue")
 COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)

### 2020-04-05 cmc-rnaseq-de/HBCC_ACC-Nirmala_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
rm zero variance ENSG00000210082.2
test alignment rate in HBCC cohort model selection

@@
 ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
-                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
+                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", "AlignmentRate",
                     "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

### 2020-04-05 cmc-rnaseq-de/HBCC_ACC-STG_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix outlier plot typo

@@
   dplyr::mutate(label = ifelse((label %in% outliers), label, NA)) %>% 
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>% 
   dplyr::mutate(Institution = "NIMH-HBCC")

### 2020-04-05 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd
lazy cache for long vectors
compute cor on half of genes to reduce computational load

@@
 Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Remove genes with percentage GC content
-```{r cpmnormalisation}
+```{r cpmnormalisation, cache.lazy = FALSE}
 counts$transcript_id = NULL
@@
 ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(NEW.COUNTS))
+# Reduce computational load and compute correlation on half of genes
+random_sampling <- sample(dim(NEW.COUNTS)[1], size = dim(NEW.COUNTS)[1]/2 , replace = F)
+tmp <- NEW.COUNTS[random_sampling,]
+cr = cor(t(tmp))
 hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')

### 2020-04-06 cmc-rnaseq-de/Differential-Expression.Rmd
ACC STG HBCC Dx.Tissue - specific vars

@@
 # Set variable class
-FactorCovariates <- c("Individual_ID", "Reported_Gender", "Dx.Tissue")
+FactorCovariates <- c("Individual_ID", "Reported_Gender", "Dx.Tissue", "IsolationBatch")
 COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
@@
                by.y="SampleID" ) %>% 
-  tidyr::separate(Dx.Tissue, into = c("Dx", "Tissue"), sep = "_")
+  tidyr::separate(Dx.Tissue, into = c("Dx", "Tissue"), sep = "_") %>% 
+  dplyr::mutate(Institution = "NIMH-HBCC")
 
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>% 
+  dplyr::mutate(Institution = "NIMH-HBCC")
 
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Institution')])
+COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue'), drop = F])
 
@@
                            main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Institution')]))
+                           groupLabels = colnames(COVARIATES.tmp))
 ```

### 2020-04-07 cmc-rnaseq-de/Differential-Expression_Dx.Rmd
DE contrast for HBCC DLPFC

--- 2020-04-06 cmc-rnaseq-de/Differential-Expression.Rmd
@@
 knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/Differential-Expression.Rmd",
-                                 parentId ="syn21887806",
-                                 folderName = "Differential Expression - Dx.Tissue")
+                                 parentId ="syn21859863",
+                                 folderName = "Differential Expression - Dx")
 ```
@@
 ```{r synapse.parameters, include=FALSE, cache=TRUE}
-parentId = 'syn21887806';
+parentId = 'syn21859863';
 activityName = 'Differential Expression';
-activityDescription = 'Differential Expression - Dx.Tissue';
+activityDescription = 'Differential Expression - Dx';
 
-thisFileName <- 'Differential-Expression.Rmd'
+thisFileName <- 'Differential-Expression_Dx.Rmd'
 
@@
 # Download CQN counts
-CQN_ID = "syn21888972"
+CQN_ID = "syn21862457"
 ALL_USED_IDs = c(CQN_ID)
@@
 # Download filtered counts
-COUNTS_ID = "syn21888969"
+COUNTS_ID = "syn21862455"
 ALL_USED_IDs = c(ALL_USED_IDs, COUNTS_ID)
@@
 # Get covariates
-COVAR_ID = "syn21888968"
+COVAR_ID = "syn21862454"
 ALL_USED_IDs = c(ALL_USED_IDs, COVAR_ID)
@@
 # Set variable class
-FactorCovariates <- c("Individual_ID", "Reported_Gender", "Dx.Tissue", "IsolationBatch")
+FactorCovariates <- c("Individual_ID", "Reported_Gender", "Dx", "IsolationBatch")
 COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
@@
 # Set covariates to adjust for from model selection
-adjust.covars <- c("Dx.Tissue", "RIN", "IntragenicRate")
+adjust.covars <- c("Dx", "RIN", "rRNARate", "IntergenicRate")
 ```
 
-### Normalisation - Dx.Tissue
-1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
+### Normalisation - Dx
+1. Dx is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
 2. Individual_ID is chosen as random effect
@@
 
-### Tissue Correlation
-```{r tissue_correlation}
-hasReps = rownameToFirstColumn(COVARIATES, "SampleID") %>%
-  dplyr::select(Individual_ID) %>% 
-  table
-
-df_stg = rownameToFirstColumn(COVARIATES, "SampleID") %>% 
-  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
-  filter(grepl("STG", Dx.Tissue)) %>% 
-  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
-df_acc = rownameToFirstColumn(COVARIATES, "SampleID") %>%
-  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
-  filter(grepl("ACC", Dx.Tissue)) %>% 
-  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
-
-df_merge = merge(df_stg, df_acc, by="Individual_ID")
-
-res_std <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id")
-
-corRes = sapply(res_std$ensembl_gene_id, function(ensGene){
-
-  x = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.x,drop=FALSE]
-  y = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.y,drop=FALSE]
-
-  cor(t(x)[,1], t(y)[,1], use="pairwise.complete.obs")
-})
-
-hist(corRes, xlab="Correlation", main="Correlation between ACC and STG sample from the sample individual for each gene")
-
-pca = prcomp(t(VOOM.GENE_EXPRESSION$E))
-
-df_pca = merge(pca$x, as.data.frame(rownameToFirstColumn(COVARIATES, "SampleID")), 
-               by.x="row.names", 
-               by.y="SampleID" ) %>% 
-  tidyr::separate(Dx.Tissue, into = c("Dx", "Tissue"), sep = "_") %>% 
-  dplyr::mutate(Institution = "NIMH-HBCC")
-
-ggplot(df_pca, aes(PC1, PC2, color=Tissue, shape=Institution)) + geom_point() + theme_bw(17) + theme(aspect.ratio=1)
-```
-
 ### Residual calculation
-Calculate weighted residuals and add back "Dx.Tissue" to the residuals
+Calculate weighted residuals and add back "Dx" to the residuals
 ```{r varsToAddBack}
 # Add variable of interest back to the residuals
-varsToAddIn = grep("Dx.Tissue", colnames(ADJUSTED.FIT$design), value = T)
+varsToAddIn = grep("Dx", colnames(ADJUSTED.FIT$design), value = T)
 RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
@@
 plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>% 
+  dplyr::mutate(Tissue = "DLPFC") %>%  
   dplyr::mutate(Institution = "NIMH-HBCC")
@@
 # Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue'), drop = F])
+COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx'), drop = F])
 
@@
                                    data = COVARIATES, 
-                                   coefficient = c("Dx.TissueOther_ACC", "Dx.TissueControl_ACC"))
+                                   coefficient = c("DxOther", "DxControl"))
 L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
@@
                                    data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueControl_ACC"))
+                                   coefficient = c("DxSCZ", "DxControl"))
 L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
@@
                                    data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_ACC", "Dx.TissueOther_ACC"))
-L4 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueOther_STG", "Dx.TissueControl_STG"))
-L5 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueControl_STG"))
-L6 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
-                                   formula = formula, 
-                                   data = COVARIATES, 
-                                   coefficient = c("Dx.TissueSCZ_STG", "Dx.TissueOther_STG"))
+                                   coefficient = c("DxSCZ", "DxOther"))
 
-L = cbind(L1, L2, L3, L4, L5, L6)
+L = cbind(L1, L2, L3)
 
@@
               glue::glue_collapse(names(L2[L2 != 0]), "-"),
-              glue::glue_collapse(names(L3[L3 != 0]), "-"),
-              glue::glue_collapse(names(L4[L4 != 0]), "-"),
-              glue::glue_collapse(names(L5[L5 != 0]), "-"),
-              glue::glue_collapse(names(L6[L6 != 0]), "-")
+              glue::glue_collapse(names(L3[L3 != 0]), "-")
 )
@@
 # Get gene lengths
-counts = 'syn21867866'
-ALL_USED_IDs = counts
+counts = 'syn21754352'
+ALL_USED_IDs = c(ALL_USED_IDs, counts)
 counts = downloadFile(counts) %>% data.frame()
@@
 # Get differential expression
-DE = lapply(glue::glue("L{1:6}"), function(i, FIT){
+DE = lapply(glue::glue("L{1:3}"), function(i, FIT){
   topTable(FIT, coef=i, number = Inf) %>%
@@
   rbindlist(idcol = 'Comparison') %>%
-  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
+  dplyr::mutate(Comparison = gsub('Dx','',Comparison),
                 Direction = logFC/abs(logFC),
@@
   tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
-  tidyr::separate(ref.state, into = c('Dx1','Tissue'), sep = '_') %>%
-  tidyr::separate(to.state, into = c('Dx2','Tissue'), sep = '_') %>%
-  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_') %>%
+  tidyr::unite(Comparison, `ref.state`, `to.state`, sep = '_vs_') %>%
   left_join(Ensemble2HGNC) %>%
@@
 
-  tissue	= 'superior temporal gyrus, anterior cingulate cortex', 
+  tissue	= 'dorsolateral prefrontal cortex', 
   study = 'CMC',
@@
 # Code
-CODE <- Folder(name = "Differential Expression - Dx.Tissue", parentId = parentId)
+CODE <- Folder(name = "Differential Expression - Dx", parentId = parentId)
 CODE <- synStore(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_STG_Dx.Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_STG_Dx.Tissue_netResidualExpression.tsv', 
-                name = 'Dx.Tissue Normalised, covariates removed residual expression (for network analysis)', 
+  write.table(file = 'CMC_HBCC_DLPFC_Dx_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+nEXP_OBJ = File('CMC_HBCC_DLPFC_Dx_netResidualExpression.tsv', 
+                name = 'Dx Normalised, covariates removed residual expression (for network analysis)', 
                 parentId = CODE$properties$id)
@@
 # Store differential expression results
-write.table(DE, file = 'CMC_ACC_STG_Dx.Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_ACC_STG_Dx.Tissue_DiffExpression.tsv', 
-                name = 'Differential Expression Results (Dx.Tissue)', 
+write.table(DE, file = 'CMC_HBCC_DLPFC_Dx_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('CMC_HBCC_DLPFC_Dx_DiffExpression.tsv', 
+                name = 'Differential Expression Results (Dx)', 
                 parentId = CODE$properties$id)
@@
 
-# Store adjusted fit linear mixed model object as RDS - Dx.Tissue
-saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_STG_Dx.Tissue_dream-output.rds')
-ADJ_OBJ = File('CMC_ACC_STG_Dx.Tissue_dream-output.rds',
-               name = 'Dream Output object - Dx.Tissue', 
+# Store adjusted fit linear mixed model object as RDS - Dx
+saveRDS(ADJUSTED.FIT, file = 'CMC_HBCC_DLPFC_Dx_dream-output.rds')
+ADJ_OBJ = File('CMC_HBCC_DLPFC_Dx_dream-output.rds',
+               name = 'Dream Output object - Dx', 
                parentId = CODE$properties$id)

### 2020-04-07 cmc-rnaseq-de/Differential-Expression_Dx.Rmd
fix knit to folder entity

@@
 ---
-title: "Differential Expression"
+title: "Differential Expression - Dx"
 author: "Kelsey Montgomery, Gabriel Hoffman, Thaneer Perumal"
@@
 
-knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/Differential-Expression.Rmd",
+knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/Differential-Expression_Dx.Rmd",
                                  parentId ="syn21859863",

### 2020-04-07 cmc-rnaseq-de/Differential-Expression_Dx.Rmd
fix factor covars hbcc dlpfc

@@
 # Set variable class
-FactorCovariates <- c("Individual_ID", "Reported_Gender", "Dx", "IsolationBatch")
+FactorCovariates <- c("Individual_ID", "Institution", "Reported_Gender", "LibraryBatch", "Dx", "RibozeroBatch", "FlowcellBatch")
 COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)

### 2020-04-07 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
test ribozero batch in hbcc dlpfc cohort

--- 2020-04-02 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
 ```{r covariates.clustering}
-FactorCovariates <- c('Individual_ID', "Reported_Gender", "LibraryBatch", "Dx", "FlowcellBatch")
+FactorCovariates <- c('Individual_ID', "Reported_Gender", "LibraryBatch", "Dx", "FlowcellBatch", "RibozeroBatch")
 ContCovariates <- c("ageOfDeath", "PMI", "RIN", "MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate",
-                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", 
+                    "GenesDetected", "ExpProfEfficiency", "rRNARate", "TotalReads", "AlignmentRate",
                     "EV.1", "EV.2", "EV.3", "EV.4", "EV.5")

### 2020-04-07 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd
input counts are genes

@@
 title: "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - SALMON - GRCh38 - Model Selection"
-author: "Kelsey Montgomery, Gabriel Hoffman, Thaneer Perumal"
+author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
 date: "`r date()`"
@@
 
-knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd",
+knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38.Rmd",
                                  parentId ="syn21435176",
-                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - SALMON - GRCh38")
+                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38")
 ```
@@
 library(edgeR)
+library(cqn)
 library(lme4)
@@
 
-```{r synapse.parameters, include=FALSE, cache=FALSE}
+```{r synapse.parameters, include=FALSE, cache=TRUE}
 parentId = 'syn21435176';
@@
 
-thisFileName <- 'MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd'
+thisFileName <- 'MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38.Rmd'
 
@@
 # Download reprocessed counts
-counts = 'syn21869686'
+counts = 'syn21899614'
 ALL_USED_IDs = counts
@@
 
+rownames(counts) <- counts$Geneid
 counts = counts[,indToRetain]
@@
 ## Get GC content from biomart
-backgroundGenes = data.frame(transcript_id = rownames(counts)) %>%
-  dplyr::mutate(id = transcript_id) %>%
-  tidyr::separate(id, c('ensembl_transcript_id','position'), sep = '\\.')
+backgroundGenes = data.frame(gene_id = rownames(counts)) %>%
+  dplyr::mutate(id = gene_id) %>%
+  tidyr::separate(id, c('ensembl_gene_id','position'), sep = '\\.')
 
@@
 # Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_transcript_id", "hgnc_trans_name", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
-                       filters = "ensembl_transcript_id", values = backgroundGenes$ensembl_transcript_id,
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
+                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                        mart = mart)
@@
   dplyr::left_join(backgroundGenes) %>% 
-  dplyr::select(transcript_id, percentage_gene_gc_content, chromosome_name) %>%
+  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
   unique
-rownames(GENE.GC.CONT) = GENE.GC.CONT$transcript_id
+rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
 ```
@@
 ```{r Reported_Gender_plots}
-counts$transcript_id = rownames(counts) 
+counts$gene_id = rownames(counts) 
 REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
@@
   filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
-  tidyr::gather(key = item, value = value, -c(transcript_id, chromosome_name)) %>%
+  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
   dplyr::mutate(value = log(value)) %>%
@@
 ##XIST and UTY expression 
-#ENST00000416330.2 and ENST00000329134.9 
+#ENSG00000229807.11 and ENSG00000183878.15 
 FILT <- REPORTED.GENDER.COUNTS %>% 
-  filter(transcript_id == "ENST00000416330.2" | transcript_id == "ENST00000329134.9") %>% 
+  filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>% 
   dplyr::select(-one_of("chromosome_name")) %>% 
-  tidyr::spread(key = transcript_id, value = `counts(log)`) %>% 
-  mutate(XIST = as.numeric(`ENST00000416330.2`)) %>% 
-  mutate(UTY = as.numeric(`ENST00000329134.9`))
+  tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
+  mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>% 
+  mutate(UTY = as.numeric(`ENSG00000183878.15`))
 
@@
 ```
-
 ### Covariates clustering
@@
 ### Filter genes
-Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Remove genes with percentage GC content
-```{r cpmnormalisation, cache.lazy = FALSE}
-counts$transcript_id = NULL
+Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Also remove genes with missing gene length and percentage GC content
+```{r cpmnormalisation}
+counts$gene_id = NULL
 
@@
   unique() %>%
-  intersect(GENE.GC.CONT$transcript_id[!is.na(GENE.GC.CONT$percentage_gene_gc_content)])
+  intersect(GENE.GC.CONT$gene_id[!is.na(GENE.GC.CONT$percentage_gene_gc_content)]) %>%
+  intersect(gene_lengths$gene_id[!is.na(gene_lengths$Length)])
 
@@
 pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
-  dplyr::rename(transcript_id = genes) %>%
+  dplyr::rename(gene_id = genes) %>%
   left_join(backgroundGenes) %>%
@@
 tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
-  rownameToFirstColumn('ensembl_transcript_id') %>%
-  tidyr::gather(SampleID, logCPM, -ensembl_transcript_id) %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
   left_join(COVARIATES %>%
@@
 gc()
+```
+
+### Distribution of samples (log cpm)
+```{r lcpm.dist, cache=FALSE}
+# Plot abberent distribution of logcpm counts
+tmp1 = (CQN.GENE_EXPRESSION$E) %>%
+  rownameToFirstColumn('gene_id') %>%
+  tidyr::gather(SampleID, logCPM, -gene_id) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID'))
+
+p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx.Tissue, scale = 'free')
+p
 ```
@@
   dplyr::rename(baseModel = BIC)
+message("passing...")
 # Iteratively identify covariates in phase I

### 2020-04-07 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd
rm CQN

@@
 gc()
-```
-
-### Distribution of samples (log cpm)
-```{r lcpm.dist, cache=FALSE}
-# Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$E) %>%
-  rownameToFirstColumn('gene_id') %>%
-  tidyr::gather(SampleID, logCPM, -gene_id) %>%
-  left_join(COVARIATES %>%
-              rownameToFirstColumn('SampleID'))
-
-p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
-p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx.Tissue, scale = 'free')
-p
 ```

### 2020-04-07 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd
fix header

@@
 
-knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38.Rmd",
+knit2synapse::createAndKnitToFolderEntity(file = "./cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd",
                                  parentId ="syn21435176",
-                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38")
+                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - SALMON - GRCh38")
 ```
@@
 
-thisFileName <- 'MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38.Rmd'
+thisFileName <- 'MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd'
 

### 2020-04-08 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
note package, correct branch

@@
 ALL_USED_IDs = METADATA_QC_DLPFC_ID
-metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 18) %>%
+metadata = downloadFile_version(METADATA_QC_DLPFC_ID, version = 19) %>%
   dplyr::select(`Individual_ID`, Institution, Cohort, `Reported_Gender`, Sex, Ethnicity, ageOfDeath, `PMI_(in_hours)`, Dx, 

### 2020-04-08 cmc-rnaseq-de/Differential-Expression_Tissue.Rmd
add rRNARate to covars to adjust

--- 2020-04-06 cmc-rnaseq-de/Differential-Expression.Rmd
@@
+###Draft
+
 ---
-title: "Differential Expression"
+title: "Differential Expression - Tissue"
 author: "Kelsey Montgomery, Gabriel Hoffman, Thaneer Perumal"
@@
                                  parentId ="syn21887806",
-                                 folderName = "Differential Expression - Dx.Tissue")
+                                 folderName = "Differential Expression - Tissue")
 ```
@@
 activityName = 'Differential Expression';
-activityDescription = 'Differential Expression - Dx.Tissue';
+activityDescription = 'Differential Expression - Tissue';
 
-thisFileName <- 'Differential-Expression.Rmd'
+thisFileName <- 'Differential-Expression_Tissue.Rmd'
 
@@
 # Set covariates to adjust for from model selection
-adjust.covars <- c("Dx.Tissue", "RIN", "IntragenicRate")
+adjust.covars <- c("Dx.Tissue", "RIN", "rRNARate", "IntragenicRate")
 ```
 
-### Normalisation - Dx.Tissue
-1. Dx.Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and tissue)
+### Normalisation - Tissue
+1. Tissue is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis and reported gender)
 2. Individual_ID is chosen as random effect
-
-```{r iterative.normalisation, results='asis'}
+```{r iterative.normalisation-tissue, results='asis'}
+###we might not need this adjustment
+COVARIATES <- tidyr::separate(COVARIATES, Dx.Tissue.Gender, c("Dx", "Tissue", "Reported_Gender")) %>% 
+  dplyr::mutate(Dx = factor(Dx, levels = c("SCZ", "Control", "Other")),
+                Tissue = factor(Tissue), 
+                Reported_Gender = factor(Reported_Gender))
+adjust.covars[1] <- "Dx"
+adjust.covars <- c(adjust.covars, "Tissue", "Reported_Gender")
 writeLines(paste('Using following covariates in the model:',
@@
                  'as fixed effects and Individual_ID is chosen as random effect'))
-
 # Post adjusted formula
@@
   glue::glue(" + (1|{random_effect})")
-
 # Estimate voom weights with DREAM
@@
 geneExpr = edgeR::calcNormFactors(geneExpr)
-
 VOOM.GENE_EXPRESSION = variancePartition::voomWithDreamWeights(counts = geneExpr, 
@@
                                                                save.plot = TRUE)
-
-# How to plot a voom object if save.plot=TRUE
-# Set plot generic to voom result EList
-#' @export
-setMethod("plot", signature(x="EList", y="missing"),
-  function(x,  y, ...){
-    if( is.null(x$voom.xy) || is.null(x$voom.line)){
-        stop("x does not contain the trend information.\nvoom() must be run with save.plot=TRUE")
-    }
-    # points
-    df = data.frame(x = x$voom.xy$x,
-                    y = x$voom.xy$y)
-    # trend line
-    df.line = data.frame(x = x$voom.line$x,
-                        y = x$voom.line$y)
-    ggplot(df, aes(x,y)) + geom_point(size=.1) + theme_bw(15) + 
-      theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) + 
-      geom_line(data=df.line, aes(x,y), color="red") + xlab(bquote(log[2](count + 0.5))) + 
-      ylab(expression( sqrt("standard deviation"))) + ggtitle("Voom: Mean-variance trend") + 
-      ylim(0, max(df$y))
-  }
-)
-
 plot(VOOM.GENE_EXPRESSION)
-
 # Fit linear model using new weights and new design
-VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION
-ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
+VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E
+Tissue_ADJUSTED.FIT = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION, 
                                         formula = formula,
@@
                                         computeResiduals = TRUE)
-
 col_type <- dplyr::select(COVARIATES, c(adjust.covars, "Individual_ID")) %>%
@@
   tidyr::gather(variable, class)
-
 # Model categorical variables as random to report variance fractions
@@
   })
-
 formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(formula_for_variance_fractions, sep = " + "))
-
 violin_plot <- variancePartition::fitExtractVarPartModel(VOOM.GENE_EXPRESSION,
@@
                                                          data = COVARIATES)
-
 plotVarPart(sortCols(violin_plot))
-
 # Residuals after normalisation
-RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION)
-
+Tissue_RESIDUAL.GENE_EXPRESSION = residuals(Tissue_ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
 # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
-residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, COVARIATES,
+residualSigCovars = runPCAandPlotCorrelations(Tissue_RESIDUAL.GENE_EXPRESSION, COVARIATES,
                                               'adjusted design(voom-normalized)',
@@
 
-### Tissue Correlation
-```{r tissue_correlation}
-hasReps = rownameToFirstColumn(COVARIATES, "SampleID") %>%
-  dplyr::select(Individual_ID) %>% 
-  table
+### Differential expression analysis (ACC - DLPFC)
+Genes that are differentially expressed at an FDR <= 0.05 are:
+```{r set-formula-tissue}
+formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+```
 
-df_stg = rownameToFirstColumn(COVARIATES, "SampleID") %>% 
-  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
-  filter(grepl("STG", Dx.Tissue)) %>% 
-  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
-df_acc = rownameToFirstColumn(COVARIATES, "SampleID") %>%
-  filter( Individual_ID %in% names(hasReps[hasReps==2])) %>%
-  filter(grepl("ACC", Dx.Tissue)) %>% 
-  dplyr::select(SampleID, Individual_ID, Dx.Tissue)
+`r formula`
 
-df_merge = merge(df_stg, df_acc, by="Individual_ID")
-
-res_std <- rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id")
-
-corRes = sapply(res_std$ensembl_gene_id, function(ensGene){
-
-  x = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.x,drop=FALSE]
-  y = VOOM.GENE_EXPRESSION$E[ensGene,df_merge$SampleID.y,drop=FALSE]
-
-  cor(t(x)[,1], t(y)[,1], use="pairwise.complete.obs")
-})
-
-hist(corRes, xlab="Correlation", main="Correlation between ACC and STG sample from the sample individual for each gene")
-
-pca = prcomp(t(VOOM.GENE_EXPRESSION$E))
-
-df_pca = merge(pca$x, as.data.frame(rownameToFirstColumn(COVARIATES, "SampleID")), 
-               by.x="row.names", 
-               by.y="SampleID" ) %>% 
-  tidyr::separate(Dx.Tissue, into = c("Dx", "Tissue"), sep = "_") %>% 
-  dplyr::mutate(Institution = "NIMH-HBCC")
-
-ggplot(df_pca, aes(PC1, PC2, color=Tissue, shape=Institution)) + geom_point() + theme_bw(17) + theme(aspect.ratio=1)
+```{r diffExp-tissue, fig.height=10, fig.width=15}
+L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("TissueDLPFC"))
+L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("DxSCZ", "DxControl"))
+L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
+                                   formula = formula, 
+                                   data = COVARIATES, 
+                                   coefficient = c("Reported_GenderMale"))
+L = cbind(L1, L2, L3)
+contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
+              glue::glue_collapse(names(L2[L2 != 0]), "-"),
+              glue::glue_collapse(names(L3[L3 != 0]), "-")
+)
+# Visualize contrast matrix
+plotContrasts(L) 
+# Fit contrast
+FIT.CONTR = variancePartition::dream(exprObj = VOOM.GENE_EXPRESSION,
+                                     formula = formula,
+                                     data = COVARIATES,
+                                     L = L)
+# Get differential expression
+DE_3 = lapply(glue::glue("L{1:3}"), function(i, FIT){
+  topTable(FIT, coef=i, number = Inf) %>%
+    rownameToFirstColumn('gene_id') %>% 
+    left_join(backgroundGenes)
+}, FIT.CONTR) 
+names(DE_3) <- contrast
+DE_3 = DE_3%>% 
+  rbindlist(idcol = 'Comparison') %>%
+  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
+                Direction = logFC/abs(logFC),
+                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
+                Direction = as.character(Direction)) %>% 
+  dplyr::mutate(Comparison = ifelse(Comparison %in% "Reported_GenderMale", "Female-Male", Comparison),
+                Comparison = ifelse(Comparison %in% "DxSCZ-DxControl", "SCZ-Control", Comparison),
+                Comparison = ifelse(Comparison %in% "TissueDLPFC", "ACC-DLPFC", Comparison)) %>% 
+  left_join(Ensemble2HGNC) %>%
+  left_join(gene_lengths %>% rownameToFirstColumn('gene_id'))
+DE_3$Direction[DE_3$adj.P.Val > 0.05 | abs(DE_3$logFC) < log2(1.2)] = 'NONE'
+tmp = DE_3 %>%
+  dplyr::filter(adj.P.Val <= 0.05) %>%
+  dplyr::select(ensembl_gene_id, Comparison) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
+tmp1 = DE_3 %>%
+  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
+  dplyr::select(ensembl_gene_id, Comparison) %>%
+  group_by(Comparison) %>%
+  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
+kable(full_join(tmp,tmp1))
+p = ggplot(DE_3, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
+p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
+p = p + facet_grid(Comparison~., scales = 'fixed')
+p
 ```
@@
 # Code
-CODE <- Folder(name = "Differential Expression - Dx.Tissue", parentId = parentId)
+CODE <- Folder(name = "Differential Expression - Tissue", parentId = parentId)
 CODE <- synStore(CODE)
@@
   rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_STG_Dx.Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
-nEXP_OBJ = File('CMC_ACC_STG_Dx.Tissue_netResidualExpression.tsv', 
-                name = 'Dx.Tissue Normalised, covariates removed residual expression (for network analysis)', 
+  write.table(file = 'CMC_ACC_STG_Tissue_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
+nEXP_OBJ = File('CMC_ACC_STG_Tissue_netResidualExpression.tsv', 
+                name = 'Tissue Normalised, covariates removed residual expression (for network analysis)', 
                 parentId = CODE$properties$id)
@@
 # Store differential expression results
-write.table(DE, file = 'CMC_ACC_STG_Dx.Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
-DEXP_OBJ = File('CMC_ACC_STG_Dx.Tissue_DiffExpression.tsv', 
-                name = 'Differential Expression Results (Dx.Tissue)', 
+write.table(DE, file = 'CMC_ACC_STG_Tissue_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
+DEXP_OBJ = File('CMC_ACC_STG_Tissue_DiffExpression.tsv', 
+                name = 'Differential Expression Results (Tissue)', 
                 parentId = CODE$properties$id)
@@
 
-# Store adjusted fit linear mixed model object as RDS - Dx.Tissue
-saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_STG_Dx.Tissue_dream-output.rds')
-ADJ_OBJ = File('CMC_ACC_STG_Dx.Tissue_dream-output.rds',
-               name = 'Dream Output object - Dx.Tissue', 
+# Store adjusted fit linear mixed model object as RDS - Tissue
+saveRDS(ADJUSTED.FIT, file = 'CMC_ACC_STG_Tissue_dream-output.rds')
+ADJ_OBJ = File('CMC_ACC_STG_Tissue_dream-output.rds',
+               name = 'Dream Output object - Tissue', 
                parentId = CODE$properties$id)

### 2020-04-08 cmc-rnaseq-de/Differential-Expression_Dx.Rmd
Institution not relevant to HBCC DLPFC cohort

--- 2020-04-07 cmc-rnaseq-de/Differential-Expression_Dx.Rmd
@@
 # Set variable class
-FactorCovariates <- c("Individual_ID", "Reported_Gender", "Dx", "IsolationBatch")
+FactorCovariates <- c("Individual_ID", "Reported_Gender", "LibraryBatch", "Dx", "RibozeroBatch", "FlowcellBatch")
 COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)

### 2020-04-08 cmc-rnaseq-de/Differential-Expression_Dx.Rmd
fix plot facet

@@
 mart <- biomaRt::useMart(biomart = "ENSEMBL_MART_ENSEMBL",
-                host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)
+                host = "useast.ensembl.org", # Ensembl Release 99 (January 2020)
                 dataset = "hsapiens_gene_ensembl")
@@
   dplyr::filter(adj.P.Val <= 0.05) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tissue) %>%
+  dplyr::select(ensembl_gene_id, Comparison) %>%
+  group_by(Comparison) %>%
   dplyr::summarise(FDR_0_05 = length(unique(ensembl_gene_id)))
@@
   dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
-  dplyr::select(ensembl_gene_id, Comparison, Tissue) %>%
-  group_by(Comparison, Tissue) %>%
+  dplyr::select(ensembl_gene_id, Comparison) %>%
+  group_by(Comparison) %>%
   dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id)))
@@
 p = p + scale_color_manual(values = c('green','grey','red')) + theme_bw() %+replace% theme(legend.position = 'top')
-p = p + facet_grid(Tissue+.~Comparison, scales = 'fixed')
+p = p + facet_grid(~Comparison, scales = 'fixed')
 p

### 2020-04-08 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd
different XIST isoform in salmon dataset

@@
 mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
-                host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)
+                host = "useast.ensembl.org", # Ensembl Release 99 (January 2020)
                 dataset = "hsapiens_gene_ensembl")
@@
 ##XIST and UTY expression 
-#ENSG00000229807.11 and ENSG00000183878.15 
+#ENSG00000229807.12 and ENSG00000183878.15 
 FILT <- REPORTED.GENDER.COUNTS %>% 
-  filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>% 
+  filter(gene_id == "ENSG00000229807.12" | gene_id == "ENSG00000183878.15") %>% 
   dplyr::select(-one_of("chromosome_name")) %>% 
   tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
-  mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>% 
+  mutate(XIST = as.numeric(`ENSG00000229807.12`)) %>% 
   mutate(UTY = as.numeric(`ENSG00000183878.15`))

### 2020-04-08 cmc-rnaseq-de/Differential-Expression_Dx.Rmd
fix plot facet

@@
 pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
-pl[[1]] = pl[[1]] + facet_grid(Tissue~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[1]] = pl[[1]] + scale_color_manual(values = c('green','grey','red'))
 pl[[1]] = pl[[1]] + theme(legend.position = 'top')
@@
 pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gene_gc_content, y = logFC))
-pl[[2]] = pl[[2]] + facet_grid(Tissue~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[2]] = pl[[2]] + scale_color_manual(values = c('green','grey','red'))
 pl[[2]] = pl[[2]] + theme(legend.position = 'top')
@@
 pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
-pl[[3]] = pl[[3]] + facet_grid(Tissue~.) + scale_color_manual(values = c('green','grey','red'))
+pl[[3]] = pl[[3]] + scale_color_manual(values = c('green','grey','red'))
 pl[[3]] = pl[[3]] + theme(legend.position = 'top')

### 2020-04-14 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
save to testing folder

@@
                                  parentId ="syn21435176",
-                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN")
+                                 folderName = "test mvBIC - MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN")
 ```
@@
 # Code
-CODE <- Folder(name = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
+CODE <- Folder(name = "test mvBIC - MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
 CODE <- synStore(CODE)

### 2020-04-14 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
biomart down - download from Synapse

@@
 ## Get GC content from biomart
-backgroundGenes = data.frame(gene_id = rownames(counts)) %>%
-  dplyr::mutate(id = gene_id) %>%
-  tidyr::separate(id, c('ensembl_gene_id','position'), sep = '\\.')
-
-# Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
-                host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)
-                dataset = "hsapiens_gene_ensembl")
-
-# Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
-                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
-                       mart = mart)
-
-GENE.GC.CONT = Ensemble2HGNC %>%
-  dplyr::left_join(backgroundGenes) %>% 
-  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
-  unique
-rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
+# backgroundGenes = data.frame(gene_id = rownames(counts)) %>%
+#   dplyr::mutate(id = gene_id) %>%
+#   tidyr::separate(id, c('ensembl_gene_id','position'), sep = '\\.')
+# 
+# # Define biomart object
+# mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
+#                 host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)
+#                 dataset = "hsapiens_gene_ensembl")
+# 
+# # Query biomart
+# Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
+#                        filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
+#                        mart = mart)
+# 
+# GENE.GC.CONT = Ensemble2HGNC %>%
+#   dplyr::left_join(backgroundGenes) %>% 
+#   dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
+#   unique
+# rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id]
+GENE.GC.CONT = downloadFile("syn21907981")
 ```

### 2020-04-14 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
comment out nonessential code for testing

@@
 
-### Sex Check
-```{r Reported_Gender_plots}
-counts$gene_id = rownames(counts) 
-REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
-  left_join(counts) %>%
-  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
-  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
-  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
-  dplyr::mutate(value = log(value)) %>%
-  dplyr::rename(`counts(log)`= value) %>% 
-  dplyr::rename(SampleID = item) %>%
-  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
-  dplyr::rename(`Reported Gender` = Reported_Gender)
+<!-- ### Sex Check -->
+<!-- ```{r Reported_Gender_plots} -->
+<!-- counts$gene_id = rownames(counts)  -->
+<!-- REPORTED.GENDER.COUNTS = GENE.GC.CONT %>%  -->
+<!--   left_join(counts) %>% -->
+<!--   dplyr::select(-one_of("percentage_gene_gc_content")) %>% -->
+<!--   filter(chromosome_name == "X" |chromosome_name == "Y") %>%  -->
+<!--   tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>% -->
+<!--   dplyr::mutate(value = log(value)) %>% -->
+<!--   dplyr::rename(`counts(log)`= value) %>%  -->
+<!--   dplyr::rename(SampleID = item) %>% -->
+<!--   left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>%  -->
+<!--   dplyr::rename(`Reported Gender` = Reported_Gender)  -->
 
-##XIST and UTY expression 
-#ENSG00000229807.11 and ENSG00000183878.15 
-FILT <- REPORTED.GENDER.COUNTS %>% 
-  filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>% 
-  dplyr::select(-one_of("chromosome_name")) %>% 
-  tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
-  mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>% 
-  mutate(UTY = as.numeric(`ENSG00000183878.15`))
+<!-- ##XIST and UTY expression  -->
+<!-- #ENSG00000229807.11 and ENSG00000183878.15  -->
+<!-- FILT <- REPORTED.GENDER.COUNTS %>%  -->
+<!--   filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>%  -->
+<!--   dplyr::select(-one_of("chromosome_name")) %>%  -->
+<!--   tidyr::spread(key = gene_id, value = `counts(log)`) %>%  -->
+<!--   mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>%  -->
+<!--   mutate(UTY = as.numeric(`ENSG00000183878.15`)) -->
 
-p = ggplot(FILT, aes (x= XIST, y = UTY)) 
-p = p + geom_point(aes(color=`Reported Gender`)) 
-p
-```
+<!-- p = ggplot(FILT, aes (x= XIST, y = UTY))  -->
+<!-- p = p + geom_point(aes(color=`Reported Gender`))  -->
+<!-- p -->
+<!-- ``` -->
 ### Covariates clustering
@@
 
-## Explore metatdata
-```{r data.explore, fig.width = 12, fig.height = 12}
-my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))
+<!-- ## Explore metatdata -->
+<!-- ```{r data.explore, fig.width = 12, fig.height = 12} -->
+<!-- my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5)) -->
 
-# RIN
-p = list()
-p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot()
-p[[1]] = p[[1]] + ggtitle('RIN') + my.theme
+<!-- # RIN -->
+<!-- p = list() -->
+<!-- p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot() -->
+<!-- p[[1]] = p[[1]] + ggtitle('RIN') + my.theme -->
 
-# Age of Death
-p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = ageOfDeath)) + geom_boxplot()
-p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme
+<!-- # Age of Death -->
+<!-- p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = ageOfDeath)) + geom_boxplot() -->
+<!-- p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme -->
 
-# PMI
-p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot()
-p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme
+<!-- # PMI -->
+<!-- p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot() -->
+<!-- p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme -->
 
-# Intronic Rate
-p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot()
-p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme
+<!-- # Intronic Rate -->
+<!-- p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot() -->
+<!-- p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme -->
 
-# IntergenicRate
-p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot()
-p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme
+<!-- # IntergenicRate -->
+<!-- p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot() -->
+<!-- p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme -->
 
-# Transcripts Detected
-p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntragenicRate)) + geom_boxplot()
-p[[6]] = p[[6]] + ggtitle('Intragenic Rate') + my.theme
+<!-- # Transcripts Detected -->
+<!-- p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntragenicRate)) + geom_boxplot() -->
+<!-- p[[6]] = p[[6]] + ggtitle('Intragenic Rate') + my.theme -->
 
-# Mapped Reads
-p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot()
-p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme
+<!-- # Mapped Reads -->
+<!-- p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot() -->
+<!-- p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme -->
 
-# PercentAligned
-p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TotalReads)) + geom_boxplot()
-p[[8]] = p[[8]] + ggtitle('Total Reads') + my.theme
+<!-- # PercentAligned -->
+<!-- p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TotalReads)) + geom_boxplot() -->
+<!-- p[[8]] = p[[8]] + ggtitle('Total Reads') + my.theme -->
 
-# rRNARate
-p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot()
-p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme
+<!-- # rRNARate -->
+<!-- p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot() -->
+<!-- p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme -->
 
-multiplot(plotlist = p, cols = 3)
+<!-- multiplot(plotlist = p, cols = 3) -->
 
-# Institution
-# vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES)
+<!-- # Institution -->
+<!-- # vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES) -->
 
-# Gender
-# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
-```
+<!-- # Gender -->
+<!-- # vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES) -->
+<!-- ``` -->
 ### Filter genes
@@
 
-### Co-expression distribution
-Coexpression of genes 
-```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
-cr = cor(t(CQN.GENE_EXPRESSION$E))
-hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
-```
+<!-- ### Co-expression distribution -->
+<!-- Coexpression of genes  -->
+<!-- ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5} -->
+<!-- cr = cor(t(CQN.GENE_EXPRESSION$E)) -->
+<!-- hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation') -->
+<!-- ``` -->
 
-### Sample clustering
-```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
-# Find principal components of expression to plot
-PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)
+<!-- ### Sample clustering -->
+<!-- ```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE} -->
+<!-- # Find principal components of expression to plot -->
+<!-- PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T) -->
 
-# Plot first 2 PCs
-plotdata <- data.frame(SampleID=rownames(PC$rotation), 
-                       PC1=PC$rotation[,1], 
-                       PC2=PC$rotation[,2])
+<!-- # Plot first 2 PCs -->
+<!-- plotdata <- data.frame(SampleID=rownames(PC$rotation),  -->
+<!--                        PC1=PC$rotation[,1],  -->
+<!--                        PC2=PC$rotation[,2]) -->
 
-plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
-  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+<!-- plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>% -->
+<!--   tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') -->
 
-p <- ggplot(plotdata, aes(x=PC1, y=PC2))
-p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
-p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
-# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
-p
-```
-Tree based clustering of samples
-```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
-# Eucledian tree based analysis
-COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue")])
-COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
+<!-- p <- ggplot(plotdata, aes(x=PC1, y=PC2)) -->
+<!-- p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath)) -->
+<!-- p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.) -->
+<!-- # p <- p + geom_text(aes(label= SampleID), size=4, hjust=0) -->
+<!-- p -->
+<!-- ``` -->
+<!-- Tree based clustering of samples -->
+<!-- ```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'} -->
+<!-- # Eucledian tree based analysis -->
+<!-- COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue")]) -->
+<!-- COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0 -->
 
-tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
-cols = WGCNA::labels2colors(COVARIATES.tmp);
+<!-- tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E))) -->
+<!-- cols = WGCNA::labels2colors(COVARIATES.tmp); -->
 
-WGCNA::plotDendroAndColors(tree, 
-                           colors = cols, 
-                           dendroLabels = FALSE, 
-                           abHeight = 0.80, 
-                           main = "Sample dendrogram",
-                           groupLabels = colnames(COVARIATES.tmp))
-```
-```{r temp, include = F}
-dev.off()
-gc()
-```
+<!-- WGCNA::plotDendroAndColors(tree,  -->
+<!--                            colors = cols,  -->
+<!--                            dendroLabels = FALSE,  -->
+<!--                            abHeight = 0.80,  -->
+<!--                            main = "Sample dendrogram", -->
+<!--                            groupLabels = colnames(COVARIATES.tmp)) -->
+<!-- ``` -->
+<!-- ```{r temp, include = F} -->
+<!-- dev.off() -->
+<!-- gc() -->
+<!-- ``` -->
 
-### Distribution of samples (log cpm)
-```{r lcpm.dist, cache=FALSE}
-# Plot abberent distribution of logcpm counts
-tmp1 = (CQN.GENE_EXPRESSION$E) %>%
-  rownameToFirstColumn('gene_id') %>%
-  tidyr::gather(SampleID, logCPM, -gene_id) %>%
-  left_join(COVARIATES %>%
-              rownameToFirstColumn('SampleID'))
+<!-- ### Distribution of samples (log cpm) -->
+<!-- ```{r lcpm.dist, cache=FALSE} -->
+<!-- # Plot abberent distribution of logcpm counts -->
+<!-- tmp1 = (CQN.GENE_EXPRESSION$E) %>% -->
+<!--   rownameToFirstColumn('gene_id') %>% -->
+<!--   tidyr::gather(SampleID, logCPM, -gene_id) %>% -->
+<!--   left_join(COVARIATES %>% -->
+<!--               rownameToFirstColumn('SampleID')) -->
 
-p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
-p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx.Tissue, scale = 'free')
-p
-```
+<!-- p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density()  -->
+<!-- p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx.Tissue, scale = 'free') -->
+<!-- p -->
+<!-- ``` -->
 

### 2020-04-14 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
comment out nonessential code for testing

@@
 ```
-Correlation/association between covariates at an FDR <= 0.1
-```{r covariates.correlation, fig.width=10, fig.height=8}
-COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
-tmp = COVARIATES.CORRELATION$ESTIMATE
-tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
-h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
-ComplexHeatmap::draw(h, heatmap_legend_side = 'left')
-```
+<!-- Correlation/association between covariates at an FDR <= 0.1 -->
+<!-- ```{r covariates.correlation, fig.width=10, fig.height=8} -->
+<!-- COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05) -->
+<!-- tmp = COVARIATES.CORRELATION$ESTIMATE -->
+<!-- tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0 -->
+<!-- h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate') -->
+<!-- ComplexHeatmap::draw(h, heatmap_legend_side = 'left') -->
+<!-- ``` -->
 

### 2020-04-14 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
random sampling of genes

@@
 ```{r gene.param}
-## Get GC content from biomart
-# backgroundGenes = data.frame(gene_id = rownames(counts)) %>%
-#   dplyr::mutate(id = gene_id) %>%
-#   tidyr::separate(id, c('ensembl_gene_id','position'), sep = '\\.')
-# 
+# Get GC content from biomart
+backgroundGenes = data.frame(gene_id = rownames(counts)) %>%
+  dplyr::mutate(id = gene_id) %>%
+  tidyr::separate(id, c('ensembl_gene_id','position'), sep = '\\.')
+
 # # Define biomart object
@@
 #                        mart = mart)
-# 
-# GENE.GC.CONT = Ensemble2HGNC %>%
-#   dplyr::left_join(backgroundGenes) %>% 
-#   dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
-#   unique
-# rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id]
-GENE.GC.CONT = downloadFile("syn21907981")
+Ensemble2HGNC = downloadFile("syn21907981")
+GENE.GC.CONT = Ensemble2HGNC %>%
+  dplyr::left_join(backgroundGenes) %>%
+  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
+  unique
+rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
 ```
@@
 gene_lengths = tibble::column_to_rownames(gene_lengths, var = 'gene_id') %>% as.data.frame()
-CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
+CQN.GENE_EXPRESSION = cqn::cqn(NEW.COUNTS, 
                           x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gene_gc_content'],
@@
 <!-- ``` -->
+
+### Subset Genes
+```{r subset-genes}
+rand_sample <- sample(1:dim(NEW.COUNTS)[1], size = round(dim(NEW.COUNTS)[1]*.6), replace = FALSE)
+subset_NEW.COUNTS<- NEW.COUNTS[rand_sample,]
+
+subset_CQN <- CQN.GENE_EXPRESSION$E[rand_sample,]
+```
 

### 2020-04-14 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
comment out nonessential code for testing

@@
 
-### Significant Covariates
-Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
-```{r preAdjusted.covariates, cache=TRUE}
-# Find correlation between PC's of gene expression with covariates
-preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
-                                                 COVARIATES,
-                                                 'NULL design(voom-normalized)', 
-                                                 isKeyPlot=TRUE, 
-                                                 MIN_PVE_PCT_PC = 1)
-```
+<!-- ### Significant Covariates -->
+<!-- Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates -->
+<!-- ```{r preAdjusted.covariates, cache=TRUE} -->
+<!-- # Find correlation between PC's of gene expression with covariates -->
+<!-- preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E,  -->
+<!--                                                  COVARIATES, -->
+<!--                                                  'NULL design(voom-normalized)',  -->
+<!--                                                  isKeyPlot=TRUE,  -->
+<!--                                                  MIN_PVE_PCT_PC = 1) -->
+<!-- ``` -->
 
-Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
-```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
-preAdjustedSigCovars[["PC_res"]][[2]]$plotData
-```
+<!-- Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars` -->
+<!-- ```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12} -->
+<!-- preAdjustedSigCovars[["PC_res"]][[2]]$plotData -->
+<!-- ``` -->
 

### 2020-04-14 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
mvBIC stepwise regression

@@
 library(githubr) # get the package from devtools::install_github('brian-bot/githubr')
+
+devtools::install_github("GabrielHoffman/mvBIC")
+library(mvBIC)
 
@@
 # Function to fit a mixed effect model and return BIC
-fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){
+fitMixedEffectModel <- function(varsToFit, random_effect, lcpm, covars, covars_to_fit){
 
-  # Get design matrix
-  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
-  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]
-
-  # Estimate voom weights
-  cnt[is.na(cnt)] = 0
-  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights
-
-  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
-                              .combine = rbind,
-                              .packages = c('lme4', 'dplyr'),
-                              .export = c('voomWeights')) %dopar% {
-
-                                # Perform weighted mixed linear models for every gene
-                                expr = data.frame(Gx = lcpm[i,])
-                                data = cbind(expr,
-                                             designMat[rownames(expr),],
-                                             Individual_ID = groupingVar)
-                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
-                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')
-
-                                weights = voomWeights[i,]
-                                BIC = NA
-                                tryCatch({
-                                  mdl = lmer(formula = formula, 
-                                             data = data, 
-                                             weights = weights)
-                                  BIC = BIC(mdl)
-                                }, error = function(e){
-                                  BIC = NA
-                                })
-                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
-                                                BIC = BIC))
-                            }
-  return(modelBIC)
-  }
-# Function to select covariates in a step-wise regression fashion
-selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){
-
-  varsSelected = c()
-  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
-           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
-  while(length(covarsToSelect) != 0){
-    allModelBIC = foreach(i = 1:length(covarsToSelect),
-                          .combine = c,
-                          .packages = pkgs,
-                          .export = c('fitMixedEffectModel')) %dopar% {
-
-                            # Fit a mixed effect linear model and get all BIC
-                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
-                                                           covars[colnames(cnt),'Individual_ID'], 
-                                                           lcpm, 
-                                                           cnt, 
-                                                           covars) %>%
-                            plyr::rename(c('BIC' = covarsToSelect[i]))
-
-                            return(list(modelBIC))
-                          }
-    allModelBIC = plyr::join_all(allModelBIC)
-
-    # Find number of genes that are changed based on BIC
-    nChange = allModelBIC %>%
-      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
-      dplyr::left_join(baseModelBIC) %>%
-      dplyr::mutate(diff = baseModel - Value) %>%
-      dplyr::group_by(ModelName) %>%
-      dplyr::summarise(better = sum(diff >= 5), 
-                       worse = sum(diff <= -5)) %>%
-      dplyr::mutate(nchange = better - worse)
-
-    print(nChange)
-
-    # Select variable
-    tmp = nChange %>%
-      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
-      dplyr::top_n(1, nchange)
-
-    if (dim(tmp)[1] == 0){
-      covarsToSelect = c()
-    } else {
-      baseModelBIC = allModelBIC %>%
-        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
-      colnames(baseModelBIC)[2] = 'baseModel'
-      varsSelected = c(varsSelected, tmp$ModelName)
-      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
-      print(paste('Variable chosen to be added:',tmp$ModelName))
-    }
-  }
-
-  return(varsSelected)
 }
 ```
+
 #### Phase I (clinical, ancestry and sample specific technical variables)
@@
 # Fit a base model with the following covariates
-adjust.covars = c('Dx.Tissue')
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-# Iteratively identify covariates in phase I
-phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'ageOfDeath', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
-vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-adjust.covars = c(adjust.covars, vars.selected)
+base_covars <- "Dx.Tissue"
+random_effect <- "Individual_ID"
+
+# Set base formula
+base_formula <- glue::glue("~ ", glue::glue_collapse(base_covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+
+# Stepwise regression
+phase1 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN,
+                         baseFormula = base_formula, 
+                         data = COVARIATES,
+                         variables = array(c("Institution", "Reported_Gender", "PMI", 
+                                                             "ageOfDeath", "RIN", "RIN2", "EV.1", 
+                                                             "EV.2", "EV.3", "EV.4", "EV.5")))
+phase1
 ```
-Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`
 
@@
 ```{r phase2}
-# Fit a base model with the following covariates
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-# Iteratively identify covariates in phase I
-phase2.covars = c('LibraryBatch', 'FlowcellBatch')
-vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-adjust.covars = c(adjust.covars, vars.selected)
+phase2 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN,
+                         baseFormula = phase1$formula, 
+                         data = COVARIATES,
+                         variables = array(c("LibraryBatch", "FlowcellBatch")))
+phase2
 ```
-Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`
 
@@
 ```{r phase3}
-# Fit a base model with the following covariates
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-# Iteratively identify covariates in phase I
-phase3.covars = c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected", "rRNARate", "TotalReads")
-vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-adjust.covars = c(adjust.covars, vars.selected)
+phase3 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN,
+                         baseFormula = phase2$formula, 
+                         data = COVARIATES,
+                         variables = array(c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected", "rRNARate", "TotalReads")))
+phase3
 ```
@@
 #### Variables included in the model 
-Covariates were added as fixed effects iteratively in three phases if model improvement by BIC criteria was observed in the majority of genes.
-Clinical, ancestry and sample-specific technical variables were tested for model improvement first.
-Covariates related to batch effects were next tested for model improvement and finally, RNASeq alignment-specific covariates.
-
-Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`
+`r phase3$formula`
 

### 2020-04-14 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
comment out nonessential code for testing

@@
 
-### Store files in synapse
-```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
-# Set annotations
-all.annotations = list(
-  dataType = 'geneExpression',
-  dataSubtype = 'processed',
-  assay	 = 'rnaSeq',
+<!-- ### Store files in synapse -->
+<!-- ```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE} -->
+<!-- # Set annotations -->
+<!-- all.annotations = list( -->
+<!--   dataType = 'geneExpression', -->
+<!--   dataSubtype = 'processed', -->
+<!--   assay	 = 'rnaSeq', -->
 
-  tissue	= 'dorsolateral prefrontal cortex, anterior cingulate cortex', 
-  study = 'CMC',
+<!--   tissue	= 'dorsolateral prefrontal cortex, anterior cingulate cortex',  -->
+<!--   study = 'CMC',  -->
 
-  species = 'Human',
-  consortium	= 'CMC',
+<!--   species = 'Human', -->
+<!--   consortium	= 'CMC', -->
 
-  analysisType	= "data normalization",
+<!--   analysisType	= "data normalization", -->
 
-  transcriptNormalizationMethod	= 'CQN',
-  transcriptQuantificationMethod = 'featureCounts',
-  referenceSet = 'GRCh38'
-)
+<!--   transcriptNormalizationMethod	= 'CQN', -->
+<!--   transcriptQuantificationMethod = 'featureCounts', -->
+<!--   referenceSet = 'GRCh38' -->
+<!-- ) -->
 
-# Code
-CODE <- Folder(name = "test mvBIC - MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
-CODE <- synStore(CODE)
+<!-- # Code -->
+<!-- CODE <- Folder(name = "test mvBIC - MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId) -->
+<!-- CODE <- synStore(CODE) -->
 
-# Store covariates
-COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
-write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
-COV_OBJ = synapser::File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
-COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                      executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(COV_OBJ, annotations = all.annotations)
+<!-- # Store covariates -->
+<!-- COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID') -->
+<!-- write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F) -->
+<!-- COV_OBJ = synapser::File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id) -->
+<!-- COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName,  -->
+<!--                       executed = thisFile, activityDescription = activityDescription) -->
+<!-- synSetAnnotations(COV_OBJ, annotations = all.annotations) -->
 
-# Store filtered counts
-PROCESSED_COUNTS$filteredExprMatrix$counts %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
-COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
-COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                      executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(COUNT_OBJ, annotations = all.annotations)
+<!-- # Store filtered counts -->
+<!-- PROCESSED_COUNTS$filteredExprMatrix$counts %>% -->
+<!--   rownameToFirstColumn('ensembl_gene_id') %>% -->
+<!--   write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F) -->
+<!-- COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id) -->
+<!-- COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName,  -->
+<!--                       executed = thisFile, activityDescription = activityDescription) -->
+<!-- synSetAnnotations(COUNT_OBJ, annotations = all.annotations) -->
 
-# Store logCPM
-CQN.GENE_EXPRESSION$E %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
-LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
-LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                      executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(LCOUNT_OBJ, annotations = all.annotations)
+<!-- # Store logCPM -->
+<!-- CQN.GENE_EXPRESSION$E %>% -->
+<!--   rownameToFirstColumn('ensembl_gene_id') %>% -->
+<!--   write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F) -->
+<!-- LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id) -->
+<!-- LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName,  -->
+<!--                       executed = thisFile, activityDescription = activityDescription) -->
+<!-- synSetAnnotations(LCOUNT_OBJ, annotations = all.annotations) -->
 
-# Store cqn offsets
-CQN.GENE_EXPRESSION$offset %>%
-  rownameToFirstColumn('ensembl_gene_id') %>%
-  write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
-OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
-OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName, 
-                      executed = thisFile, activityDescription = activityDescription)
-synSetAnnotations(OFFSET_OBJ, annotations = all.annotations)
-```
+<!-- # Store cqn offsets -->
+<!-- CQN.GENE_EXPRESSION$offset %>% -->
+<!--   rownameToFirstColumn('ensembl_gene_id') %>% -->
+<!--   write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F) -->
+<!-- OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id) -->
+<!-- OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName,  -->
+<!--                       executed = thisFile, activityDescription = activityDescription) -->
+<!-- synSetAnnotations(OFFSET_OBJ, annotations = all.annotations) -->
+<!-- ``` -->
 
-|  *Results*                                  |  *SynapseID*                     |
-|  -----------------------------------------  |   ---------                      |
-|  Covariates                                 |  `r COV_OBJ$properties$id`       |
-|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
-|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
-|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
+<!-- |  *Results*                                  |  *SynapseID*                     | -->
+<!-- |  -----------------------------------------  |   ---------                      | -->
+<!-- |  Covariates                                 |  `r COV_OBJ$properties$id`       | -->
+<!-- |  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     | -->
+<!-- |  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    | -->
+<!-- |  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    | -->
 

### 2020-04-14 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
save to test folder on Synapse

@@
 
-devtools::install_github("GabrielHoffman/mvBIC")
-library(mvBIC)
+library(mvBIC) # devtools::install_github("GabrielHoffman/mvBIC")
 
@@
 # Github link
-thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='master')
+thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='test-mvBIC')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))
@@
 
-<!-- # Code -->
-<!-- CODE <- Folder(name = "test mvBIC - MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId) -->
-<!-- CODE <- synStore(CODE) -->
-
+# Code
+``` {r save_code}
+CODE <- Folder(name = "test mvBIC - MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
+CODE <- synStore(CODE)
+```
 <!-- # Store covariates -->

### 2020-04-14 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
note random effects

@@
 
-<!-- ### Significant Covariates -->
-<!-- Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates -->
-<!-- ```{r preAdjusted.covariates, cache=TRUE} -->
-<!-- # Find correlation between PC's of gene expression with covariates -->
-<!-- preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E,  -->
-<!--                                                  COVARIATES, -->
-<!--                                                  'NULL design(voom-normalized)',  -->
-<!--                                                  isKeyPlot=TRUE,  -->
-<!--                                                  MIN_PVE_PCT_PC = 1) -->
-<!-- ``` -->
-
-<!-- Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars` -->
-<!-- ```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12} -->
-<!-- preAdjustedSigCovars[["PC_res"]][[2]]$plotData -->
-<!-- ``` -->
-
-### Model Identification
-```{r fxn}
-# Function to fit a mixed effect model and return BIC
-fitMixedEffectModel <- function(varsToFit, random_effect, lcpm, covars, covars_to_fit){
-
-}
-```
-
 #### Phase I (clinical, ancestry and sample specific technical variables)
@@
                          data = COVARIATES,
-                         variables = array(c("Institution", "Reported_Gender", "PMI", 
+                         variables = array(c("(1|Institution)", "Reported_Gender", "PMI", 
                                                              "ageOfDeath", "RIN", "RIN2", "EV.1", 
@@
                          data = COVARIATES,
-                         variables = array(c("LibraryBatch", "FlowcellBatch")))
+                         variables = array(c("(1|LibraryBatch)", "(1|FlowcellBatch)")))
 phase2

### 2020-04-16 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
categorical variables as random effects

@@
                          data = COVARIATES,
-                         variables = array(c("(1|Institution)", "Reported_Gender", "PMI", 
+                         variables = array(c("(1|Institution)", "(1|Reported_Gender)", "PMI", 
                                                              "ageOfDeath", "RIN", "RIN2", "EV.1", 
@@
                          data = COVARIATES,
-                         variables = array(c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected", "rRNARate", "TotalReads")))
+                         variables = array(c("IntragenicRate", "IntronicRate","IntergenicRate","rRNARate", "TotalReads", "GenesDetected", "MappedReads")))
 phase3

### 2020-04-17 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
parallel processing with SnowParam
test Total Reads as random effect

@@
 
-library(doParallel)
+library(BiocParallel)
 library(foreach)
@@
 
-cl = makeCluster((availableCores()-2)/2)
-
-registerDoParallel(cl)
+param = SnowParam((availableCores()-2), "SOCK", progressbar=TRUE)
+register(param)
 
@@
                          data = COVARIATES,
-                         variables = array(c("IntragenicRate", "IntronicRate","IntergenicRate","rRNARate", "TotalReads", "GenesDetected", "MappedReads")))
+                         variables = array(c("IntragenicRate", "IntronicRate","IntergenicRate","rRNARate", "(1|TotalReads)", "GenesDetected", "MappedReads")))
 phase3

### 2020-04-17 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
test phase III

@@
 
-#### Phase I (clinical, ancestry and sample specific technical variables)
-```{r phase1}
-# Fit a base model with the following covariates
-base_covars <- "Dx.Tissue"
-random_effect <- "Individual_ID"
+<!-- #### Phase I (clinical, ancestry and sample specific technical variables) -->
+<!-- ```{r phase1} -->
+<!-- # Fit a base model with the following covariates -->
+<!-- base_covars <- "Dx.Tissue" -->
+<!-- random_effect <- "Individual_ID" -->
 
-# Set base formula
-base_formula <- glue::glue("~ ", glue::glue_collapse(base_covars, sep = " + ")) %>% 
-  glue::glue(" + (1|{random_effect})")
+<!-- # Set base formula -->
+<!-- base_formula <- glue::glue("~ ", glue::glue_collapse(base_covars, sep = " + ")) %>%  -->
+<!--   glue::glue(" + (1|{random_effect})") -->
 
-# Stepwise regression
-phase1 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN,
-                         baseFormula = base_formula, 
-                         data = COVARIATES,
-                         variables = array(c("(1|Institution)", "(1|Reported_Gender)", "PMI", 
-                                                             "ageOfDeath", "RIN", "RIN2", "EV.1", 
-                                                             "EV.2", "EV.3", "EV.4", "EV.5")))
-phase1
-```
+<!-- # Stepwise regression -->
+<!-- phase1 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN, -->
+<!--                          baseFormula = base_formula,  -->
+<!--                          data = COVARIATES, -->
+<!--                          variables = array(c("(1|Institution)", "(1|Reported_Gender)", "PMI",  -->
+<!--                                                              "ageOfDeath", "RIN", "RIN2", "EV.1",  -->
+<!--                                                              "EV.2", "EV.3", "EV.4", "EV.5"))) -->
+<!-- phase1 -->
+<!-- ``` -->
 
-#### Phase II (batch effects)
-```{r phase2}
-phase2 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN,
-                         baseFormula = phase1$formula, 
-                         data = COVARIATES,
-                         variables = array(c("(1|LibraryBatch)", "(1|FlowcellBatch)")))
-phase2
-```
+<!-- #### Phase II (batch effects) -->
+<!-- ```{r phase2} -->
+<!-- phase2 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN, -->
+<!--                          baseFormula = phase1$formula,  -->
+<!--                          data = COVARIATES, -->
+<!--                          variables = array(c("(1|LibraryBatch)", "(1|FlowcellBatch)"))) -->
+<!-- phase2 -->
+<!-- ``` -->
 
@@
 ```{r phase3}
+phase2_formula <- "~Dx.Tissue + (1 | Individual_ID) + RIN2 + (1 | Institution) + ageOfDeath + RIN + PMI + EV.1 + (1 | Reported_Gender) + EV.2 + EV.3 + EV.4"
 phase3 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN,
-                         baseFormula = phase2$formula, 
+                         baseFormula = phase2_formula, 
                          data = COVARIATES,
-                         variables = array(c("IntragenicRate", "IntronicRate","IntergenicRate","rRNARate", "(1|TotalReads)", "GenesDetected", "MappedReads")))
+                         variables = array(c("IntragenicRate", "IntronicRate","IntergenicRate","rRNARate", "TotalReads", "GenesDetected", "MappedReads")))
 phase3

### 2020-04-17 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove commented code

@@
 
-# # Define biomart object
-# mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
-#                 host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)
-#                 dataset = "hsapiens_gene_ensembl")
-# 
-# # Query biomart
-# Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
-#                        filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
-#                        mart = mart)
 Ensemble2HGNC = downloadFile("syn21907981")
@@
 
-<!-- ### Sex Check -->
-<!-- ```{r Reported_Gender_plots} -->
-<!-- counts$gene_id = rownames(counts)  -->
-<!-- REPORTED.GENDER.COUNTS = GENE.GC.CONT %>%  -->
-<!--   left_join(counts) %>% -->
-<!--   dplyr::select(-one_of("percentage_gene_gc_content")) %>% -->
-<!--   filter(chromosome_name == "X" |chromosome_name == "Y") %>%  -->
-<!--   tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>% -->
-<!--   dplyr::mutate(value = log(value)) %>% -->
-<!--   dplyr::rename(`counts(log)`= value) %>%  -->
-<!--   dplyr::rename(SampleID = item) %>% -->
-<!--   left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>%  -->
-<!--   dplyr::rename(`Reported Gender` = Reported_Gender)  -->
-
-<!-- ##XIST and UTY expression  -->
-<!-- #ENSG00000229807.11 and ENSG00000183878.15  -->
-<!-- FILT <- REPORTED.GENDER.COUNTS %>%  -->
-<!--   filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>%  -->
-<!--   dplyr::select(-one_of("chromosome_name")) %>%  -->
-<!--   tidyr::spread(key = gene_id, value = `counts(log)`) %>%  -->
-<!--   mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>%  -->
-<!--   mutate(UTY = as.numeric(`ENSG00000183878.15`)) -->
-
-<!-- p = ggplot(FILT, aes (x= XIST, y = UTY))  -->
-<!-- p = p + geom_point(aes(color=`Reported Gender`))  -->
-<!-- p -->
-<!-- ``` -->
 ### Covariates clustering
@@
 ```
-<!-- Correlation/association between covariates at an FDR <= 0.1 -->
-<!-- ```{r covariates.correlation, fig.width=10, fig.height=8} -->
-<!-- COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05) -->
-<!-- tmp = COVARIATES.CORRELATION$ESTIMATE -->
-<!-- tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0 -->
-<!-- h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate') -->
-<!-- ComplexHeatmap::draw(h, heatmap_legend_side = 'left') -->
-<!-- ``` -->
 
-<!-- ## Explore metatdata -->
-<!-- ```{r data.explore, fig.width = 12, fig.height = 12} -->
-<!-- my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5)) -->
-
-<!-- # RIN -->
-<!-- p = list() -->
-<!-- p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot() -->
-<!-- p[[1]] = p[[1]] + ggtitle('RIN') + my.theme -->
-
-<!-- # Age of Death -->
-<!-- p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = ageOfDeath)) + geom_boxplot() -->
-<!-- p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme -->
-
-<!-- # PMI -->
-<!-- p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot() -->
-<!-- p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme -->
-
-<!-- # Intronic Rate -->
-<!-- p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot() -->
-<!-- p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme -->
-
-<!-- # IntergenicRate -->
-<!-- p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot() -->
-<!-- p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme -->
-
-<!-- # Transcripts Detected -->
-<!-- p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntragenicRate)) + geom_boxplot() -->
-<!-- p[[6]] = p[[6]] + ggtitle('Intragenic Rate') + my.theme -->
-
-<!-- # Mapped Reads -->
-<!-- p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot() -->
-<!-- p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme -->
-
-<!-- # PercentAligned -->
-<!-- p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TotalReads)) + geom_boxplot() -->
-<!-- p[[8]] = p[[8]] + ggtitle('Total Reads') + my.theme -->
-
-<!-- # rRNARate -->
-<!-- p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot() -->
-<!-- p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme -->
-
-<!-- multiplot(plotlist = p, cols = 3) -->
-
-<!-- # Institution -->
-<!-- # vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES) -->
-
-<!-- # Gender -->
-<!-- # vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES) -->
-<!-- ``` -->
 ### Filter genes
@@
 
-<!-- ### Co-expression distribution -->
-<!-- Coexpression of genes  -->
-<!-- ```{r coexp1, cache=FALSE, fig.height=5, fig.width=5} -->
-<!-- cr = cor(t(CQN.GENE_EXPRESSION$E)) -->
-<!-- hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation') -->
-<!-- ``` -->
-
-<!-- ### Sample clustering -->
-<!-- ```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE} -->
-<!-- # Find principal components of expression to plot -->
-<!-- PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T) -->
-
-<!-- # Plot first 2 PCs -->
-<!-- plotdata <- data.frame(SampleID=rownames(PC$rotation),  -->
-<!--                        PC1=PC$rotation[,1],  -->
-<!--                        PC2=PC$rotation[,2]) -->
-
-<!-- plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>% -->
-<!--   tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') -->
-
-<!-- p <- ggplot(plotdata, aes(x=PC1, y=PC2)) -->
-<!-- p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath)) -->
-<!-- p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.) -->
-<!-- # p <- p + geom_text(aes(label= SampleID), size=4, hjust=0) -->
-<!-- p -->
-<!-- ``` -->
-<!-- Tree based clustering of samples -->
-<!-- ```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'} -->
-<!-- # Eucledian tree based analysis -->
-<!-- COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue")]) -->
-<!-- COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0 -->
-
-<!-- tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E))) -->
-<!-- cols = WGCNA::labels2colors(COVARIATES.tmp); -->
-
-<!-- WGCNA::plotDendroAndColors(tree,  -->
-<!--                            colors = cols,  -->
-<!--                            dendroLabels = FALSE,  -->
-<!--                            abHeight = 0.80,  -->
-<!--                            main = "Sample dendrogram", -->
-<!--                            groupLabels = colnames(COVARIATES.tmp)) -->
-<!-- ``` -->
-<!-- ```{r temp, include = F} -->
-<!-- dev.off() -->
-<!-- gc() -->
-<!-- ``` -->
-
-<!-- ### Distribution of samples (log cpm) -->
-<!-- ```{r lcpm.dist, cache=FALSE} -->
-<!-- # Plot abberent distribution of logcpm counts -->
-<!-- tmp1 = (CQN.GENE_EXPRESSION$E) %>% -->
-<!--   rownameToFirstColumn('gene_id') %>% -->
-<!--   tidyr::gather(SampleID, logCPM, -gene_id) %>% -->
-<!--   left_join(COVARIATES %>% -->
-<!--               rownameToFirstColumn('SampleID')) -->
-
-<!-- p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density()  -->
-<!-- p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx.Tissue, scale = 'free') -->
-<!-- p -->
-<!-- ``` -->
-
 ### Subset Genes
@@
 ```
-
-<!-- #### Phase I (clinical, ancestry and sample specific technical variables) -->
-<!-- ```{r phase1} -->
-<!-- # Fit a base model with the following covariates -->
-<!-- base_covars <- "Dx.Tissue" -->
-<!-- random_effect <- "Individual_ID" -->
-
-<!-- # Set base formula -->
-<!-- base_formula <- glue::glue("~ ", glue::glue_collapse(base_covars, sep = " + ")) %>%  -->
-<!--   glue::glue(" + (1|{random_effect})") -->
-
-<!-- # Stepwise regression -->
-<!-- phase1 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN, -->
-<!--                          baseFormula = base_formula,  -->
-<!--                          data = COVARIATES, -->
-<!--                          variables = array(c("(1|Institution)", "(1|Reported_Gender)", "PMI",  -->
-<!--                                                              "ageOfDeath", "RIN", "RIN2", "EV.1",  -->
-<!--                                                              "EV.2", "EV.3", "EV.4", "EV.5"))) -->
-<!-- phase1 -->
-<!-- ``` -->
-
-<!-- #### Phase II (batch effects) -->
-<!-- ```{r phase2} -->
-<!-- phase2 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN, -->
-<!--                          baseFormula = phase1$formula,  -->
-<!--                          data = COVARIATES, -->
-<!--                          variables = array(c("(1|LibraryBatch)", "(1|FlowcellBatch)"))) -->
-<!-- phase2 -->
-<!-- ``` -->
 
@@
 
-<!-- ### Store files in synapse -->
-<!-- ```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE} -->
-<!-- # Set annotations -->
-<!-- all.annotations = list( -->
-<!--   dataType = 'geneExpression', -->
-<!--   dataSubtype = 'processed', -->
-<!--   assay	 = 'rnaSeq', -->
-
-<!--   tissue	= 'dorsolateral prefrontal cortex, anterior cingulate cortex',  -->
-<!--   study = 'CMC',  -->
-
-<!--   species = 'Human', -->
-<!--   consortium	= 'CMC', -->
-
-<!--   analysisType	= "data normalization", -->
-
-<!--   transcriptNormalizationMethod	= 'CQN', -->
-<!--   transcriptQuantificationMethod = 'featureCounts', -->
-<!--   referenceSet = 'GRCh38' -->
-<!-- ) -->
-
 # Code
@@
 ```
-<!-- # Store covariates -->
-<!-- COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID') -->
-<!-- write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F) -->
-<!-- COV_OBJ = synapser::File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id) -->
-<!-- COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName,  -->
-<!--                       executed = thisFile, activityDescription = activityDescription) -->
-<!-- synSetAnnotations(COV_OBJ, annotations = all.annotations) -->
-
-<!-- # Store filtered counts -->
-<!-- PROCESSED_COUNTS$filteredExprMatrix$counts %>% -->
-<!--   rownameToFirstColumn('ensembl_gene_id') %>% -->
-<!--   write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F) -->
-<!-- COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id) -->
-<!-- COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName,  -->
-<!--                       executed = thisFile, activityDescription = activityDescription) -->
-<!-- synSetAnnotations(COUNT_OBJ, annotations = all.annotations) -->
-
-<!-- # Store logCPM -->
-<!-- CQN.GENE_EXPRESSION$E %>% -->
-<!--   rownameToFirstColumn('ensembl_gene_id') %>% -->
-<!--   write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F) -->
-<!-- LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id) -->
-<!-- LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName,  -->
-<!--                       executed = thisFile, activityDescription = activityDescription) -->
-<!-- synSetAnnotations(LCOUNT_OBJ, annotations = all.annotations) -->
-
-<!-- # Store cqn offsets -->
-<!-- CQN.GENE_EXPRESSION$offset %>% -->
-<!--   rownameToFirstColumn('ensembl_gene_id') %>% -->
-<!--   write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F) -->
-<!-- OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id) -->
-<!-- OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName,  -->
-<!--                       executed = thisFile, activityDescription = activityDescription) -->
-<!-- synSetAnnotations(OFFSET_OBJ, annotations = all.annotations) -->
-<!-- ``` -->
-
-<!-- |  *Results*                                  |  *SynapseID*                     | -->
-<!-- |  -----------------------------------------  |   ---------                      | -->
-<!-- |  Covariates                                 |  `r COV_OBJ$properties$id`       | -->
-<!-- |  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     | -->
-<!-- |  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    | -->
-<!-- |  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    | -->
-
-### Source Rmd
-[Source markdown](`r thisFile`)

### 2020-04-17 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
mvBIC to all three phases

@@
                                  parentId ="syn21435176",
-                                 folderName = "test mvBIC - MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN")
+                                 folderName = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN")
 ```
@@
 library(githubr) # get the package from devtools::install_github('brian-bot/githubr')
-
-library(mvBIC) # devtools::install_github("GabrielHoffman/mvBIC")
 
@@
 # Github link
-thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='test-mvBIC')
+thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='add-dream')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))
@@
 ```{r gene.param}
-# Get GC content from biomart
+## Get GC content from biomart
 backgroundGenes = data.frame(gene_id = rownames(counts)) %>%
@@
 
-Ensemble2HGNC = downloadFile("syn21907981")
+# Define biomart object
+mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
+                host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)
+                dataset = "hsapiens_gene_ensembl")
+
+# Query biomart
+Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
+                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
+                       mart = mart)
+
 GENE.GC.CONT = Ensemble2HGNC %>%
@@
 
+### Sex Check
+```{r Reported_Gender_plots}
+counts$gene_id = rownames(counts) 
+REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
+  left_join(counts) %>%
+  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
+  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
+  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
+  dplyr::mutate(value = log(value)) %>%
+  dplyr::rename(`counts(log)`= value) %>% 
+  dplyr::rename(SampleID = item) %>%
+  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
+  dplyr::rename(`Reported Gender` = Reported_Gender)
+
+##XIST and UTY expression 
+#ENSG00000229807.11 and ENSG00000183878.15 
+FILT <- REPORTED.GENDER.COUNTS %>% 
+  filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>% 
+  dplyr::select(-one_of("chromosome_name")) %>% 
+  tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
+  mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>% 
+  mutate(UTY = as.numeric(`ENSG00000183878.15`))
+
+p = ggplot(FILT, aes (x= XIST, y = UTY)) 
+p = p + geom_point(aes(color=`Reported Gender`)) 
+p
+```
 ### Covariates clustering
@@
 ```
+Correlation/association between covariates at an FDR <= 0.1
+```{r covariates.correlation, fig.width=10, fig.height=8}
+COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
+tmp = COVARIATES.CORRELATION$ESTIMATE
+tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
+h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
+ComplexHeatmap::draw(h, heatmap_legend_side = 'left')
+```
 
+## Explore metatdata
+```{r data.explore, fig.width = 12, fig.height = 12}
+my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))
+
+# RIN
+p = list()
+p[[1]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = RIN)) + geom_boxplot()
+p[[1]] = p[[1]] + ggtitle('RIN') + my.theme
+
+# Age of Death
+p[[2]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = ageOfDeath)) + geom_boxplot()
+p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + my.theme
+
+# PMI
+p[[3]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = PMI)) + geom_boxplot()
+p[[3]] = p[[3]] + ggtitle('PMI (in hours)') + my.theme
+
+# Intronic Rate
+p[[4]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntronicRate)) + geom_boxplot()
+p[[4]] = p[[4]] + ggtitle('Intronic Rate') + my.theme
+
+# IntergenicRate
+p[[5]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntergenicRate)) + geom_boxplot()
+p[[5]] = p[[5]] + ggtitle('Intergenic Rate') + my.theme
+
+# Transcripts Detected
+p[[6]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = IntragenicRate)) + geom_boxplot()
+p[[6]] = p[[6]] + ggtitle('Intragenic Rate') + my.theme
+
+# Mapped Reads
+p[[7]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = MappedReads)) + geom_boxplot()
+p[[7]] = p[[7]] + ggtitle('Mapped Reads') + my.theme
+
+# PercentAligned
+p[[8]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = TotalReads)) + geom_boxplot()
+p[[8]] = p[[8]] + ggtitle('Total Reads') + my.theme
+
+# rRNARate
+p[[9]] = ggplot(COVARIATES, aes(x = Dx.Tissue, y = rRNARate)) + geom_boxplot()
+p[[9]] = p[[9]] + ggtitle('rRNARate') + my.theme
+
+multiplot(plotlist = p, cols = 3)
+
+# Institution
+# vcd::mosaic(~ Institution + Dx.Tissue, data = COVARIATES)
+
+# Gender
+# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
+```
 ### Filter genes
@@
 gene_lengths = tibble::column_to_rownames(gene_lengths, var = 'gene_id') %>% as.data.frame()
-CQN.GENE_EXPRESSION = cqn::cqn(NEW.COUNTS, 
+CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                           x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gene_gc_content'],
@@
 
-### Subset Genes
-```{r subset-genes}
-rand_sample <- sample(1:dim(NEW.COUNTS)[1], size = round(dim(NEW.COUNTS)[1]*.6), replace = FALSE)
-subset_NEW.COUNTS<- NEW.COUNTS[rand_sample,]
+### Co-expression distribution
+Coexpression of genes 
+```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
+cr = cor(t(CQN.GENE_EXPRESSION$E))
+hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
+```
 
-subset_CQN <- CQN.GENE_EXPRESSION$E[rand_sample,]
+### Sample clustering
+```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
+# Find principal components of expression to plot
+PC <- prcomp(CQN.GENE_EXPRESSION$E, scale.=T, center = T)
+
+# Plot first 2 PCs
+plotdata <- data.frame(SampleID=rownames(PC$rotation), 
+                       PC1=PC$rotation[,1], 
+                       PC2=PC$rotation[,2])
+
+plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
+  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_')
+
+p <- ggplot(plotdata, aes(x=PC1, y=PC2))
+p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
+p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
+# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
+p
 ```
+Tree based clustering of samples
+```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
+# Eucledian tree based analysis
+COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx.Tissue")])
+COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
+
+tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$E)))
+cols = WGCNA::labels2colors(COVARIATES.tmp);
+
+WGCNA::plotDendroAndColors(tree, 
+                           colors = cols, 
+                           dendroLabels = FALSE, 
+                           abHeight = 0.80, 
+                           main = "Sample dendrogram",
+                           groupLabels = colnames(COVARIATES.tmp))
+```
+
+```{r temp, include = F}
+dev.off()
+gc()
+```
+
+### Distribution of samples (log cpm)
+```{r lcpm.dist, cache=FALSE}
+# Plot abberent distribution of logcpm counts
+tmp1 = (CQN.GENE_EXPRESSION$E) %>%
+  rownameToFirstColumn('gene_id') %>%
+  tidyr::gather(SampleID, logCPM, -gene_id) %>%
+  left_join(COVARIATES %>%
+              rownameToFirstColumn('SampleID'))
+
+p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
+p = p + theme_bw() %+replace% theme(legend.position = 'NONE') + facet_grid(.~Dx.Tissue, scale = 'free')
+p
+```
+
+### Significant Covariates
+Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
+```{r preAdjusted.covariates, cache=TRUE}
+# Find correlation between PC's of gene expression with covariates
+preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$E, 
+                                                 COVARIATES,
+                                                 'NULL design(voom-normalized)', 
+                                                 isKeyPlot=TRUE, 
+                                                 MIN_PVE_PCT_PC = 1)
+```
+
+Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
+```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
+preAdjustedSigCovars[["PC_res"]][[2]]$plotData
+```
+
+### Model Identification
+
+#### Phase I (clinical, ancestry and sample specific technical variables)
+```{r phase1}
+# Fit a base model with the following covariates
+base_covars <- "Dx.Tissue"
+random_effect <- "Individual_ID"
+
+# Set base formula
+base_formula <- glue::glue("~ ", glue::glue_collapse(base_covars, sep = " + ")) %>% 
+  glue::glue(" + (1|{random_effect})")
+
+# Stepwise regression
+phase1 <- mvBIC::mvForwardStepwise(exprObj = CQN.GENE_EXPRESSION$E,
+                         baseFormula = base_formula, 
+                         data = COVARIATES,
+                         variables = array(c("(1|Institution)", "(1|Reported_Gender)", "PMI", 
+                                                             "ageOfDeath", "RIN", "RIN2", "EV.1", 
+                                                             "EV.2", "EV.3", "EV.4", "EV.5")))
+phase1
+```
+Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`
+
+#### Phase II (batch effects)
+```{r phase2}
+# Fit a base model with the following covariates
+phase2 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN,
+                         baseFormula = phase1$formula, 
+                         data = COVARIATES,
+                         variables = array(c("(1|LibraryBatch)", "(1|FlowcellBatch)")))
+phase2
+```
+Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`
 
@@
 ```{r phase3}
-phase2_formula <- "~Dx.Tissue + (1 | Individual_ID) + RIN2 + (1 | Institution) + ageOfDeath + RIN + PMI + EV.1 + (1 | Reported_Gender) + EV.2 + EV.3 + EV.4"
+# Fit a base model with the following covariates
 phase3 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN,
-                         baseFormula = phase2_formula, 
+                         baseFormula = phase2$formula, 
                          data = COVARIATES,
-                         variables = array(c("IntragenicRate", "IntronicRate","IntergenicRate","rRNARate", "TotalReads", "GenesDetected", "MappedReads")))
+                         variables = array(c("scale(IntragenicRate)", "scale(IntronicRate)","scale(IntergenicRate)","scale(rRNARate)", 
+                                             "scale(TotalReads)", "scale(GenesDetected)", "scale(MappedReads)")))
 phase3
@@
 #### Variables included in the model 
-`r phase3$formula`
+Covariates were added as fixed and random effects iteratively in three phases if model improvement by BIC criteria was observed.
+Clinical, ancestry and sample-specific technical variables were tested for model improvement first.
+Covariates related to batch effects were next tested for model improvement and finally, RNASeq alignment-specific covariates.
+
+The model for clinical, technical and batch variables selected after phase I, II and III is `r paste(phase3$formula)`
+
+### Store files in synapse
+```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
+# Set annotations
+all.annotations = list(
+  dataType = 'geneExpression',
+  dataSubtype = 'processed',
+  assay	 = 'rnaSeq',
+
+  tissue	= 'dorsolateral prefrontal cortex, anterior cingulate cortex', 
+  study = 'CMC',
+
+  species = 'Human',
+  consortium	= 'CMC',
+
+  analysisType	= "data normalization",
+
+  transcriptNormalizationMethod	= 'CQN',
+  transcriptQuantificationMethod = 'featureCounts',
+  referenceSet = 'GRCh38'
+)
 
 # Code
-``` {r save_code}
-CODE <- Folder(name = "test mvBIC - MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
+CODE <- Folder(name = "MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN", parentId = parentId)
 CODE <- synStore(CODE)
+
+# Store covariates
+COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
+write.table(COVARIATES, file = 'CMC_ACC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
+COV_OBJ = synapser::File('CMC_ACC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
+COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                      executed = thisFile, activityDescription = activityDescription)
+synSetAnnotations(COV_OBJ, annotations = all.annotations)
+
+# Store filtered counts
+PROCESSED_COUNTS$filteredExprMatrix$counts %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
+COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
+COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                      executed = thisFile, activityDescription = activityDescription)
+synSetAnnotations(COUNT_OBJ, annotations = all.annotations)
+
+# Store logCPM
+CQN.GENE_EXPRESSION$E %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'CMC_ACC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
+LCOUNT_OBJ = File('CMC_ACC_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
+LCOUNT_OBJ = synStore(LCOUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                      executed = thisFile, activityDescription = activityDescription)
+synSetAnnotations(LCOUNT_OBJ, annotations = all.annotations)
+
+# Store cqn offsets
+CQN.GENE_EXPRESSION$offset %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
+  write.table(file = 'CMC_ACC_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
+OFFSET_OBJ = File('CMC_ACC_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
+OFFSET_OBJ = synStore(OFFSET_OBJ, used = ALL_USED_IDs, activityName = activityName, 
+                      executed = thisFile, activityDescription = activityDescription)
+synSetAnnotations(OFFSET_OBJ, annotations = all.annotations)
 ```
+
+|  *Results*                                  |  *SynapseID*                     |
+|  -----------------------------------------  |   ---------                      |
+|  Covariates                                 |  `r COV_OBJ$properties$id`       |
+|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
+|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
+|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
+
+### Source Rmd
+[Source markdown](`r thisFile`)

### 2020-04-17 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
biomart obj on Synapse while site is down

@@
 
+# Biomart down until April 20th maintenance
 # Define biomart object
-mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
-                host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)
-                dataset = "hsapiens_gene_ensembl")
-
-# Query biomart
-Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
-                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
-                       mart = mart)
+# mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
+#                 host = "uswest.ensembl.org", # Ensembl Release 99 (January 2020)
+#                 dataset = "hsapiens_gene_ensembl")
+# 
+# # Query biomart
+# Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
+#                        filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
+#                        mart = mart)
+Ensemble2HGNC = downloadFile("syn21907981")
 

### 2020-04-17 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
note branch

@@
 # Github link
-thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='add-dream')
+thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='mvBIC-all-cohorts')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))

### 2020-04-20 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
scale continuous vars

@@
                          data = COVARIATES,
-                         variables = array(c("(1|Institution)", "(1|Reported_Gender)", "PMI", 
-                                                             "ageOfDeath", "RIN", "RIN2", "EV.1", 
-                                                             "EV.2", "EV.3", "EV.4", "EV.5")))
+                         variables = array(c("(1|Institution)", "(1|Reported_Gender)", "scale(PMI)", 
+                                                             "scale(ageOfDeath)", "scale(RIN)", "scale(RIN2)", "scale(EV.1)", 
+                                                             "scale(EV.2)", "scale(EV.3)", "scale(EV.4)", "scale(EV.5)")))
 phase1

### 2020-04-21 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
fix markdown output

--- 2020-04-17 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
@@
                          data = COVARIATES,
-                         variables = array(c("(1|Institution)", "(1|Reported_Gender)", "PMI", 
-                                                             "ageOfDeath", "RIN", "RIN2", "EV.1", 
-                                                             "EV.2", "EV.3", "EV.4", "EV.5")))
+                         variables = array(c("(1|Institution)", "(1|Reported_Gender)", "scale(PMI)", 
+                                                             "scale(ageOfDeath)", "scale(RIN)", "scale(RIN2)", "scale(EV.1)", 
+                                                             "scale(EV.2)", "scale(EV.3)", "scale(EV.4)", "scale(EV.5)")))
 phase1
 ```
-Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`
 
@@
 ```
-Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`
 

### 2020-04-21 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
model selection on full gene set

@@
 # Fit a base model with the following covariates
-phase2 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN,
+phase2 <- mvBIC::mvForwardStepwise(exprObj = CQN.GENE_EXPRESSION$E,
                          baseFormula = phase1$formula, 
@@
 # Fit a base model with the following covariates
-phase3 <- mvBIC::mvForwardStepwise(exprObj = subset_CQN,
+phase3 <- mvBIC::mvForwardStepwise(exprObj = CQN.GENE_EXPRESSION$E,
                          baseFormula = phase2$formula, 

### 2020-04-22 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd

@@
   dplyr::select(Individual_ID, `SNP_report:Genotyping_Sample_ID`, `SNP_report:Exclude?`) %>% 
-  dplyr::inner_join(ANCESTRY)
+  dplyr::inner_join(ANCESTRY) %>% 
+  dplyr::filter(is.na(`SNP_report:Exclude?`))
 ```

### 2020-04-22 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
mvBIC

@@
 ### Model Identification
-```{r fxn}
-# Function to fit a mixed effect model and return BIC
-fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){
 
-  # Get design matrix
-  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
-  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]
-
-  # Estimate voom weights
-  cnt[is.na(cnt)] = 0
-  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights
-
-  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
-                              .combine = rbind,
-                              .packages = c('lme4', 'dplyr'),
-                              .export = c('voomWeights')) %dopar% {
-
-                                # Perform weighted mixed linear models for every gene
-                                expr = data.frame(Gx = lcpm[i,])
-                                data = cbind(expr,
-                                             designMat[rownames(expr),],
-                                             Individual_ID = groupingVar)
-                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
-                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')
-
-                                weights = voomWeights[i,]
-                                BIC = NA
-                                tryCatch({
-                                  mdl = lmer(formula = formula, 
-                                             data = data, 
-                                             weights = weights)
-                                  BIC = BIC(mdl)
-                                }, error = function(e){
-                                  BIC = NA
-                                })
-                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
-                                                BIC = BIC))
-                            }
-  return(modelBIC)
-  }
-
-# Function to select covariates in a step-wise regression fashion
-selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){
-
-  varsSelected = c()
-  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
-           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
-  while(length(covarsToSelect) != 0){
-    allModelBIC = foreach(i = 1:length(covarsToSelect),
-                          .combine = c,
-                          .packages = pkgs,
-                          .export = c('fitMixedEffectModel')) %dopar% {
-
-                            # Fit a mixed effect linear model and get all BIC
-                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
-                                                           covars[colnames(cnt),'Individual_ID'], 
-                                                           lcpm, 
-                                                           cnt, 
-                                                           covars) %>%
-                            plyr::rename(c('BIC' = covarsToSelect[i]))
-
-                            return(list(modelBIC))
-                          }
-    allModelBIC = plyr::join_all(allModelBIC)
-
-    # Find number of genes that are changed based on BIC
-    nChange = allModelBIC %>%
-      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
-      dplyr::left_join(baseModelBIC) %>%
-      dplyr::mutate(diff = baseModel - Value) %>%
-      dplyr::group_by(ModelName) %>%
-      dplyr::summarise(better = sum(diff >= 5), 
-                       worse = sum(diff <= -5)) %>%
-      dplyr::mutate(nchange = better - worse)
-
-    print(nChange)
-
-    # Select variable
-    tmp = nChange %>%
-      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
-      dplyr::top_n(1, nchange)
-
-    if (dim(tmp)[1] == 0){
-      covarsToSelect = c()
-    } else {
-      baseModelBIC = allModelBIC %>%
-        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
-      colnames(baseModelBIC)[2] = 'baseModel'
-      varsSelected = c(varsSelected, tmp$ModelName)
-      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
-      print(paste('Variable chosen to be added:',tmp$ModelName))
-    }
-  }
-
-  return(varsSelected)
-}
-```
 #### Phase I (clinical, ancestry and sample specific technical variables)
 ```{r phase1}
-# Fit a base model with the following covariates
-adjust.covars = c('Dx')
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-# Iteratively identify covariates in phase I
-phase1.covars = c('Reported_Gender', 'PMI', 'ageOfDeath', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
-vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
+#Fit a base model with the following covariates
+base_covars <- "Dx"
+random_effect <- "Individual_ID"
 
-adjust.covars = c(adjust.covars, vars.selected)
+# Set base formula
+base_formula <- glue::glue("~ ", glue::glue_collapse(base_covars, sep = " + ")) %>%
+  glue::glue(" + (1|{random_effect})")
+
+# Stepwise regression
+phase1 <- mvBIC::mvForwardStepwise(exprObj = CQN.GENE_EXPRESSION$E,
+                         baseFormula = base_formula,
+                         data = COVARIATES,
+                         variables = array(c("(1|Institution)", "(1|Reported_Gender)", "scale(PMI)",
+                                                             "scale(ageOfDeath)", "scale(RIN)", "scale(RIN2)", "scale(EV.1)",
+                                                             "scale(EV.2)", "scale(EV.3)", "scale(EV.4)", "scale(EV.5)")))
+phase1
 ```
-Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`
 
@@
 # Fit a base model with the following covariates
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-# Iteratively identify covariates in phase I
-phase2.covars = c('LibraryBatch', 'FlowcellBatch')
-vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-
-adjust.covars = c(adjust.covars, vars.selected)
+phase2 <- mvBIC::mvForwardStepwise(exprObj = CQN.GENE_EXPRESSION$E,
+                         baseFormula = phase1$formula, 
+                         data = COVARIATES,
+                         variables = array(c("(1|LibraryBatch)", "(1|FlowcellBatch)")))
+phase2
 ```
-Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`
 
@@
 # Fit a base model with the following covariates
-baseModelBIC = fitMixedEffectModel(adjust.covars, 
-                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
-                                   CQN.GENE_EXPRESSION$E, 
-                                   NEW.COUNTS, 
-                                   COVARIATES) %>%
-  dplyr::rename(baseModel = BIC)
-# Iteratively identify covariates in phase I
-phase3.covars = c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected",
-                  "ExpProfEfficiency", "rRNARate", "TotalReads")
-vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
-                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
-                                     NEW.COUNTS, baseModelBIC)
-
-adjust.covars = c(adjust.covars, vars.selected)
+phase3 <- mvBIC::mvForwardStepwise(exprObj = CQN.GENE_EXPRESSION$E,
+                         baseFormula = phase2$formula, 
+                         data = COVARIATES,
+                         variables = array(c("scale(IntragenicRate)", "scale(IntronicRate)","scale(IntergenicRate)","scale(rRNARate)", 
+                                             "scale(TotalReads)", "scale(GenesDetected)", "scale(MappedReads)")))
+phase3
 ```
@@
 #### Variables included in the model 
-Covariates were added as fixed effects iteratively in three phases if model improvement by BIC criteria was observed in the majority of genes.
+Covariates were added as fixed and random effects iteratively in three phases if model improvement by BIC criteria was observed.
 Clinical, ancestry and sample-specific technical variables were tested for model improvement first.
@@
 
-Clinical, technical and batch variables selected after phase I, II and III of model identification are `r paste(adjust.covars, collapse = ', ')`
-
+The model for clinical, technical and batch variables selected after phase I, II and III is `r phase3$formula`
 ### Store files in synapse

### 2020-04-23 cmc-rnaseq-de/HBCC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
not testing institution for model sel

@@
 library(stringr)
+library(mvBIC)
 
@@
 # Github link
-thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='master')
+thisRepo <- getRepo(repository = "kelshmo/analysis", ref="branch", refName='mvBIC-all-cohorts')
 thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('cmc-rnaseq-de/',thisFileName))
@@
                          data = COVARIATES,
-                         variables = array(c("(1|Institution)", "(1|Reported_Gender)", "scale(PMI)",
+                         variables = array(c("(1|Reported_Gender)", "scale(PMI)",
                                              "scale(ageOfDeath)", "scale(RIN)", "scale(RIN2)", "scale(EV.1)",

### 2021-02-20 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd

@@
                                 })
-                              return(data.frame(ensembl_transcript_id = rownames(cnt)[i],
+                              return(data.frame(ensembl_gene_id = rownames(cnt)[i],
                                                 BIC = BIC))
@@
     nChange = allModelBIC %>%
-      tidyr::gather(ModelName, Value, -ensembl_transcript_id) %>%
+      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
       dplyr::left_join(baseModelBIC) %>%
@@
       baseModelBIC = allModelBIC %>%
-        dplyr::select(ensembl_transcript_id, one_of(tmp$ModelName))
+        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
       colnames(baseModelBIC)[2] = 'baseModel'
@@
 PROCESSED_COUNTS$filteredExprMatrix$counts %>%
-  rownameToFirstColumn('ensembl_transcript_id') %>%
+  rownameToFirstColumn('ensembl_gene_id') %>%
   write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
